---
title: "Clean-Room TMLE Workflow"
author: "cleanTMLE"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clean-Room TMLE Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = TRUE
)
```

## Overview

This vignette demonstrates the full **clean-room staged TMLE workflow**
using the `run_clean_tmle()` wrapper.
The staged workflow enforces outcome blinding through structured gates:

| Stage | Purpose | Output |
|-------|---------|--------|
| **1** | Lock analytic specification | `cleanroom_lock` |
| **2a** | Propensity-score diagnostics & gating | PS diagnostics, ESS, love plot |
| **2b** | Negative controls / plasmode simulation | Bias, RMSE, coverage |
| **3** | Primary TMLE estimation | ATE, IC, weight diagnostics |

## Simulated Data

We simulate a dataset with a binary treatment, baseline covariates, and
a binary 24-month event outcome.

```{r simulate-data}
library(cleanTMLE)
set.seed(42)

# Simulate via built-in DGP (or manually if sim_func1 not available)
n <- 800
age       <- rnorm(n, 55, 10)
sex       <- rbinom(n, 1, 0.45)
biomarker <- rnorm(n, 0, 1)
comorbid  <- rbinom(n, 1, 0.30)

# Treatment mechanism (confounded)
lp_trt <- -1.0 + 0.02 * age + 0.3 * sex + 0.4 * biomarker + 0.5 * comorbid
ps_true <- plogis(lp_trt)
treatment <- rbinom(n, 1, ps_true)

# Outcome mechanism (binary at 24 months)
lp_out <- -2.0 + 0.01 * age + 0.2 * sex + 0.3 * biomarker +
          0.4 * comorbid - 0.5 * treatment
event_24 <- rbinom(n, 1, plogis(lp_out))

# Negative-control outcome (unrelated to treatment)
nc_outcome <- rbinom(n, 1, plogis(-1.5 + 0.01 * age))

dat <- data.frame(
  age, sex, biomarker, comorbid, treatment, event_24, nc_outcome
)

head(dat)
```

## Stage 1: Analysis Lock

The wrapper validates that a pre-specified SuperLearner library and
truncation rules are provided, then creates a locked specification.

```{r stage1-lock}
# These would normally be pre-registered
sl_lib <- c("SL.glm", "SL.mean")
trunc  <- c(0.01, 0.99)
```

## Full Pipeline via `run_clean_tmle()`

```{r run-pipeline}
result <- run_clean_tmle(
  data                 = dat,
  Avar                 = "treatment",
  Yvar                 = "event_24",
  covariates           = c("age", "sex", "biomarker", "comorbid"),
  learner_lib          = sl_lib,
  truncation           = trunc,
  neg_control_outcomes = "nc_outcome",
  simulation_spec      = list(reps = 20, effect_sizes = c(0.05)),
  stage2a_thresholds   = list(ESS = 0.20, max_smd = 0.25),
  seed                 = 42,
  verbose              = TRUE
)

print(result)
```

## Stage 2a Diagnostics

### Weight Distribution

```{r weight-summary}
result$weight_summary
```

### Extreme Weights

```{r extreme-wts}
result$extreme_weights
```

### PS Overlap Histogram

```{r ps-histogram, fig.cap = "Propensity score overlap by treatment group."}
result$ps_diagnostics$overlap_plot
```

### Love Plot (Covariate Balance)

```{r love-plot, fig.cap = "Standardised mean differences before and after weighting."}
result$love_plot
```

### Effective Sample Size

```{r ess}
result$ps_diagnostics$ess
```

## Stage 2b: Negative Controls

If negative-control outcomes were specified, the pipeline estimates the
treatment effect on each and tests whether it differs from zero.

```{r nc-results}
if (!is.null(result$nc_results)) {
  for (nc in result$nc_results) {
    print(nc)
    cat("\n")
  }
}
```

## Stage 2b: Plasmode Simulation

If a `simulation_spec` was provided, the pipeline runs plasmode
replicates to assess bias, RMSE, and coverage.

```{r plasmode}
if (!is.null(result$plasmode_results)) {
  print(result$plasmode_results$metrics)
}
```

## Stage 3: Primary TMLE Estimation

```{r stage3-estimate}
if (!is.null(result$tmle_result)) {
  print(result$tmle_result)
}
```

### Influence Curve Histogram

```{r ic-hist, fig.cap = "Distribution of the efficient influence curve."}
if (!is.null(result$tmle_result)) {
  ic_histogram(result$tmle_result)
}
```

### IC Summary

```{r ic-summary}
result$ic_summary
```

## Cross-Fitted TMLE (Comparison)

We can also run a 2-fold cross-fitted TMLE for comparison.

```{r cross-fitted}
tmle_cf <- estimate_tmle_risk_point(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbid"),
  sl_library = sl_lib,
  truncate   = 0.01,
  n_folds    = 2L
)

cat("Cross-fitted ATE:\n")
cat(sprintf("  Estimate: %.5f (95%% CI: %.5f, %.5f)\n",
            tmle_cf$estimates$ATE$estimate,
            tmle_cf$estimates$ATE$ci_lower,
            tmle_cf$estimates$ATE$ci_upper))
```

## Comparison with IPW and Matching

To illustrate improvements, we compare the TMLE estimate against
crude, IPTW, and matching approaches using the cleanTMLE infrastructure.

```{r comparison}
# Create lock for comparison workflows
lock <- create_analysis_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbid"),
  seed       = 42
)

ps_fit  <- fit_ps_glm(lock)
iptw    <- run_iptw_workflow(lock, ps_fit, override_clean_room = TRUE)
matched <- run_match_workflow(lock, ps_fit, override_clean_room = TRUE)

# Crude
Y <- dat$event_24; A <- dat$treatment
crude_rd <- mean(Y[A == 1]) - mean(Y[A == 0])

comparison <- data.frame(
  Method   = c("Crude", "IPTW", "PS Matching", "TMLE"),
  Estimate = round(c(crude_rd, iptw$estimate, matched$estimate,
                     result$tmle_result$estimates$ATE$estimate), 5),
  SE       = round(c(NA, iptw$se, matched$se,
                     result$tmle_result$estimates$ATE$se), 5),
  stringsAsFactors = FALSE
)
print(comparison)
```

## Integration Utilities

### External PS and Parallel Support

```{r external-ps}
# Wrap externally computed PS scores
my_ps <- predict(glm(treatment ~ age + sex + biomarker + comorbid,
                      data = dat, family = binomial()), type = "response")
ext_fit <- wrap_ps_fit(lock, ps_scores = my_ps, method = "custom_glm")
print(ext_fit)
```

### Covariate Sanitisation

```{r sanitize}
clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "biomarker", "comorbid"),
  verbose = TRUE)
```

### TMLE-Based Negative Controls

```{r nc-tmle}
nc_tmle <- run_negative_control_tmle(lock, "nc_outcome", ps_fit)
cat(sprintf("TMLE NC: est=%.5f, p=%.4f (method: %s)\n",
            nc_tmle$estimate, nc_tmle$p_value, nc_tmle$method))
```

### Matched-Cohort TMLE

```{r matched-tmle}
matched_idx <- c(which(dat$treatment == 1)[1:50],
                 which(dat$treatment == 0)[1:50])
matched_tmle <- run_matched_tmle(lock, ps_fit, subset_idx = matched_idx,
                                  override_clean_room = TRUE)
cat(sprintf("Matched TMLE ATE: %.5f (n=%d)\n",
            matched_tmle$estimates$ATE$estimate, matched_tmle$n_matched))
```

### Clever Covariate Plot

```{r clever-plot, fig.cap = "Clever covariate H(A,W) by treatment group."}
clever_covariate_plot(ps_fit = ps_fit, lock = lock)
```

## Decision Log

The full decision log records every stage, metric, and gating decision
with timestamps.

```{r decision-log}
knitr::kable(result$decision_log, caption = "Decision Log")
```

## Session Info

```{r session-info}
sessionInfo()
```
