---
title: "cleanTMLE Function Reference"
subtitle: "API Patterns and Quick Examples"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{cleanTMLE Function Reference}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

This article is a compact index of every exported function in
**cleanTMLE**, organised by clean-room stage.
For a narrative worked example, see the companion vignette
*A Staged Clean-Room TMLE Analysis*.

```{r load}
library(cleanTMLE)
dat <- sim_func1(n = 400, seed = 42)
```

---

## Function Index

The table below lists every exported function, its clean-room stage,
and whether it accesses the real outcome.

| Stage | Function | Outcome? | Purpose |
|-------|----------|:--------:|---------|
| Util | `expit()`, `logit()` | -- | Inverse-logit / logit transforms |
| Data | `sim_func1()` | -- | Simulate causal-inference data |
| 1a | `create_analysis_lock()` | No | Lock analytic specification |
| 1a | `validate_analysis_lock()` | No | Verify lock integrity |
| 1a | `attach_estimand()` | No | Attach estimand metadata |
| 1a | `declare_sensitivity_plan()` | No | Declare a sensitivity plan |
| 1a | `define_negative_control()` | No | Register a negative control variable |
| 1b | `checkpoint_cohort_adequacy()` | Yes* | Check Point 1: sample size / events |
| 2 | `fit_ps_glm()` | No | Logistic regression PS |
| 2 | `fit_ps_superlearner()` | No | SuperLearner PS |
| 2 | `compute_ps_diagnostics()` | No | Overlap, ESS, SMDs |
| 2 | `checkpoint_balance()` | No | Check Point 2: balance decision |
| 2b | `tmle_candidate()` | No | Define a TMLE candidate specification |
| 2b | `expand_tmle_candidate_grid()` | No | Generate candidate grid |
| 2b | `validate_tmle_candidates()` | No | Validate candidate list |
| 2b | `run_plasmode_feasibility()` | No† | Plasmode simulation evaluation |
| 2b | `summarize_plasmode_results()` | No | Print plasmode metrics |
| 2b | `select_tmle_candidate()` | No | Select best TMLE specification |
| 2b | `lock_primary_tmle_spec()` | No | Lock selected spec in the lock |
| 2b | `get_primary_tmle_spec()` | No | Retrieve locked spec |
| 2b | `gate_check()` | No | GO / FLAG / STOP for plasmode |
| 3 | `run_negative_control()` | Yes‡ | Negative control analysis |
| 3 | `checkpoint_residual_bias()` | No | Check Point 3: residual bias |
| 4 | `run_crude_workflow()` | **Yes** | Unadjusted risk difference |
| 4 | `run_match_workflow()` | **Yes** | PS matching |
| 4 | `run_iptw_workflow()` | **Yes** | Stabilised IPTW |
| 4 | `fit_tmle_treatment_mechanism()` | No | TMLE g-step (re-uses PS) |
| 4 | `fit_tmle_outcome_mechanism()` | **Yes** | TMLE Q-step |
| 4 | `run_tmle_targeting_step()` | **Yes** | TMLE targeting / fluctuation |
| 4 | `extract_tmle_estimate()` | **Yes** | Final TMLE estimate |
| 4 | `fit_tmle_candidate_set()` | **Yes** | Fit multiple TMLE specs |
| 4 | `fit_final_workflows()` | **Yes** | Run all Stage 4 workflows |
| 4 | `sensitivity_truncation()` | **Yes** | Truncation sensitivity |
| -- | `compute_evalue()` | No | E-value for unmeasured confounding |
| -- | `summarize_cleanroom_results()` | No | Cross-workflow comparison table |
| -- | `create_audit_log()` | No | Initialise audit trail |
| -- | `record_stage()` | No | Append stage entry |
| -- | `record_checkpoint()` | No | Append checkpoint entry |
| -- | `export_audit_trail()` | No | Export trail as data frame |
| -- | `build_stage_manifest()` | No | Print stage path summary |

\* Accesses event counts only.
† Generates simulated outcomes; does not use the real outcome association.
‡ Accesses the negative control outcome, not the primary outcome.

---

## Utilities

```{r utils}
expit(0)            # 0.5
logit(0.5)          # 0
expit(logit(0.73))  # 0.73
```

---

## Stage 1a: Lock, Estimand, Plans

```{r lock}
lock <- create_analysis_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  sl_library = c("SL.glm", "SL.mean"),
  plasmode_reps = 40L,
  seed       = 42L
)

lock <- attach_estimand(lock,
  description          = "Treatment effect on 24-month event risk",
  contrast             = "risk_difference",
  statistical_estimand = "E_W[E(Y|A=1,W) - E(Y|A=0,W)]"
)

lock <- declare_sensitivity_plan(lock, "truncation",
  description = "Vary PS truncation",
  settings    = list(thresholds = c(0.01, 0.05, 0.10)))

lock <- define_negative_control(lock, "nc_outcome")

validate_analysis_lock(lock)
print(lock)
```

---

## Stage 1b: Cohort Adequacy (Check Point 1)

```{r cp1}
cp1 <- checkpoint_cohort_adequacy(lock)
print(cp1)
```

---

## Stage 2: Propensity Score and Balance (Check Point 2)

```{r ps}
ps_fit <- fit_ps_glm(lock, truncate = 0.01)
diag   <- compute_ps_diagnostics(ps_fit)
print(diag)
```

```{r overlap, fig.cap = "PS overlap."}
plot(diag)
```

```{r cp2}
cp2 <- checkpoint_balance(diag, lock_hash = lock$lock_hash)
print(cp2)
```

---

## Stage 2b: TMLE Candidate Selection via Plasmode

```{r candidates}
# Define TMLE candidate specifications
candidates <- list(
  tmle_candidate("glm_t01", "GLM trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05)
)
# Or use the default grid:
# candidates <- expand_tmle_candidate_grid()
```

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(lock, tmle_candidates = candidates,
                                 effect_sizes = c(0.05, 0.10),
                                 reps = 40L)
print(plas)
```

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- lock_primary_tmle_spec(lock, selected)
print(selected)
```

---

## Stage 3: Negative Control (Check Point 3)

```{r nc}
nc  <- run_negative_control(lock, "nc_outcome", ps_fit)
cp3 <- checkpoint_residual_bias(nc, lock_hash = lock$lock_hash)
print(nc)
print(cp3)
```

---

## Stage 4: Estimation

```{r crude}
run_crude_workflow(lock)
```

```{r matching}
match_fit <- run_match_workflow(lock, ps_fit)
print(match_fit)
```

```{r iptw}
iptw_fit <- run_iptw_workflow(lock, ps_fit)
print(iptw_fit)
```

```{r tmle}
g_fit    <- fit_tmle_treatment_mechanism(lock, ps_fit)
Q_fit    <- fit_tmle_outcome_mechanism(lock, g_fit)
tmle_upd <- run_tmle_targeting_step(g_fit, Q_fit)
tmle_fit <- extract_tmle_estimate(tmle_upd)
print(tmle_fit)
```

```{r compare}
summarize_cleanroom_results(
  list(Matching = match_fit, IPTW = iptw_fit, TMLE = tmle_fit)
)
```

---

## Sensitivity and Diagnostics

```{r trunc-sens}
sensitivity_truncation(lock, thresholds = c(0.01, 0.05, 0.10))
```

```{r evalue}
compute_evalue(1.3, ci_bound = 1.05)
```

---

## Model Specification DSL

The `specify_models()` + `identify_*()` API provides a pipe-friendly
interface for time-to-event estimation functions.

```{r spec}
spec <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph") |>
  identify_subject(id)
print(spec)
```

---

## Time-to-Event Estimation

```{r ipwrisk, cache = TRUE}
fit_ipw <- estimate_ipwrisk(spec, risk_time = c(12, 24),
                             weight_type = "iptw", nboot = 20, seed = 42)
print(fit_ipw)
```

```{r ipwrisk-plot, fig.cap = "IPTW risk curves."}
plot(fit_ipw)
```

```{r gcomp}
spec_gc <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment)

fit_gc <- estimate_gcomprisk(spec_gc, risk_time = c(12, 24))
print(fit_gc)
```

```{r aipw}
spec_aipw <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph")

fit_aipw <- estimate_aipwrisk(spec_aipw, risk_time = c(12, 24))
print(fit_aipw)
```

```{r ipwhr}
hr_fit <- estimate_ipwhr(spec)
print(hr_fit)
```

```{r compare-tte}
compare_fits(IPW = fit_ipw, GComp = fit_gc, AIPW = fit_aipw, risk_time = 24)
```

---

## Tables and Weight Diagnostics

```{r table1}
make_table1(dat, vars = c("age", "sex", "biomarker", "comorbidity"),
            by = "treatment", smd = TRUE)
```

```{r table2}
make_table2(fit_ipw, risk_time = 24)
```

```{r wt-summary}
make_wt_summary_table(fit_ipw)
```

```{r extreme}
extreme_weights(fit_ipw, k = 5)
```

---

## Audit Trail

```{r audit}
audit <- create_audit_log(lock)
audit <- record_stage(audit, "Stage 1a", "Lock created")
audit <- record_checkpoint(audit, cp1)
audit <- record_checkpoint(audit, cp2)
audit <- record_checkpoint(audit, cp3)
audit <- record_stage(audit, "Stage 4", "Final estimation complete")
export_audit_trail(audit)
```

```{r manifest}
build_stage_manifest(audit)
```

```{r session-info}
sessionInfo()
```
