---
title: "cleanTMLE Function Reference"
subtitle: "API Patterns and Quick Examples"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{cleanTMLE Function Reference}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

This article is a compact index of every exported function in
**cleanTMLE**, organised by outcome-blind stage.
For a narrative worked example, see the companion vignette
*A Staged Outcome-Blind TMLE Workflow with cleanTMLE*.

```{r load}
library(cleanTMLE)
dat <- sim_func1(n = 400, seed = 42)
```

---

## Outcome-access taxonomy

The function index below labels each exported function by what it
uses internally from the outcome side of the data:

- **No outcome access** --- uses W, A, metadata, lock objects,
  synthetic outcomes, or diagnostics not involving Y.
- **Marginal Y only** --- uses event counts or marginal outcome
  support (no estimation of the A--Y association).
- **Synthetic Y only** --- uses plasmode-generated outcomes; no
  observed Y.
- **Negative-control Y only** --- uses prespecified NC outcomes,
  not the primary outcome.
- **Primary Y** --- estimates the observed
  treatment-primary-outcome association; requires authorization.

These labels describe what exported package functions use
internally, not what a user with raw data access could inspect
outside the package.

---

## Function Index

The table below lists every exported function, its outcome-blind
stage, and what it uses from the outcome side of the data.

| Stage | Function | Outcome? | Purpose |
|-------|----------|:--------:|---------|
| Util | `expit()`, `logit()` | n/a | Inverse-logit / logit transforms |
| Data | `sim_func1()` | n/a | Simulate causal-inference data |
| 1a | `create_analysis_lock()` | No | Lock analytic specification |
| 1a | `validate_analysis_lock()` | No | Verify lock integrity |
| 1a | `attach_estimand()` | No | Attach estimand metadata |
| 1a | `declare_sensitivity_plan()` | No | Declare a sensitivity plan |
| 1a | `define_negative_control()` | No | Register a negative control variable |
| 1b | `checkpoint_cohort_adequacy()` | Marginal Y only | Check Point 1: sample size / events |
| 1b | `estimate_design_precision()` | Marginal Y only | Design-stage precision/MDD |
| 1b | `summarize_event_support()` | Marginal Y only | Event counts by arm |
| 2 | `fit_ps_glm()` | No | Logistic regression PS |
| 2 | `fit_ps_superlearner()` | No | SuperLearner PS |
| 2 | `compute_ps_diagnostics()` | No | Overlap, ESS, SMDs |
| 2 | `checkpoint_balance()` | No | Check Point 2: balance decision |
| 2b | `tmle_candidate()` | No | Define a TMLE candidate specification |
| 2b | `expand_tmle_candidate_grid()` | No | Generate candidate grid |
| 2b | `validate_tmle_candidates()` | No | Validate candidate list |
| 2b | `run_plasmode_feasibility()` | Synthetic Y only | Plasmode simulation evaluation |
| 2b | `summarize_plasmode_results()` | No | Print plasmode metrics |
| 2b | `select_tmle_candidate()` | No | Select best TMLE specification |
| 2b | `lock_primary_tmle_spec()` | No | Lock selected spec in the lock |
| 2b | `get_primary_tmle_spec()` | No | Retrieve locked spec |
| 2b | `gate_check()` | No | GO / FLAG / STOP for plasmode |
| 3 | `run_negative_control()` | Negative-control Y only | Negative control analysis |
| 3 | `run_residual_confounding_stage()` | Negative-control Y only | Stage 3 wrapper |
| 3 | `checkpoint_residual_bias()` | No | Check Point 3: residual bias |
| Gate | `authorize_outcome_analysis()` | No | Pre-outcome GO/STOP gate |
| Gate | `assert_outcome_authorized()` | No | Gate assertion (errors if not GO) |
| 4 | `run_crude_workflow()` | Primary Y | Unadjusted risk difference |
| 4 | `run_match_workflow()` | Primary Y | PS matching |
| 4 | `run_iptw_workflow()` | Primary Y | Stabilised IPTW |
| 4 | `fit_tmle_treatment_mechanism()` | No | TMLE g-step (re-uses PS) |
| 4 | `fit_tmle_outcome_mechanism()` | Primary Y | TMLE Q-step |
| 4 | `run_tmle_targeting_step()` | Primary Y | TMLE targeting / fluctuation |
| 4 | `extract_tmle_estimate()` | Primary Y | Final TMLE estimate |
| 4 | `fit_tmle_candidate_set()` | Primary Y | Fit multiple TMLE specs |
| 4 | `fit_final_workflows()` | Primary Y | Run all Stage 4 workflows |
| 4 | `sensitivity_truncation()` | Primary Y | Truncation sensitivity |
| Any |`compute_evalue()` | No | E-value for unmeasured confounding |
| Any |`summarize_cleanroom_results()` | No | Cross-workflow comparison table |
| Any |`record_decision_log_entry()` | No | Add structured decision log entry |
| Any |`export_decision_log()` | No | Export decision log as data frame |
| Any |`summarize_stage_path()` | No | Compact stage narrative |
| Any |`create_audit_log()` | No | Initialise audit trail |
| Any |`record_stage()` | No | Append stage entry |
| Any |`record_checkpoint()` | No | Append checkpoint entry |
| Any |`export_audit_trail()` | No | Export trail as data frame |
| Any |`build_stage_manifest()` | No | Print stage path summary |
| Util | `mask_outcome()` | No | Mask outcome column with NA |
| Util | `unmask_outcome()` | No | Restore masked outcome |
| Wrapper | `run_clean_tmle()` | Primary Y | Full staged clean-room pipeline |
| Wrapper | `init_decision_log()` | No | Create empty decision log |
| Wrapper | `log_decision_entry()` | No | Append row to decision log |
| Wrapper | `save_decision_log()` | No | Write decision log to CSV |
| Plot | `love_plot()` | No | SMD dot-plot (weighted vs unweighted) |
| Plot | `ic_histogram()` | No | Influence curve histogram |
| Integ | `wrap_ps_fit()` | No | Wrap external PS scores into ps_fit |
| Integ | `fit_ps_parallel()` | No | SuperLearner PS with parallel cluster |
| Integ | `save_lock()` / `load_lock()` | No | Serialise / deserialise lock to RDS |
| Integ | `save_audit()` / `load_audit()` | No | Serialise / deserialise audit log |
| 3 | `run_negative_control_tmle()` | Negative-control Y only | TMLE-based negative control (doubly robust) |
| 4 | `run_matched_tmle()` | Primary Y | Modular TMLE on a matched subset |
| Integ | `sanitize_covariates()` | No | Clean covariates for SuperLearner |
| Integ | `create_analysis_lock_from_yaml()` | No | Lock from YAML config file |
| Plot | `clever_covariate_plot()` | No | H(A,W) distribution by group |
| Plot | `love_plot_threeway()` | No | Unweighted / IPTW / matched SMDs |
| Integ | `compute_matched_smds()` | No | SMDs on a matched subset |
| 1 | `attrition_table()` | No | Cohort flow / attrition documentation |
| 1 | `get_final_cohort()` | No | Extract filtered data from attrition table |
| 2b | `refine_ps_after_nco()` | Negative-control Y only | Iterative PS refinement after NCO signal |
| 2b | `run_plasmode_dq_stress()` | Synthetic Y only | DQ stress test (missingness, misclass, U) |
| 2b | `summarize_dq_degradation()` | No | DQ degradation comparison table |
| Gate | `gate_all()` | No | Composite GO/FLAG/STOP across checkpoints |

**Marginal Y only**: accesses event counts or marginal outcome support, not the A--Y association.
**Synthetic Y only**: generates simulated outcomes; does not use the real outcome association.
**Negative-control Y only**: accesses prespecified negative-control outcomes, not the primary outcome.
**Primary Y**: estimates the observed treatment--primary-outcome association and requires authorization.

---

## Utilities

```{r utils}
expit(0)            # 0.5
logit(0.5)          # 0
expit(logit(0.73))  # 0.73
```

---

## Stage 1a: Lock, Estimand, Plans

```{r lock}
lock <- create_analysis_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  sl_library = c("SL.glm", "SL.mean"),
  plasmode_reps = 40L,
  seed       = 42L
)

lock <- attach_estimand(lock,
  description          = "Treatment effect on 24-month event risk",
  contrast             = "risk_difference",
  statistical_estimand = "E_W[E(Y|A=1,W) - E(Y|A=0,W)]"
)

lock <- declare_sensitivity_plan(lock, "truncation",
  description = "Vary PS truncation",
  settings    = list(thresholds = c(0.01, 0.05, 0.10)))

lock <- define_negative_control(lock, "nc_outcome")

validate_analysis_lock(lock)
print(lock)
```

### Outcome Masking

```{r masking}
masked <- mask_outcome(lock)
all(is.na(masked$data[["event_24"]]))
unmasked <- unmask_outcome(masked, lock)
```

---

## Stage 1b: Cohort Adequacy (Check Point 1)

```{r cp1}
cp1 <- checkpoint_cohort_adequacy(lock)
print(cp1)
```

---

## Design-Stage Diagnostics

```{r design-precision}
dp <- estimate_design_precision(lock, target_mdd = 0.05)
print(dp)
summarize_event_support(lock)
```

---

## Stage 2a: Design diagnostics (Check Point 2)

```{r ps}
ps_fit <- fit_ps_glm(lock, truncate = 0.01)
diag   <- compute_ps_diagnostics(ps_fit)
print(diag)
```

```{r overlap, fig.cap = "PS overlap."}
plot(diag)
```

```{r cp2}
cp2 <- checkpoint_balance(diag, lock_hash = lock$lock_hash)
print(cp2)
```

---

## Stage 2b: Baseline plasmode candidate selection

```{r candidates}
# Define TMLE candidate specifications
candidates <- list(
  tmle_candidate("glm_t01", "GLM trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05)
)
# Or use the default grid:
# candidates <- expand_tmle_candidate_grid()
```

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(lock, tmle_candidates = candidates,
                                 effect_sizes = c(0.05, 0.10),
                                 reps = 40L)
print(plas)
```

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- lock_primary_tmle_spec(lock, selected)
print(selected)
```

---

## Stage 2c: Data-quality stress testing

`run_plasmode_dq_stress()` perturbs the synthetic outcome under
locked data-quality severity ranges. The severity values are not
chosen here; they come from external sources (validation studies,
data profiling, or expert elicitation) and are locked before the
stress test runs. The example below shows two scenarios with
locked severity values and the downstream summarisation and
gating calls.

```{r dq-stress-example, eval = FALSE}
# Severity values below are illustrative. In practice they are
# locked before this chunk runs and come from:
#   - validation studies (e.g. measured sens/spec for the outcome
#     phenotype against a gold-standard chart review),
#   - data profiling (e.g. observed missingness fractions), or
#   - expert elicitation (e.g. plausible OR ranges for an
#     unmeasured confounder).

dq_results <- run_plasmode_dq_stress(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05),
  reps            = 30L,
  data_quality_scenarios = list(
    # Scenario 1: outcome misclassification, two (sens, spec) pairs
    outcome_misclass = list(
      sensitivity = c(0.95, 0.90),
      specificity = c(0.99, 0.95)
    ),
    # Scenario 2: unmeasured confounding, two beta (OR) values
    unmeasured_confounding = list(
      U_prevalence   = 0.20,
      U_treatment_OR = c(1.5, 2.0),
      U_outcome_OR   = c(1.5, 2.0)
    )
  ),
  verbose = FALSE
)

# Summarise degradation relative to the baseline plasmode
deg <- summarize_dq_degradation(dq_results)
head(deg)

# Score all checkpoints together (GO / FLAG / STOP)
composite <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
print(composite)

# Or check the audit-trail-based pre-outcome gate
gate <- authorize_outcome_analysis(audit)
print(gate)
```

::: {.callout-warning}
## Common misuse

- Do not use DQ stress testing to claim causal validity.
- Do not interpret low-replicate examples inferentially.
- Do not treat truncation as purely numerical when it changes
  the effective target population.
- Do not use the package audit log as a substitute for
  institutional audit trails.
- Do not interpret a GO decision as proof that assumptions hold.
:::

---

## Stage 3: Negative-control checks (Check Point 3)

```{r nc}
nc  <- run_negative_control(lock, "nc_outcome", ps_fit)
cp3 <- checkpoint_residual_bias(nc, lock_hash = lock$lock_hash)
print(nc)
print(cp3)
```

---

## Audit Trail (initialise early for gate)

```{r audit}
audit <- create_audit_log(lock)
audit <- record_stage(audit, "Stage 1a", "Lock created")
audit <- record_checkpoint(audit, cp1)
audit <- record_checkpoint(audit, cp2)
audit <- record_checkpoint(audit, cp3)
```

## Gate and authorisation

```{r gate}
gate <- authorize_outcome_analysis(audit)
print(gate)
```

---

## Stage 4: Authorised primary analysis

```{r crude}
run_crude_workflow(lock)
```

```{r matching}
match_fit <- run_match_workflow(lock, ps_fit)
print(match_fit)
```

```{r iptw}
iptw_fit <- run_iptw_workflow(lock, ps_fit)
print(iptw_fit)
```

```{r tmle}
g_fit    <- fit_tmle_treatment_mechanism(lock, ps_fit)
Q_fit    <- fit_tmle_outcome_mechanism(lock, g_fit)
tmle_upd <- run_tmle_targeting_step(g_fit, Q_fit)
tmle_fit <- extract_tmle_estimate(tmle_upd)
print(tmle_fit)
```

```{r compare}
summarize_cleanroom_results(
  list(Matching = match_fit, IPTW = iptw_fit, TMLE = tmle_fit)
)
```

---

## Sensitivity and Diagnostics

```{r trunc-sens}
sensitivity_truncation(lock, thresholds = c(0.01, 0.05, 0.10))
```

```{r evalue}
compute_evalue(1.3, ci_bound = 1.05)
```

---

## Model Specification DSL

The `specify_models()` + `identify_*()` API provides a pipe-friendly
interface for time-to-event estimation functions.

```{r spec}
spec <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph") |>
  identify_subject(id)
print(spec)
```

---

## Time-to-Event Estimation

```{r ipwrisk, cache = TRUE}
fit_ipw <- estimate_ipwrisk(spec, risk_time = c(12, 24),
                             weight_type = "iptw", nboot = 20, seed = 42)
print(fit_ipw)
```

```{r ipwrisk-plot, fig.cap = "IPTW risk curves."}
plot(fit_ipw)
```

```{r gcomp}
spec_gc <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment)

fit_gc <- estimate_gcomprisk(spec_gc, risk_time = c(12, 24))
print(fit_gc)
```

```{r aipw}
spec_aipw <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph")

fit_aipw <- estimate_aipwrisk(spec_aipw, risk_time = c(12, 24))
print(fit_aipw)
```

```{r ipwhr}
hr_fit <- estimate_ipwhr(spec)
print(hr_fit)
```

```{r compare-tte}
compare_fits(IPW = fit_ipw, GComp = fit_gc, AIPW = fit_aipw, risk_time = 24)
```

---

## Tables and Weight Diagnostics

```{r table1}
make_table1(dat, vars = c("age", "sex", "biomarker", "comorbidity"),
            by = "treatment", smd = TRUE)
```

```{r table2}
make_table2(fit_ipw, risk_time = 24)
```

```{r wt-summary}
make_wt_summary_table(fit_ipw)
```

```{r extreme}
extreme_weights(fit_ipw, k = 5)
```

---

## Audit Trail (continued)

```{r audit-final}
audit <- record_stage(audit, "Stage 4", "Final estimation complete")
export_audit_trail(audit)
```

### Decision Log

```{r decision-log}
audit <- record_decision_log_entry(audit, "Stage 2", "model_specification",
                                   "Selected GLM PS model")
export_decision_log(audit)
summarize_stage_path(audit)
```

```{r manifest}
build_stage_manifest(audit)
```

---

## Real-World Pipeline Integration

When applying cleanTMLE to actual data, several practical concerns
arise: externally fitted PS, dirty covariate matrices, file-based
multi-stage scripts, and matched-cohort analyses.
This section demonstrates the integration utilities that address
those needs.

### Dirty covariates → clean matrix

Real registry data often has missing values, zero-variance columns,
or column names with special characters.
`sanitize_covariates()` handles all of these in one call:

```{r sanitize}
clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  verbose = TRUE)
cat("Covariates after cleaning:", paste(clean$covariates, collapse = ", "), "\n")
```

### Bring your own propensity scores

If PS were estimated outside cleanTMLE (parallel SuperLearner,
Bayesian model, external software), `wrap_ps_fit()` wraps them into
a `ps_fit` so all downstream functions work identically:

```{r wrap-ps}
my_ps <- predict(glm(treatment ~ age + sex + biomarker + comorbidity,
                      data = dat, family = binomial()), type = "response")
ext_ps_fit <- wrap_ps_fit(lock, ps_scores = my_ps, method = "custom_glm")
ext_diag   <- compute_ps_diagnostics(ext_ps_fit)
print(ext_diag$ess)
```

### Serialise lock and audit between stages

In production pipelines, each stage runs as a separate R script.
`save_lock()` / `load_lock()` serialise with automatic hash
re-validation on reload:

```{r save-load-lock}
# Round-trip: save, reload, verify
tmp <- tempfile("lock_", fileext = ".rds")
save_lock(lock, tmp)
lock_rt <- load_lock(tmp)
cat("Hash intact:", identical(lock$lock_hash, lock_rt$lock_hash), "\n")
unlink(tmp)
```

### TMLE-based negative controls

The default `run_negative_control()` uses IPTW.
`run_negative_control_tmle()` is doubly robust --- more consistent
with a TMLE-centred pipeline:

```{r nc-tmle}
nc_tmle <- run_negative_control_tmle(lock, "nc_outcome", ps_fit)
print(nc_tmle)
cat("Method:", nc_tmle$method, "\n")
```

### Matched-cohort TMLE

`run_matched_tmle()` runs the full TMLE update on a matched subset
without creating a separate lock:

```{r matched-tmle}
matched_idx <- c(which(dat$treatment == 1)[1:50],
                 which(dat$treatment == 0)[1:50])
matched_fit <- run_matched_tmle(lock, ps_fit, subset_idx = matched_idx,
                                override_clean_room = TRUE)
print(matched_fit)
```

### Diagnostic plots

```{r clever-plot, fig.cap = "Clever covariate H(A,W) by treatment group."}
clever_covariate_plot(ps_fit = ps_fit, lock = lock)
```

`compute_matched_smds()` + `love_plot_threeway()` compare balance
across all three adjustment strategies:

```{r threeway-love, fig.cap = "Unweighted vs IPTW vs matched SMDs."}
m_smds <- compute_matched_smds(dat, "treatment",
  c("age", "sex", "biomarker", "comorbidity"),
  subset_idx = matched_idx)
love_plot_threeway(diag, matched_smds = m_smds)
```

### Composite gating

`gate_all()` evaluates all checkpoints at once ---
convenient when the pipeline has many stages:

```{r gate-all}
composite <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
print(composite)
```

```{r session-info}
sessionInfo()
```
