---
title: "A Staged Clean-Room TMLE Analysis"
subtitle: "Worked Example with cleanTMLE"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{A Staged Clean-Room TMLE Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  warning    = FALSE,
  message    = FALSE
)
```

## Why staged clean-room analysis?

Observational comparative effectiveness studies are vulnerable to analytic
decisions that --- consciously or not --- are influenced by preliminary
knowledge of the treatment--outcome association.
Muntner et al. (2020) proposed a staged "clean-room" workflow in which
design decisions (variable selection, model specification, estimator
choice) are locked *before* the investigator accesses the observed
outcome data.
The approach enforces outcome blinding through software-mediated stage
gates, reducing degrees of freedom in the final comparative analysis.

**cleanTMLE** operationalises this workflow in R.
It is not a full protocol template or a regulatory submission tool;
it is a package that encodes estimand-oriented, stage-gated
causal analyses with TMLE as the primary estimator.
This vignette demonstrates the package on a simulated dataset,
walking through each stage from analytic specification to final
inference.

The stages follow the Muntner framework:
```
Stage 1a  Specify the analytic plan and lock it
Stage 1b  Assess cohort adequacy                   (Check Point 1)
Stage 2   Assess treatment-arm comparability via PS (Check Point 2)
Stage 3   Assess residual bias via negative controls(Check Point 3)
Gate      Pre-outcome authorization                 (Go / No-Go)
Stage 4   Conduct the comparative analysis
```

Each checkpoint produces a structured GO / FLAG / STOP decision.
The analysis may proceed to the next stage only after all preceding
checkpoints have been evaluated.
An audit trail accumulates entries automatically.

---

## Define the causal question and estimand

Before writing any analysis code we state the causal question in
target-trial-emulation style.

| Element | Specification |
|---------|---------------|
| **Population** | Adults with simulated disease (the `sim_func1()` cohort) |
| **Treatment strategies** | Treatment (A = 1) versus control (A = 0), assigned at baseline |
| **Follow-up** | 24 months from cohort entry |
| **Outcome** | Primary event occurring by 24 months (`event_24`) |
| **Causal contrast** | Average treatment effect on the risk-difference scale |

The causal estimand is $\psi = E[Y^{a=1}] - E[Y^{a=0}]$, the
population-level risk difference under two hypothetical interventions.
Because we work with observational data, the quantity we can identify
from the observed distribution is the *statistical estimand*

$$
\psi = E_W\!\bigl[E[Y \mid A=1, W] - E[Y \mid A=0, W]\bigr],
$$

which equals the causal parameter under the standard identification
assumptions (consistency, exchangeability, positivity).
The package estimates this statistical estimand; causal interpretation
relies on the investigator's subject-matter defence of those assumptions.

We choose the risk difference as the primary estimand because it has a
transparent absolute-scale interpretation, admits doubly-robust and
semiparametric-efficient estimation via TMLE, and is directly relevant
to clinical decision making.

---

## Stage 1a: Analysis specification and lock

```{r stage1a}
library(cleanTMLE)
set.seed(42)

dat <- sim_func1(n = 400, seed = 42)
```

The analysis lock captures the full specification --- treatment, outcome,
covariates, learner library, plasmode settings, and seed --- as an
immutable object.
Any downstream function that receives the lock can verify its integrity
against the stored hash.

```{r lock}
lock <- create_analysis_lock(
  data          = dat,
  treatment     = "treatment",
  outcome       = "event_24",
  covariates    = c("age", "sex", "biomarker", "comorbidity"),
  sl_library    = c("SL.glm", "SL.mean"),
  plasmode_reps = 40L,
  seed          = 42L
)
```

::: {.callout-tip}
## Optional: outcome masking

For analyses that require stricter blinding, `mask_outcome()` can
replace the real outcome column with `NA` values inside the lock,
preventing any code from accessing the outcome before Stage 4.
Call `unmask_outcome(lock, original_lock)` when the pre-outcome
gates have been passed and the outcome is needed.
In this vignette we proceed without masking for clarity, but the
functions are available for production workflows.
:::

We enrich the lock with the estimand, sensitivity plan, and a negative
control declaration.
These metadata do not alter the reproducibility hash (which depends only
on the core analytic specification) but provide the structured record
that a clean-room review requires.

```{r estimand}
lock <- attach_estimand(lock,
  description          = "Effect of treatment on 24-month event risk",
  population           = "Adults with simulated disease",
  treatment_strategies = c("Treatment", "Control"),
  outcome_label        = "Primary event by 24 months",
  followup             = "24 months",
  contrast             = "risk_difference",
  statistical_estimand = "E_W[E(Y|A=1,W) - E(Y|A=0,W)] under exchangeability + positivity"
)
```

```{r sensitivity-plan}
lock <- declare_sensitivity_plan(lock,
  label       = "truncation_sensitivity",
  description = "Re-estimate under alternate PS truncation thresholds",
  settings    = list(truncation = c(0.01, 0.05, 0.10))
)
```

```{r neg-control-decl}
lock <- define_negative_control(lock, "nc_outcome",
  description = "Binary outcome driven by covariates only; treatment has no causal effect"
)
```

```{r print-lock}
validate_analysis_lock(lock)
print(lock)
```

We initialise the audit log at this point so that every subsequent
stage is recorded.

```{r audit-init}
audit <- create_audit_log(lock)
audit <- record_stage(audit, "Stage 1a",
                      "Analysis lock created and estimand attached")
```

We also begin populating the decision log, which records the
rationale behind key analytic choices.

```{r dlog-lock}
audit <- record_decision_log_entry(audit,
  stage         = "Stage 1a",
  decision_type = "specification",
  description   = "Analysis lock created with GLM + Mean SL library and 40 plasmode reps",
  rationale     = "Minimal library chosen for vignette reproducibility"
)
```

---

## Stage 1b / Check Point 1: Cohort adequacy

Before any modelling, we confirm that the cohort has adequate size,
event counts, and positivity to support the planned analysis.
If this checkpoint fails, the analysis stops before any comparative
outcome modelling begins.

```{r cp1}
cp1 <- checkpoint_cohort_adequacy(lock,
  min_n_per_arm  = 50,
  min_events     = 30,
  min_prevalence = 0.01
)
print(cp1)
audit <- record_checkpoint(audit, cp1)
```

The decision is **`r cp1$decision`**.
With `r nrow(dat)` subjects and adequate events in both arms, we proceed.

### Design-stage diagnostics

Before moving to propensity-score modelling we run two additional
design-stage diagnostics.
`estimate_design_precision()` reports the detectable effect size
given the locked sample and event counts, optionally compared against
a target minimum detectable difference (MDD).
`summarize_event_support()` tabulates event counts by treatment arm
and covariate strata, flagging any cells with sparse support.

```{r design-precision}
dp <- estimate_design_precision(lock, target_mdd = 0.05)
print(dp)
```

```{r event-support}
es <- summarize_event_support(lock)
print(es)
```

---

## Stage 2 / Check Point 2: Treatment-arm comparability

Stage 2 estimates the propensity score and evaluates overlap, balance,
and effective sample size.
The outcome is never accessed at this stage: only covariates (W) and
treatment (A) are used.

### Propensity score estimation

```{r ps}
ps_fit <- fit_ps_glm(lock, truncate = 0.01)
print(ps_fit)
```

### Diagnostics

```{r ps-diag}
diag <- compute_ps_diagnostics(ps_fit)
print(diag)
```

```{r ps-overlap, fig.cap = "Propensity score overlap by treatment arm."}
plot(diag)
```

### Structured checkpoint

The `checkpoint_balance()` function converts diagnostics into a
GO / FLAG / STOP decision based on prespecified thresholds for
the maximum weighted SMD and minimum ESS percentage.

```{r cp2}
cp2 <- checkpoint_balance(diag,
  max_smd     = 0.10,
  min_ess_pct = 50,
  lock_hash   = lock$lock_hash
)
print(cp2)
audit <- record_checkpoint(audit, cp2)
```

The decision is **`r cp2$decision`**.
All covariates are well balanced after weighting, and the effective
sample size is adequate.

---

## Stage 2b: Outcome-blind TMLE specification selection

Before accessing the real outcome, we compare *prespecified TMLE
specifications* on plasmode-simulated data.
Each candidate varies the propensity-score truncation threshold and/or
the SuperLearner library used for nuisance estimation.
Plasmode simulation preserves the real covariate distribution and
treatment mechanism but generates synthetic outcomes with a known true
effect, allowing us to evaluate bias, RMSE, and coverage for each
candidate specification.

The purpose is *prespecified specification selection*, not performance
chasing: the selection rule (here, minimum RMSE) is declared in
advance, and the selected TMLE specification is locked into the
analysis lock and carried forward exactly to Stage 4 without
ad hoc modification.

### Define TMLE candidates

```{r tmle-candidates}
# Define candidate TMLE specifications that vary truncation
candidates <- list(
  tmle_candidate("glm_t01", "GLM, trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM, trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05),
  tmle_candidate("glm_mean_t01", "GLM+Mean, trunc=0.01",
                 g_library = c("SL.glm", "SL.mean"), truncation = 0.01)
)
```

### Plasmode feasibility evaluation

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05, 0.10),
  reps            = 40L,
  verbose         = FALSE
)
print(plas)
```

### Select and lock the primary TMLE specification

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- lock_primary_tmle_spec(lock, selected)
print(selected)
```

The selected TMLE specification for Stage 4 is
**`r selected$label`** (truncation = `r selected$truncation`).
This specification is now locked into the analysis lock and will be
used automatically by `fit_tmle_treatment_mechanism()` and
`fit_tmle_outcome_mechanism()`.

```{r audit-stage2b}
audit <- record_stage(audit, "Stage 2b",
  paste("Plasmode evaluation complete; selected TMLE spec:",
        selected$candidate_id))
```

```{r dlog-tmle-select}
audit <- record_decision_log_entry(audit,
  stage         = "Stage 2b",
  decision_type = "estimator_selection",
  description   = paste("Selected TMLE candidate:", selected$candidate_id),
  rationale     = "Minimum RMSE rule applied to plasmode simulation results"
)
```

---

## Stage 3 / Check Point 3: Residual bias assessment

Stage 3 assesses residual confounding using a negative control outcome ---
a variable that is known (or strongly believed) to be unaffected by
treatment.
If the same analytic pipeline detects an association with the negative
control, this suggests that confounders remain unbalanced after
adjustment.

The negative control outcome `nc_outcome` was generated from
covariates only, with no treatment effect, so any estimated association
is attributable to residual confounding.

The `run_residual_confounding_stage()` wrapper runs the negative control
analysis and generates the checkpoint in a single call, reducing
boilerplate and ensuring the two steps stay paired.

```{r stage3}
stage3 <- run_residual_confounding_stage(lock, ps_fit)
print(stage3)
cp3 <- stage3$checkpoint
audit <- record_checkpoint(audit, cp3)
```

The decision is **`r cp3$decision`**.
No significant association with the negative control was detected, so
we proceed to the pre-outcome authorization gate.

---

## Pre-outcome authorization gate

Before any code touches the real outcome we run a formal go/no-go
check.
`authorize_outcome_analysis()` inspects the audit trail and verifies
that every required checkpoint has been recorded with a GO (or, if
`allow_flag = TRUE`, a FLAG) decision.
If any checkpoint is missing or returned STOP, the gate refuses to
authorize and the analysis cannot proceed to Stage 4.

```{r gate}
gate <- authorize_outcome_analysis(audit)
print(gate)
audit <- record_checkpoint(audit, gate)
```

```{r dlog-gate}
audit <- record_decision_log_entry(audit,
  stage         = "Pre-outcome gate",
  decision_type = "authorization",
  description   = paste("Outcome analysis authorized:", gate$decision),
  rationale     = "All preceding checkpoints passed; gate opened automatically"
)
```

The gate decision is **`r gate$decision`** --- outcome data may now
be accessed.

---

## Stage 4: Final comparative analysis

Only after all preceding checkpoints pass do we access the real
outcome and run the locked analysis.

### Unadjusted benchmark

```{r crude}
crude <- run_crude_workflow(lock)
cat(sprintf("Crude RD: %.4f (95%% CI: %.4f, %.4f)\n",
            crude$estimate, crude$ci_lower, crude$ci_upper))
```

### PS matching

```{r matching}
match_fit <- run_match_workflow(lock, ps_fit)
print(match_fit)
```

### IPTW

```{r iptw}
iptw_fit <- run_iptw_workflow(lock, ps_fit)
print(iptw_fit)
```

### Modular TMLE (primary estimator)

TMLE is the prespecified primary estimator, using the specification
selected at Stage 2b and locked into the analysis lock.
The locked truncation threshold and learner libraries are applied
automatically by the modular TMLE functions --- no manual
specification is needed at this stage.

The package separates the TMLE procedure into explicit steps aligned
with the clean-room stages:

1. **Treatment mechanism** ($g$-model): estimated at Stage 2 without
   outcome access.  The locked truncation is applied automatically.
2. **Outcome mechanism** ($Q$-model): requires the real outcome and
   therefore belongs to Stage 4.  The locked Q-library is used.
3. **Targeting step**: fluctuates the initial outcome estimates using
   the clever covariate $H(A,W) = A/g(W) - (1-A)/(1-g(W))$.
4. **Estimate extraction**: computes the plug-in ATE, EIC-based SE,
   and confidence interval.

```{r tmle-g}
# Step 1: treatment mechanism (re-uses Stage 2 PS)
g_fit <- fit_tmle_treatment_mechanism(lock, ps_fit)
```

```{r tmle-Q}
# Step 2: outcome mechanism (FIRST access to the real outcome)
Q_fit <- fit_tmle_outcome_mechanism(lock, g_fit)
```

```{r tmle-target}
# Step 3: targeting step
tmle_upd <- run_tmle_targeting_step(g_fit, Q_fit)
```

```{r tmle-extract}
# Step 4: extract final estimate
tmle_fit <- extract_tmle_estimate(tmle_upd)
print(tmle_fit)
```

```{r audit-stage4}
audit <- record_stage(audit, "Stage 4",
  sprintf("Primary TMLE estimate: %.4f", tmle_fit$estimates$ATE$estimate))
```

### Cross-estimator comparison

```{r compare}
all_fits <- list(
  Crude    = list(estimate = crude$estimate, se = crude$se,
                  ci_lower = crude$ci_lower, ci_upper = crude$ci_upper,
                  p_value  = crude$p_value),
  Matching = match_fit,
  IPTW     = iptw_fit,
  `TMLE (primary)` = tmle_fit
)
summary_tbl <- summarize_cleanroom_results(
  list(Matching = match_fit, IPTW = iptw_fit, `TMLE (primary)` = tmle_fit)
)
summary_tbl
```

TMLE is the locked primary analysis; matching and IPTW serve as
contextual comparators.
Agreement across estimators increases confidence that the result is
not an artefact of a single modelling strategy.

---

## Sensitivity analysis

### Truncation sensitivity

The prespecified sensitivity plan varies the PS truncation threshold
to assess how much the IPTW estimate depends on extreme weights.

```{r sensitivity-trunc}
sens <- sensitivity_truncation(lock,
  thresholds = c(0.01, 0.025, 0.05, 0.10))
print(sens, row.names = FALSE)
```

### E-value

The E-value quantifies the minimum strength of unmeasured confounding
(on the risk-ratio scale) needed to explain away the observed
association.

```{r evalue}
ate <- tmle_fit$estimates$ATE
# Approximate risk ratio from TMLE risk difference
# Using crude arm risks for illustration
r0_hat <- crude$r0
rr_hat <- (r0_hat + ate$estimate) / r0_hat
rr_lo  <- (r0_hat + ate$ci_lower) / r0_hat
ev <- compute_evalue(rr_hat, ci_bound = min(rr_lo, 1/rr_lo))
cat(sprintf("E-value: %.2f (CI bound: %.2f)\n", ev["e_value"], ev["e_value_ci"]))
```

---

## Audit trail and reproducibility

The audit log documents every stage decision, its timestamp, and
the lock hash, providing a traceable record of the analytic path.

```{r audit-trail}
print(audit)
```

```{r stage-manifest}
build_stage_manifest(audit)
```

```{r audit-df}
trail_df <- export_audit_trail(audit)
knitr::kable(trail_df, caption = "Audit trail for the staged analysis.")
```

### Stage path narrative

`summarize_stage_path()` produces a human-readable narrative of the
stages traversed, their decisions, and the overall outcome.
This is useful for inclusion in study reports or supplementary
materials.

```{r stage-path}
summarize_stage_path(audit)
```

### Decision log

The decision log collects the rationale behind key analytic choices
made throughout the staged workflow.
`export_decision_log()` returns these entries as a data frame suitable
for archival or tabular display.

```{r decision-log}
dlog_df <- export_decision_log(audit)
knitr::kable(dlog_df, caption = "Decision log for the staged analysis.")
```

---

## Summary

This vignette demonstrated how **cleanTMLE** operationalises a staged
clean-room analysis:

1. The causal question and statistical estimand were defined first.
2. The analytic specification was locked before any outcome modelling.
3. Cohort adequacy, treatment comparability, and residual bias were
   assessed through structured checkpoints with prespecified
   decision thresholds.
4. Design precision and event support diagnostics confirmed the
   study was adequately powered.
5. The primary estimator (TMLE) was selected via outcome-blind
   plasmode simulation.
6. A formal pre-outcome authorization gate verified that all
   checkpoints passed before outcome data were accessed.
7. The final comparative analysis was conducted only after the gate
   opened.
8. Sensitivity analysis and an E-value were computed as prespecified.
9. An audit trail, stage path narrative, and decision log document
   the full analytic path and the rationale behind key choices.

The package does not enforce institutional governance (role separation,
independent review boards, blinded data managers); those are
organisational responsibilities.
What it does provide is software scaffolding that makes the staged
workflow reproducible, auditable, and less susceptible to unintentional
outcome-informed decision making.

```{r session-info}
sessionInfo()
```
