---
title: "Get started with cleanTMLE"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    self-contained: true
vignette: >
  %\VignetteIndexEntry{Get started with cleanTMLE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      warning = FALSE, message = FALSE)
library(cleanTMLE)
```

cleanTMLE answers a comparative-effectiveness question in two moves:
first it decides, before the outcome is examined, whether the design
can support the comparison and with which estimand; then it estimates
with targeted maximum likelihood. This vignette runs the shortest
honest path through that workflow, five verbs on a simulated cohort.
The full staged workflow, with the candidate grid, the data-quality
stress test, and the negative-control ladder, is in the *Full
workflow* article.

Read the simulated cohort as a pharmaco-epidemiology study: a new
medication (`treatment`) started at enrolment versus standard care, a
binary event at 24 months (`event_24`), and four baseline covariates.

```{r data}
dat <- sim_func1(n = 1000, seed = 42)
str(dat[, c("treatment", "event_24", "age", "sex", "biomarker")])
```

## 1. Lock the specification

`create_analysis_lock()` seals the analytic specification, data,
variable roles, learner library, and seed, behind a SHA-256
fingerprint, before any outcome is read. The returned lock is the one
object every later verb takes. Since 0.3.0 the lock's data frame holds
design data only; the primary outcome lives in a separate store that
only the Stage 4 estimators join back.

```{r lock}
lock <- create_analysis_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  seed       = 42)
print(lock)
```

Any later change to the design data, the roles, or the declarations
invalidates the fingerprint, so the review team can verify that what
was estimated is what was declared.

## 2. Fit the propensity score

`fit_ps()` fits the treatment mechanism on the lock. The
SuperLearner method is the default in real studies; the plain
logistic model keeps this vignette fast.

```{r ps}
ps <- fit_ps(lock, method = "glm")
print(ps)
```

## 3. Assess support

`assess_support()` turns the fitted propensity into the design team's
overlap verdict: the share of patients outside the support band, the
weight tails, the effective sample size, and a graded reading
(PASS, FLAG, SEVERE, FAIL). Everything in it is computable without the
outcome.

```{r support}
sup <- assess_support(ps)
print(sup)
```

## 4. Map the feasible estimands

Poor overlap does not have to end a study; it often means the design
answers a different question than the one first asked.
`estimand_feasibility()` walks the estimand ladder, from the ATE
through trimmed, treated-only, overlap, and matched estimands, and
reports which of them this design can deliver.

```{r feasibility}
feas <- estimand_feasibility(ps)
print(feas)
```

## 5. Estimate

`estimate_effect()` is the single front door for Stage 4: it joins the
outcome back from the store, fits the declared estimand with the
requested estimator, and returns the risk difference with its
influence-function confidence interval.

```{r estimate}
fit <- estimate_effect(lock, ps_fit = ps, estimand = "ATE",
                       estimator = "tmle",
                       sl_library = c("SL.glm", "SL.mean"))
print(fit)
```

## Where to go next

The pre-outcome half of a real study has more to it than these five
verbs, and each addition is one more verb on the same lock:

- `declare_negative_controls()` and `declare_estimand_ladder()` put
  the decision rules on the lock, fingerprinted, before outcome
  access.
- `define_candidates()`, `stress_test()`, and `select_candidate()`
  choose the TMLE specification on simulated outcomes and grade it
  under declared data-quality threats, with a locked GO / FLAG / STOP
  verdict and tipping points.
- `negative_control_ladder()` and `check_process_indicators()` screen
  for residual confounding and collider structure.
- `design_report()` and `export_design_log()` assemble what the
  review team reads, and the decision record it releases.

The *Full workflow* article runs all sixteen verbs on this same
cohort.
