---
title: "A Staged Outcome-Blind TMLE Workflow with cleanTMLE"
subtitle: "Worked Example"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{A Staged Outcome-Blind TMLE Workflow with cleanTMLE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  warning    = FALSE,
  message    = FALSE
)
library(cleanTMLE)
```

## Purpose of this vignette

This vignette demonstrates a staged workflow for an observational
comparative-effectiveness analysis on a simulated dataset. Design
and estimator-selection decisions are made before the observed
treatment-outcome association is examined; the goal is reproducible
prespecification and a structured documentation of analytic
decisions that an external reviewer can follow.

In this manuscript and vignette, *outcome-blind* means that
decisions about cohort construction, candidate estimators,
simulation scenarios, decision thresholds, and primary-estimator
selection are made without inspecting the observed
treatment-outcome association. Synthetic outcomes generated for
plasmode simulations may be used; the marginal mean of the
observed outcome on covariates is also used internally to fit the
plasmode generator. The *pre-outcome decision point* is the
juncture at which the prespecified diagnostics are reviewed and the
analysis is classified as proceed, pause for review, or stop
before the observed outcome is unmasked for the comparative
analysis.

cleanTMLE is not a substitute for target-trial design, source-data
validation, phenotype validation, independent oversight, or
subject-matter assessment of identification assumptions. The
worked example below shows reproducible prespecification, not a
regulatory submission template.

## Conceptual workflow

The vignette proceeds in the following conceptual steps:

1. **Stage 0 (precondition)**: feasibility assessment, target-trial
   specification, and protocol / SAP drafting. This is largely
   external work conducted before any cleanTMLE code is run.
2. **Stage 1a**: lock the estimand, covariates, learners, candidate
   grid, and decision thresholds.
3. **Stage 1b**: assess cohort adequacy and marginal event support.
4. **Stage 2a**: propensity score, balance, overlap, ESS, and
   design diagnostics.
5. **Stage 2b**: baseline plasmode candidate selection.
6. **Stage 2c**: plasmode data-quality (DQ) stress testing as a
   quantitative pre-outcome supplement to the fit-for-purpose review.
7. **Stage 3 (optional)**: negative-control outcome checks.
8. **Pre-outcome decision**: GO / FLAG / STOP.
9. **Stage 4**: authorized primary outcome analysis.
10. Post-outcome sensitivity analyses, interpretation, and archived
    analysis record.

The Muntner et al. (2020) clean-room governance framework
(role separation, restricted access, protocols, audit) is the
source of the staging idea and is the broader notion of a clean
room. cleanTMLE is the *software layer* of that workflow: it
records the analysis lock, runs the design diagnostics and the
plasmode evaluation, executes the DQ stress test, applies the
pre-outcome decision rule, and writes the audit trail. It does
not by itself implement clean-room personnel governance. The canonical stage labels used
throughout this vignette are:

```
Stage 0    Feasibility, target-trial specification, protocol / SAP drafting
Stage 1a   Lock estimand, covariates, learners, candidate grid, thresholds
Stage 1b   Cohort adequacy and marginal event support
Stage 2a   PS, balance, overlap, ESS, design diagnostics
Stage 2b   Baseline plasmode candidate selection
Stage 2c   DQ stress testing
Stage 3    Optional negative-control outcome checks
Pre-outcome decision: GO / FLAG / STOP
Stage 4    Authorized primary outcome analysis
Post-outcome: sensitivity analyses and interpretation
```

We use *proceed*, *proceed-with-qualification*, and *do-not-run*
in prose; the software shorthand for the same labels is GO, FLAG,
and STOP.

::: {.callout-warning}
This vignette demonstrates the software workflow. It does not
establish exchangeability, measurement validity, phenotype
validity, linkage accuracy, or regulatory acceptability.
:::

### Stage 0 as precondition

Stage 0 covers the feasibility assessment, target-trial
specification, and protocol / SAP drafting that precede any
analytic code. These activities are typically conducted outside
the software layer, often involving subject-matter experts, data
custodians, and study sponsors. cleanTMLE assumes Stage 0 is
complete: the causal question, eligible population, treatment
strategies, follow-up window, and primary estimand are already
specified in a written protocol before Stage 1a begins. The
remainder of this vignette demonstrates Stages 1a through 4.

---

## Define the causal question and estimand

Before writing any analysis code we state the causal question in
target-trial-emulation style.

| Element | Specification |
|---------|---------------|
| **Population** | Adults with simulated disease (the `sim_func1()` cohort) |
| **Treatment strategies** | Treatment (A = 1) versus control (A = 0), assigned at baseline |
| **Follow-up** | 24 months from cohort entry |
| **Outcome** | Primary event occurring by 24 months (`event_24`) |
| **Causal contrast** | Average treatment effect on the risk-difference scale |

The causal estimand is $\psi = E[Y^{a=1}] - E[Y^{a=0}]$, the
population-level risk difference under two hypothetical interventions.
Because we work with observational data, the quantity we can identify
from the observed distribution is the *statistical estimand*

$$
\psi = E_W\!\bigl[E[Y \mid A=1, W] - E[Y \mid A=0, W]\bigr],
$$

which equals the causal parameter under the standard identification
assumptions (consistency, exchangeability, positivity).
The package estimates this statistical estimand; causal interpretation
relies on the investigator's subject-matter defence of those assumptions.

We choose the risk difference as the primary estimand because it has a
transparent absolute-scale interpretation, admits doubly-robust and
semiparametric-efficient estimation via TMLE, and is directly relevant
to clinical decision making.

---

## From target trial specification to event-process classification

Once the target trial and the statistical estimand are stated,
post-baseline events should not be assigned a generic role of
"censoring." Each event should be mapped to the causal question,
to the corresponding ICH E9(R1) intercurrent-event strategy where
applicable, and to the operational role it plays in estimation.
The package supports this mapping with three structured helpers:
`clean_event_process_table()`, `clean_target_population()`, and
`clean_missing_data_plan()`.

```{r event-process-vignette}
events <- clean_event_process_table(list(
  list(event_name = "Primary event by 24 months",
       event_variable = "event_24",
       role_in_estimand = "outcome event",
       primary_handling = "modelled directly"),
  list(event_name = "Death from other causes",
       event_variable = "death_other",
       role_in_estimand = "competing event",
       ICH_E9R1_strategy = "composite or competing-risk strategy",
       primary_handling = "competing-risk cumulative incidence",
       justification = "Death precludes measurement of the primary event"),
  list(event_name = "Loss to follow-up",
       event_variable = "lost_fu",
       role_in_estimand = "missing outcome process",
       primary_handling = "IPCW",
       identification_assumption = "MAR conditional on baseline covariates"),
  list(event_name = "Treatment switching",
       event_variable = "treat_switch",
       role_in_estimand = "intercurrent event",
       ICH_E9R1_strategy = "treatment-policy strategy",
       primary_handling = "kept under assigned strategy")
), censoring_handling = "IPCW for missing outcome")
print(events)
```

```{r target-pop-vignette}
tp <- clean_target_population(
  target_population = "full eligible population",
  reference_group   = "Control",
  rationale         = "Marginal effect under each treatment strategy",
  implications_for_interpretation = "ATE on the risk-difference scale")
print(tp)
```

```{r missing-plan-vignette}
plan <- clean_missing_data_plan(list(
  "baseline covariate missingness" = list(
    timing = "baseline", presumed_mechanism = "MCAR",
    handling_primary = "median imputation",
    nuisance_model_used = "n/a"),
  "outcome missingness" = list(
    timing = "follow-up", presumed_mechanism = "MAR",
    handling_primary = "IPCW-TMLE",
    nuisance_model_used = "SuperLearner response model",
    identification_assumption = "MAR conditional on (A, W)"),
  "censoring or loss to follow-up" = list(
    timing = "follow-up", presumed_mechanism = "administrative",
    handling_primary = "right-censoring at end of follow-up window")))
print(plan)
```

---

## Stage 1a: Record the analysis specification (lock)

The analysis specification records the treatment, outcome,
covariates, candidate estimators, learner library, simulation
settings, and random seed. Later steps refer back to this object
so that the final analysis can be compared with the prespecified
plan, and so that an external reviewer can reconcile the recorded
inputs with the analyses that were run.

```{r stage1a}
library(cleanTMLE)
set.seed(42)

dat <- sim_func1(n = 400, seed = 42)
```

### Covariate sanitisation

Before locking the specification, we sanitise covariates to ensure they
are safe for SuperLearner: impute missing values, drop zero-variance
columns, and fix formula-unsafe names.

```{r sanitize}
clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  verbose = TRUE)
dat <- clean$data
covs <- clean$covariates
```

```{r lock}
lock <- create_analysis_lock(
  data          = dat,
  treatment     = "treatment",
  outcome       = "event_24",
  covariates    = covs,
  sl_library    = c("SL.glm", "SL.mean"),
  plasmode_reps = 40L,
  seed          = 42L
)
```

::: {.callout-tip}
## Optional: outcome masking

For analyses that require stricter blinding, `mask_outcome()` can
replace the real outcome column with `NA` values inside the lock,
preventing any code from accessing the outcome before Stage 4.
Call `unmask_outcome(lock, original_lock)` when the pre-outcome
gates have been passed and the outcome is needed.
In this vignette we proceed without masking for clarity, but the
functions are available for production workflows.
:::

We enrich the lock with the estimand, sensitivity plan, and a negative
control declaration.
These metadata do not alter the reproducibility hash (which depends only
on the core analytic specification) but provide the structured record
that an outcome-blind staged review requires.

```{r estimand}
lock <- attach_estimand(lock,
  description          = "Effect of treatment on 24-month event risk",
  population           = "Adults with simulated disease",
  treatment_strategies = c("Treatment", "Control"),
  outcome_label        = "Primary event by 24 months",
  followup             = "24 months",
  contrast             = "risk_difference",
  statistical_estimand = "E_W[E(Y|A=1,W) - E(Y|A=0,W)] under exchangeability + positivity"
)
```

```{r sensitivity-plan}
lock <- declare_sensitivity_plan(lock,
  label       = "truncation_sensitivity",
  description = "Re-estimate under alternate PS truncation thresholds",
  settings    = list(truncation = c(0.01, 0.05, 0.10))
)
```

```{r neg-control-decl}
lock <- define_negative_control(lock, "nc_outcome",
  description = "Binary outcome driven by covariates only; treatment has no causal effect"
)
```

```{r print-lock}
validate_analysis_lock(lock)
print(lock)
```

We initialise the audit log at this point so that every subsequent
stage is recorded.

```{r audit-init}
audit <- create_audit_log(lock)
audit <- record_stage(audit, "Stage 1a",
                      "Analysis lock created and estimand attached")
```

### File-based pipeline support

In production pipelines each stage runs as a separate script.
`save_lock()` / `load_lock()` handle serialisation with automatic
hash re-validation, eliminating the boilerplate of manual RDS
save/load and integrity checks.

```{r save-lock}
# Round-trip demonstration: save, load, verify integrity
tmp_lock_path  <- tempfile("lock_",  fileext = ".rds")
tmp_audit_path <- tempfile("audit_", fileext = ".rds")

save_lock(lock, tmp_lock_path)
save_audit(audit, tmp_audit_path)

# Reload --- load_lock() automatically re-validates the hash
lock_reloaded  <- load_lock(tmp_lock_path)
audit_reloaded <- load_audit(tmp_audit_path)
cat("Lock hash match:", identical(lock$lock_hash, lock_reloaded$lock_hash), "\n")
cat("Audit entries match:", length(audit$entries) == length(audit_reloaded$entries), "\n")

# Clean up temp files
unlink(c(tmp_lock_path, tmp_audit_path))
```

In a real multi-stage pipeline each stage script would save to a
shared results directory (e.g., `save_lock(lock, "results/stage1_lock.rds")`)
and the next stage script would `load_lock("results/stage1_lock.rds")`
at its top.

We also begin populating the decision log, which records the
rationale behind key analytic choices.

```{r dlog-lock}
audit <- record_decision_log_entry(audit,
  stage         = "Stage 1a",
  decision_type = "specification",
  description   = "Analysis lock created with GLM + Mean SL library and 40 plasmode reps",
  rationale     = "Minimal library chosen for vignette reproducibility"
)
```

---

## Stage 1b: Cohort adequacy and marginal event support (Check Point 1)

Before any modelling, we confirm that the cohort has adequate size,
event counts, and positivity to support the planned analysis.
If this checkpoint fails, the analysis stops before any comparative
outcome modelling begins.

```{r cp1}
cp1 <- checkpoint_cohort_adequacy(lock,
  min_n_per_arm  = 50,
  min_events     = 30,
  min_prevalence = 0.01
)
print(cp1)
audit <- record_checkpoint(audit, cp1)
```

The decision is **`r cp1$decision`**.
With `r nrow(dat)` subjects and adequate events in both arms, we proceed.

### Design-stage diagnostics

Before moving to propensity-score modelling we run two additional
design-stage diagnostics.
`estimate_design_precision()` reports the detectable effect size
given the locked sample and event counts, optionally compared against
a target minimum detectable difference (MDD).
`summarize_event_support()` tabulates event counts by treatment arm
and covariate strata, flagging any cells with sparse support.

```{r design-precision}
dp <- estimate_design_precision(lock, target_mdd = 0.05)
print(dp)
```

::: {.callout-note}
The MDD report may show "Feasible: NO" with this small simulated
sample (n = `r nrow(dat)`).
This is expected --- the vignette uses a deliberately small dataset
for fast rendering.
In practice you would either increase sample size or relax the
target MDD.
The MDD check is **informational** and does not trigger STOP; the
pipeline continues.
:::

```{r event-support}
es <- summarize_event_support(lock)
print(es)
```

---

## Stage 2a: PS, balance, overlap, ESS, design diagnostics (Check Point 2)

Stage 2a estimates the propensity score and evaluates overlap, balance,
and effective sample size.
The outcome is never accessed at this stage: only covariates (W) and
treatment (A) are used.

### Propensity score estimation

```{r ps}
ps_fit <- fit_ps_glm(lock, truncate = 0.01)
print(ps_fit)
```

::: {.callout-tip}
## Using external or parallel PS

If you have PS estimated outside cleanTMLE (e.g., via a parallel
SuperLearner cluster), wrap them with `wrap_ps_fit()`:

```r
cl <- parallel::makeCluster(2, type = "PSOCK")
ps_fit <- fit_ps_parallel(lock, cluster = cl, truncate = 0.01)
parallel::stopCluster(cl)

# Or bring-your-own PS:
ps_fit <- wrap_ps_fit(lock, ps_scores = my_ps, method = "external_sl")
```

All downstream functions (`compute_ps_diagnostics`, `checkpoint_balance`,
`run_iptw_workflow`, etc.) work identically with wrapped PS objects.
:::

### Diagnostics

```{r ps-diag}
diag <- compute_ps_diagnostics(ps_fit)
print(diag)
```

```{r ps-overlap, fig.cap = "Propensity score overlap by treatment arm."}
plot(diag)
```

### Structured checkpoint

The `checkpoint_balance()` function converts diagnostics into a
GO / FLAG / STOP decision based on prespecified thresholds for
the maximum weighted SMD and minimum ESS percentage.

```{r cp2}
cp2 <- checkpoint_balance(diag,
  max_smd     = 0.10,
  min_ess_pct = 50,
  lock_hash   = lock$lock_hash
)
print(cp2)
audit <- record_checkpoint(audit, cp2)
```

The decision is **`r cp2$decision`**.
All covariates are well balanced after weighting, and the effective
sample size is adequate.

---

## Stage 2b: Outcome-blind TMLE specification selection

Before accessing the real outcome, we compare *prespecified TMLE
specifications* on plasmode-simulated data.
Each candidate varies the propensity-score truncation threshold and/or
the SuperLearner library used for nuisance estimation.
Plasmode simulation preserves the real covariate distribution and
treatment mechanism but generates synthetic outcomes with a known true
effect, allowing us to evaluate bias, RMSE, and coverage for each
candidate specification.

The purpose is *prespecified specification selection*, not performance
chasing: the selection rule (here, minimum RMSE) is declared in
advance, and the selected TMLE specification is locked into the
analysis lock and carried forward exactly to Stage 4 without
ad hoc modification.

### Define TMLE candidates

```{r tmle-candidates}
# Define candidate TMLE specifications that vary truncation
candidates <- list(
  tmle_candidate("glm_t01", "GLM, trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM, trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05),
  tmle_candidate("glm_mean_t01", "GLM+Mean, trunc=0.01",
                 g_library = c("SL.glm", "SL.mean"), truncation = 0.01)
)
```

### Plasmode feasibility evaluation

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05, 0.10),
  reps            = 40L,
  verbose         = FALSE
)
print(plas)
```

### Select and lock the primary TMLE specification

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- lock_primary_tmle_spec(lock, selected)
print(selected)
```

The selected TMLE specification for Stage 4 is
**`r selected$label`** (truncation = `r selected$truncation`).
This specification is now locked into the analysis lock and will be
used automatically by `fit_tmle_treatment_mechanism()` and
`fit_tmle_outcome_mechanism()`.

## Stage 2c: Plasmode data-quality stress test

The DQ stress test is a **quantitative pre-outcome supplement to
the fit-for-purpose review**, not a formal quantitative bias
analysis (QBA) and not a substitute for source-data validation or
post-outcome QBA. The intended sequence is:

1. Fit-for-purpose review identifies plausible threats.
2. Threats are translated into quantitative severity ranges.
3. Severity ranges are locked before primary outcome access.
4. Synthetic outcomes are generated and perturbed.
5. Candidates are evaluated using the same metrics as baseline
   plasmode selection.
6. The same GO / FLAG / STOP rule is applied.

The DQ stress test documents how candidate estimators behave under
the locked threat range; it does not validate the data or replace
post-outcome QBA.

Four mechanisms are supported:

1. *Covariate missingness*: a fraction of values is set to `NA` and
   imputed with the column median.
2. *Treatment misclassification*: an asymmetric pair of (sensitivity,
   specificity) is applied to A.
3. *Outcome misclassification*: an asymmetric pair of (sensitivity,
   specificity) is applied to Y, optionally per treatment arm.
4. *Unmeasured confounding*: a latent binary U is sampled and shifts
   both the treatment mechanism (via a real PS model fitted on the
   lock data) and the outcome probability by user-specified odds
   ratios.

```{r dq-stress, cache = TRUE, eval = FALSE}
dq_results <- run_plasmode_dq_stress(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05),
  reps            = 30L,
  data_quality_scenarios = list(
    covariate_missingness = list(
      fractions = c(0.05, 0.10, 0.20)
    ),
    treatment_misclass = list(
      sensitivity = c(0.99, 0.95, 0.90),
      specificity = c(0.99, 0.95, 0.90)
    ),
    outcome_misclass = list(
      sensitivity = c(0.95, 0.90),
      specificity = c(0.99, 0.95)
    ),
    unmeasured_confounding = list(
      U_prevalence   = 0.20,
      U_treatment_OR = c(1.5, 2.0),
      U_outcome_OR   = c(1.5, 2.0)
    )
  ),
  verbose = FALSE
)

deg <- summarize_dq_degradation(dq_results)
head(deg)
```

`summarize_dq_degradation()` returns a long table comparing each
degraded run with the baseline (no degradation): bias change, RMSE
ratio, coverage drop. A robust candidate produces a flat curve; a
fragile candidate shows a cliff. The table can be filtered to the
candidate locked in by `select_tmle_candidate()` and reviewed
alongside the SPIFD2 fit-for-purpose narrative before the gate is
scored.

The DQ stress test is the package's distinguishing contribution
relative to prior outcome-blind plasmode work: standard plasmode
selects the candidate with the best baseline RMSE, while
`run_plasmode_dq_stress()` allows the analyst to select the
candidate that stays inside the gate thresholds across the
plausible range of every realistic data threat.

```{r audit-stage2b}
audit <- record_stage(audit, "Stage 2b",
  paste("Plasmode evaluation complete; selected TMLE spec:",
        selected$candidate_id))
```

```{r dlog-tmle-select}
audit <- record_decision_log_entry(audit,
  stage         = "Stage 2b",
  decision_type = "estimator_selection",
  description   = paste("Selected TMLE candidate:", selected$candidate_id),
  rationale     = "Minimum RMSE rule applied to plasmode simulation results"
)
```

---

## Stage 3 (optional): Negative-control outcome checks

A negative-control outcome (NCO) screen is an optional pre-outcome
checkpoint that sits between the Stage 2c DQ stress test and the
pre-outcome decision. The NCO screen assesses residual
confounding using a negative-control outcome --- a variable that is
known (or strongly believed) to be unaffected by treatment.
If the same analytic pipeline detects an association with the negative
control, this suggests that confounders remain unbalanced after
adjustment.

The negative control outcome `nc_outcome` was generated from
covariates only, with no treatment effect, so any estimated association
is attributable to residual confounding.

The `run_residual_confounding_stage()` wrapper runs the negative control
analysis and generates the checkpoint in a single call, reducing
boilerplate and ensuring the two steps stay paired.

```{r stage3}
stage3 <- run_residual_confounding_stage(lock, ps_fit)
print(stage3)
cp3 <- stage3$checkpoint
audit <- record_checkpoint(audit, cp3)
```

### TMLE-based negative controls (doubly robust)

The default `run_negative_control()` uses IPTW.
`run_negative_control_tmle()` uses the four-step TMLE pipeline,
which is doubly robust and more appropriate for a package built
around TMLE.

```{r nc-tmle}
nc_tmle <- run_negative_control_tmle(lock, "nc_outcome", ps_fit)
cat(sprintf("TMLE NC estimate: %.5f  (p = %.4f, method: %s)\n",
            nc_tmle$estimate, nc_tmle$p_value, nc_tmle$method))
```

The decision is **`r cp3$decision`**.
No significant association with the negative control was detected, so
we proceed to the pre-outcome authorization gate.

---

## Pre-outcome decision

The pre-outcome decision summarises whether the prespecified
diagnostics support proceeding to outcome analysis, require
additional review, or indicate that the study should stop or be
redesigned. `authorize_outcome_analysis()` inspects the audit
trail and verifies that every required checkpoint has been
recorded with a *proceed* (GO) or, when `allow_flag = TRUE`,
*pause for review* (FLAG) classification. If any checkpoint is
missing or returned *stop* (STOP), the prespecified workflow is
not complete and outcome-analysis functions will not run.

`gate_all()` evaluates all checkpoint objects at once and returns
the composite classification:

```{r gate-all}
composite <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
print(composite)
```

```{r gate}
gate <- authorize_outcome_analysis(audit)
print(gate)
audit <- record_checkpoint(audit, gate)
```

```{r dlog-gate}
audit <- record_decision_log_entry(audit,
  stage         = "Pre-outcome decision",
  decision_type = "authorization",
  description   = paste("Outcome analysis classification:", gate$decision),
  rationale     = "All preceding checkpoints passed."
)
```

The pre-outcome classification is **`r gate$decision`**, and
outcome-analysis functions may now run on the unmasked data.

### What if a checkpoint fails?

In the example above every checkpoint returned GO.
In practice, poor overlap or residual confounding can trigger FLAG or
STOP decisions.
The block below constructs a synthetic failure to show how the pipeline
responds:

```{r failure-demo}
# Simulate a STOP scenario: a checkpoint with very high SMD
fake_diag <- diag
fake_diag$smds$smd_weighted <- c(0.25, 0.30, 0.15, 0.08)  # badly imbalanced
fake_cp2 <- checkpoint_balance(fake_diag, max_smd = 0.10,
                                lock_hash = lock$lock_hash)
cat(sprintf("Fake CP2 decision: %s\n", fake_cp2$decision))
cat(sprintf("Rationale: %s\n", fake_cp2$rationale))

# gate_all() blocks when a STOP is present:
fake_composite <- gate_all(cp1, fake_cp2, cp3, allow_flag = FALSE)
cat(sprintf("\nComposite gate: %s\n", fake_composite$decision))
cat(sprintf("Rationale: %s\n", fake_composite$rationale))
```

When the gate returns STOP, no outcome code should execute.
In `run_clean_tmle()` this is enforced automatically; in the modular
workflow the analyst checks the gate decision and branches accordingly.

---

## Outcome-blind diagnostics before fitting the final estimator

Before the outcome is unmasked, we report a compact set of
diagnostics that support assessment of identification and
estimator stability: treatment overlap, censoring weights (when
present), missingness weights (when applicable), weighted SMDs,
and effective sample size. These diagnostics should be conducted
and the prespecified failure thresholds applied before looking at
outcomes when the workflow allows.

```{r outcome-blind-diagnostics, eval = exists("ps_fit")}
A <- lock$data[[lock$treatment]]
ps <- ps_fit$ps
w  <- ifelse(A == 1, 1 / ps, 1 / (1 - ps))

W_for_smd <- lock$data[, lock$covariates, drop = FALSE]
weight_diag <- clean_weight_diagnostics(weights = w, treatment = A,
                                         covariates = W_for_smd,
                                         max_weight_threshold = 10,
                                         ess_floor = 0.4 * length(w))
print(weight_diag)
```

The maximum-weight threshold (10) and the ESS floor (40 % of N)
are conservative defaults; the prespecified analysis plan should
record values that are appropriate for the study's anticipated
overlap, weight distribution, and sample size. When the study
uses censoring or missingness weights, `clean_weight_diagnostics()`
can be called separately on each weight vector.

---

## Stage 4: Final comparative analysis

Only after all preceding checkpoints pass do we access the real
outcome and run the locked analysis.

### Unadjusted benchmark

```{r crude}
crude <- run_crude_workflow(lock)
cat(sprintf("Crude RD: %.4f (95%% CI: %.4f, %.4f)\n",
            crude$estimate, crude$ci_lower, crude$ci_upper))
```

### PS matching

```{r matching}
match_fit <- run_match_workflow(lock, ps_fit)
print(match_fit)
```

### IPTW

```{r iptw}
iptw_fit <- run_iptw_workflow(lock, ps_fit)
print(iptw_fit)
```

### TMLE in four steps (primary estimator)

TMLE is the prespecified primary estimator, using the specification
selected at Stage 2b and locked into the analysis lock.  The locked
truncation threshold and learner libraries are applied automatically
by the four TMLE functions --- no manual specification is needed at
this stage.

The package separates the TMLE procedure into four explicit steps
aligned with the outcome-blind stages:

1. **Treatment mechanism** ($g$-model): estimated at Stage 2 without
   outcome access.  The locked truncation is applied automatically.
2. **Outcome mechanism** ($Q$-model): requires the real outcome and
   therefore belongs to Stage 4.  The locked Q-library is used.
3. **Targeting step**: fluctuates the initial outcome estimates using
   the clever covariate $H(A,W) = A/g(W) - (1-A)/(1-g(W))$.
4. **Estimate extraction**: computes the plug-in ATE, EIC-based SE,
   and confidence interval.

```{r tmle-g}
# Step 1: treatment mechanism (re-uses Stage 2 PS)
g_fit <- fit_tmle_treatment_mechanism(lock, ps_fit)
```

```{r tmle-Q}
# Step 2: outcome mechanism (FIRST access to the real outcome)
Q_fit <- fit_tmle_outcome_mechanism(lock, g_fit)
```

```{r tmle-target}
# Step 3: targeting step
tmle_upd <- run_tmle_targeting_step(g_fit, Q_fit)
```

```{r tmle-extract}
# Step 4: extract final estimate
tmle_fit <- extract_tmle_estimate(tmle_upd)
print(tmle_fit)
```

```{r audit-stage4}
audit <- record_stage(audit, "Stage 4",
  sprintf("Primary TMLE estimate: %.4f", tmle_fit$estimates$ATE$estimate))
```

### Matched-cohort TMLE

`run_matched_tmle()` runs the full TMLE pipeline on a matched subset
without creating a separate lock --- addressing a common need in
pharmacoepidemiology.

```{r matched-tmle}
# Extract matched row indices from the matching workflow.
# run_match_workflow() preserves original rownames in matched_data.
matched_idx <- as.integer(rownames(match_fit$matched_data))
cat(sprintf("Matched observations: %d (of %d total)\n",
            length(matched_idx), nrow(dat)))

# Run TMLE on just the matched subset
matched_tmle_fit <- run_matched_tmle(lock, ps_fit,
  subset_idx = matched_idx)
print(matched_tmle_fit)
```

### IPCW-weighted TMLE (missing-outcome sensitivity)

When the outcome is missing for a non-trivial fraction of the
analytic cohort and the analyst suspects the missingness mechanism
is at-random rather than completely-at-random, `run_ipcw_tmle()`
fits a SuperLearner response model
$P(R = 1 \mid A, W)$, builds stabilised inverse-probability-of-
censoring weights, and runs TMLE with the censoring weights so the
ATE targets the full-cohort marginal estimand rather than the
complete-case subset.

```{r ipcw-tmle}
ipcw_fit <- run_ipcw_tmle(lock, ps_fit)
print(ipcw_fit)
```

The returned object is a `tmle_fit` and can be combined with the
other estimators in [`summarize_cleanroom_results()`] or
[`forest_plot()`] for a side-by-side comparison.

### TMLE diagnostics

```{r clever-covariate, fig.cap = "Clever covariate H(A,W) by treatment group."}
clever_covariate_plot(ps_fit = ps_fit, lock = lock)
```

```{r ic-hist, fig.cap = "Efficient influence curve distribution."}
ic_histogram(tmle_fit)
```

### Three-way balance comparison

`compute_matched_smds()` and `love_plot_threeway()` show covariate
balance under three adjustment strategies side by side:

```{r threeway-love, fig.cap = "Unweighted, IPTW-weighted, and matched SMDs."}
m_smds <- compute_matched_smds(dat, "treatment", lock$covariates,
                                subset_idx = matched_idx)
love_plot_threeway(diag, matched_smds = m_smds)
```

### Cross-estimator comparison

```{r compare}
all_fits <- list(
  Crude    = list(estimate = crude$estimate, se = crude$se,
                  ci_lower = crude$ci_lower, ci_upper = crude$ci_upper,
                  p_value  = crude$p_value),
  Matching = match_fit,
  IPTW     = iptw_fit,
  `TMLE (primary)` = tmle_fit
)
summary_tbl <- summarize_cleanroom_results(
  list(Matching = match_fit, IPTW = iptw_fit, `TMLE (primary)` = tmle_fit)
)
summary_tbl
```

TMLE is the locked primary analysis; matching and IPTW serve as
contextual comparators.
Agreement across estimators increases confidence that the result is
not an artefact of a single modelling strategy.

---

## Reporting cumulative risks and contrasts

Treatment-specific risks at the prespecified follow-up time, the
risk difference, the risk ratio, event counts, and the estimator
that produced each row are reported in a single structured table
via `clean_risk_report_table()`.

```{r risk-report-vignette, eval = exists("tmle_fit")}
ate <- tmle_fit$estimates$ATE
crude_r1 <- mean(lock$data[[lock$outcome]][lock$data[[lock$treatment]] == 1])
crude_r0 <- mean(lock$data[[lock$outcome]][lock$data[[lock$treatment]] == 0])

risk_rows <- list(
  list(time_point = 24, treatment_strategy = "Treated",
       n_observed = sum(lock$data[[lock$treatment]] == 1),
       risk = crude_r1 + ate$estimate / 2,
       estimator = "Modular TMLE",
       nuisance_specification = paste(lock$sl_library, collapse = "+")),
  list(time_point = 24, treatment_strategy = "Control",
       n_observed = sum(lock$data[[lock$treatment]] == 0),
       risk = crude_r0 - ate$estimate / 2,
       risk_difference = ate$estimate,
       risk_ci_lower = ate$ci_lower,
       risk_ci_upper = ate$ci_upper,
       estimator = "Modular TMLE",
       nuisance_specification = paste(lock$sl_library, collapse = "+")))
print(clean_risk_report_table(risk_rows))
```

### Why hazard ratios are secondary in this workflow

cleanTMLE emphasises cumulative-risk contrasts because they map
directly to treatment-specific risks under specified strategies and
are generally more interpretable for regulatory RWE studies.
Hazard ratios may be useful as secondary summaries when
proportional hazards is plausible and the estimand is explicitly
hazard-based, but they should not replace prespecified
cumulative-risk contrasts when the causal question concerns
absolute or relative risk by a clinically meaningful follow-up
time.

---

## Sensitivity analysis

### Truncation sensitivity

The prespecified sensitivity plan varies the PS truncation threshold
to assess how much the IPTW estimate depends on extreme weights.

```{r sensitivity-trunc}
sens <- sensitivity_truncation(lock,
  thresholds = c(0.01, 0.025, 0.05, 0.10))
print(sens, row.names = FALSE)
```

### E-value

The E-value quantifies the minimum strength of unmeasured confounding
(on the risk-ratio scale) needed to explain away the observed
association.

```{r evalue}
ate <- tmle_fit$estimates$ATE
# Approximate risk ratio from TMLE risk difference
# Using crude arm risks for illustration
r0_hat <- crude$r0
rr_hat <- (r0_hat + ate$estimate) / r0_hat
rr_lo  <- (r0_hat + ate$ci_lower) / r0_hat
ev <- compute_evalue(rr_hat, ci_bound = min(rr_lo, 1/rr_lo))
cat(sprintf("E-value: %.2f (CI bound: %.2f)\n", ev["e_value"], ev["e_value_ci"]))
```

---

## Audit trail and reproducibility

The audit log documents every stage decision, its timestamp, and
the lock hash, providing a traceable record of the analytic path.

```{r audit-trail}
print(audit)
```

```{r stage-manifest}
build_stage_manifest(audit)
```

```{r audit-df}
trail_df <- export_audit_trail(audit)
knitr::kable(trail_df, caption = "Audit trail for the staged analysis.")
```

### Stage path narrative

`summarize_stage_path()` produces a human-readable narrative of the
stages traversed, their decisions, and the overall outcome.
This is useful for inclusion in study reports or supplementary
materials.

```{r stage-path}
summarize_stage_path(audit)
```

### Decision log

The decision log collects the rationale behind key analytic choices
made throughout the staged workflow.
`export_decision_log()` returns these entries as a data frame suitable
for archival or tabular display.

```{r decision-log}
dlog_df <- export_decision_log(audit)
knitr::kable(dlog_df, caption = "Decision log for the staged analysis.")
```

---

## Summary

The worked example illustrates the canonical stages:

1. **Stage 0** --- feasibility, target-trial specification, and
   protocol / SAP drafting (precondition, external to the package).
2. **Stage 1a** --- lock the estimand, covariates, learners,
   candidate grid, and decision thresholds.
3. **Stage 1b** --- cohort adequacy and marginal event support
   (Check Point 1, with design-precision diagnostics).
4. **Stage 2a** --- propensity score, balance, overlap, ESS, and
   design diagnostics (Check Point 2).
5. **Stage 2b** --- baseline plasmode candidate selection using the
   prespecified rule.
6. **Stage 2c** --- plasmode DQ stress testing on the locked
   severity ranges from the fit-for-purpose review.
7. **Stage 3 (optional)** --- negative-control outcome checks.
8. **Pre-outcome decision** --- GO (proceed),
   FLAG (proceed-with-qualification), or STOP (do-not-run), based
   on the recorded checkpoint results.
9. **Stage 4** --- authorized primary outcome analysis on the
   unmasked data.
10. **Post-outcome** --- sensitivity analyses, interpretation, and
    archived analysis record (analysis specification, audit trail,
    and decision log).

The package includes additional functions used along the way:
`sanitize_covariates()` for covariate preparation,
`run_negative_control_tmle()` for residual-bias screening, the
composite `gate_all()` for evaluating checkpoint objects together,
`run_matched_tmle()` and the four-step TMLE pipeline for the final
analysis, `compute_evalue()` and `sensitivity_truncation()` for
sensitivity analyses, and `run_plasmode_dq_stress()` with
`summarize_dq_degradation()` for studies that want to combine
candidate selection with a quantitative pre-outcome stress test.

### Integration with real-world pipelines

The package now supports common production patterns:

- **External PS**: `wrap_ps_fit()` wraps PS from parallel SuperLearner,
  Bayesian models, or external software into a compatible `ps_fit`.
- **Parallel SuperLearner**: `fit_ps_parallel()` accepts a PSOCK cluster.
- **File-based pipelines**: `save_lock()` / `load_lock()` and
  `save_audit()` / `load_audit()` handle RDS serialisation with
  automatic hash re-validation.
- **YAML config**: `create_analysis_lock_from_yaml()` reads analytic
  specifications from a YAML configuration file.
- **Matched-cohort TMLE**: `run_matched_tmle()` runs TMLE on a subset
  without creating a separate lock.

The package does not enforce institutional governance (role separation,
independent review boards, blinded data managers); those are
organisational responsibilities.
What it does provide is software scaffolding that makes the staged
workflow reproducible, auditable, and less susceptible to unintentional
outcome-informed decision making.

```{r session-info}
sessionInfo()
```
