---
title: "A Staged Outcome-Blind TMLE Workflow with cleanTMLE"
subtitle: "Worked Example"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{A Staged Outcome-Blind TMLE Workflow with cleanTMLE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  warning    = FALSE,
  message    = FALSE
)
library(cleanTMLE)
```

## Read this first

This vignette has two halves. The **first half** teaches the core estimator and shows what a plain end-to-end analysis looks like using `cleanTMLE` as a regular RWE estimation package. The **second half** adds the staged-analysis governance layer (analysis lock, audit log, GO / FLAG / STOP gate) for studies that need a reproducible pre-outcome dossier.

If you are new to targeted-learning methods, read the first half straight through. The vocabulary you need is in the **Concept primer** below. The governance vocabulary (clean room, gate, dossier) does not appear until §"The staged workflow" — by then you will have already seen Table 1, the propensity-score overlap, and a risk-curve plot, so the governance ideas have something concrete to attach to.

## A clinical anchor

We use a small simulated cohort (`sim_func1()`) that we will treat as if it came from a pharmaco-epidemiology study. Read it as:

- **Population.** Adult patients enrolled at a tertiary hospital network.
- **Treatment** (`treatment`). A new oral medication started at enrolment versus standard care.
- **Outcome** (`event_24`). A binary event at 24 months (for example, a major cardiovascular event).
- **Baseline covariates** (`age`, `sex`, `biomarker`, `comorbidity`). Standard pre-treatment demographics and risk factors recorded at enrolment.

The numbers are simulated, but the question we are asking is the one a real pharmaco-epi study would ask: **does the new medication change the 24-month event risk, after we adjust for baseline differences between the treated and untreated groups?**

```{r clinical-anchor}
set.seed(2026)
dat <- sim_func1(n = 600, seed = 2026)
head(dat[, c("treatment", "event_24", "age", "sex", "biomarker",
             "comorbidity")])
cat("\nGroup sizes:\n")
table(treatment = dat$treatment, useNA = "ifany")
cat("\nRaw 24-month event rates by treatment arm:\n")
round(tapply(dat$event_24, dat$treatment, mean, na.rm = TRUE), 3)
```

## Concept primer

Eight terms that appear repeatedly in this vignette and in the targeted-learning literature, in plain language:

- **Estimand.** The exact quantity you want to estimate. For us: the population-average 24-month event-risk difference if everyone took the new medication versus if no one did.
- **Propensity score (PS).** The probability of receiving treatment given a person's baseline covariates. Predicted from a model fit on the data.
- **Overlap (positivity).** Whether treated and untreated groups have similar baseline profiles. Poor overlap means adjustment cannot recover the marginal effect for the full cohort.
- **SMD (standardised mean difference).** A balance metric: the difference in covariate means between treated and untreated, standardised by the pooled SD. Smaller is better; < 0.10 is the convention.
- **ESS (effective sample size, Kish).** The "real" sample size after weighting; a few heavy weights can shrink ESS far below `n`.
- **TMLE (targeted maximum-likelihood estimation).** A four-step recipe: (1) predict the outcome from covariates and treatment; (2) predict the propensity score; (3) *target* the outcome predictions using the propensity score so the resulting treatment-effect estimate has the right calibration; (4) read off the contrast and its standard error.
- **Clever covariate.** The specific function of the propensity score that TMLE uses in step 3. You do not have to compute it by hand; `run_clean_tmle()` does it.
- **EIC (efficient influence curve) variance.** The standard-error formula TMLE uses. Interpretable as a regular Wald-type SE under standard regularity conditions.

Two more terms come up in the second half of the vignette: **plasmode simulation** (a simulation that keeps the real covariate and treatment structure but generates a synthetic outcome with a known truth) and **IPCW** (inverse-probability-of-censoring weighting, used to handle missing outcomes).

## The plain workflow: four estimators on one figure

Before introducing any governance machinery, we run the full estimator menu on the clinical anchor and compare them on a single figure. This is the workflow you can lift straight into a methods-development analysis.

```{r plain-workflow}
# 1. Create a "simple" lock that disables the staged-analysis machinery
#    (no outcome guard, no gate, no required audit). The fingerprint
#    and design diagnostics remain available.
plain_lock <- create_simple_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  sl_library = c("SL.glm", "SL.glmnet", "SL.mean"),
  seed       = 2026
)

# 2. Propensity-score model + overlap and balance diagnostics
ps      <- fit_ps_superlearner(plain_lock)
ps_diag <- compute_ps_diagnostics(ps)
print(ps_diag)
```

```{r plain-overlap, fig.cap = "Propensity-score overlap by treatment arm. Both arms should have density across the same range; gaps signal positivity strain."}
plot(ps_diag)
```

The overlap plot is the first thing a reviewer should see. If the two density curves do not cover the same range, adjustment cannot recover a marginal contrast for the full eligible cohort and you will need to consider matching or trimming.

```{r plain-estimators, results = "hide"}
# 3. Attach a primary TMLE specification to the lock so the four-step
#    TMLE knows which library and truncation level to use.
primary_spec <- tmle_candidate(
  candidate_id = "plain_tmle",
  label        = "TMLE for the plain workflow",
  q_library    = c("SL.glm", "SL.glmnet"),
  g_library    = c("SL.glm", "SL.glmnet"),
  truncation   = 0.01
)
plain_lock <- lock_primary_tmle_spec(plain_lock, primary_spec)

# 4. Run four estimators on the same lock and same PS
crude  <- run_crude_workflow(plain_lock)
match  <- run_match_workflow(plain_lock, ps_fit = ps)
iptw   <- run_iptw_workflow(plain_lock, ps_fit = ps)

# 5. TMLE via the four-step modular API
g_fit    <- fit_tmle_treatment_mechanism(plain_lock, ps)
Q_fit    <- fit_tmle_outcome_mechanism(plain_lock, g_fit)
tmle_upd <- run_tmle_targeting_step(g_fit, Q_fit)
tmle     <- extract_tmle_estimate(tmle_upd)
```

```{r plain-summary}
# 6. Combine the four estimates in one table.
#    Each result has $estimate; CI may be on the top level (run_*_workflow)
#    or under $estimates$ATE (modular TMLE extract).
pick_ci <- function(x) {
  if (!is.null(x$estimates) && !is.null(x$estimates$ATE)) {
    list(est = x$estimates$ATE$estimate,
         lo  = x$estimates$ATE$ci_lower,
         hi  = x$estimates$ATE$ci_upper)
  } else {
    se_or_na <- if (is.null(x$se) || length(x$se) == 0) NA_real_ else x$se
    lo_top   <- if (is.null(x$ci_lower)) x$estimate - 1.96 * se_or_na else x$ci_lower
    hi_top   <- if (is.null(x$ci_upper)) x$estimate + 1.96 * se_or_na else x$ci_upper
    list(est = x$estimate, lo = lo_top, hi = hi_top)
  }
}
plain_results <- data.frame(
  estimator = c("Crude (unadjusted)", "PS matching",
                "IPTW (stabilised)", "TMLE (full cohort)"),
  estimate  = c(pick_ci(crude)$est, pick_ci(match)$est,
                pick_ci(iptw)$est,  pick_ci(tmle)$est),
  ci_lower  = c(pick_ci(crude)$lo,  pick_ci(match)$lo,
                pick_ci(iptw)$lo,   pick_ci(tmle)$lo),
  ci_upper  = c(pick_ci(crude)$hi,  pick_ci(match)$hi,
                pick_ci(iptw)$hi,   pick_ci(tmle)$hi),
  stringsAsFactors = FALSE
)
knitr::kable(plain_results, digits = 4,
             caption = "Four estimators on the plain-mode lock.")
```

(`run_ipcw_tmle()` adds an IPCW sensitivity path for studies with missing outcomes; `estimate_gcomprisk()` and `estimate_aipwrisk()` are also available, but they consume a `specify_models()` object rather than the `cleanroom_lock` shown here. See the function-reference vignette for both.)

```{r plain-forest, fig.cap = "Forest plot of the four estimators. The dashed line marks no effect; the adjusted estimators agree closely and shift the crude estimate.", fig.height = 3.5}
library(ggplot2)
ggplot(plain_results,
       aes(x = estimate, y = factor(estimator, levels = rev(plain_results$estimator)))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.2) +
  labs(x = "Risk difference (95% CI)", y = NULL,
       title = "Four estimators on one figure",
       subtitle = "Crude vs PS matching vs IPTW vs TMLE") +
  theme_minimal(base_size = 11)
```

A few things to read from the forest plot:

- The three adjusted estimators (PS matching, IPTW, TMLE) agree closely. When the design diagnostics look reasonable, this is exactly what you should see; substantial disagreement is a signal to slow down.
- The crude (unadjusted) estimate sits noticeably further from the others. The shift between crude and the adjusted estimators is the contribution of the baseline confounders.
- The confidence intervals are not the same width. IPTW's IF-based SE assumes the propensity score is known; in practice it has been estimated, so the IPTW SE is conservative. TMLE's SE accounts for the estimation of the propensity score.
- The matched estimator's CI is wider than the full-cohort TMLE's CI because matching restricts the analysis to the overlap subset and discards unmatched controls.

That is the entire core workflow. Everything in the rest of this vignette is *governance and audit machinery* on top of these four estimators, for studies that need a reproducible pre-outcome dossier, and is not a different estimator.

## When to read the rest

Continue reading if any of the following apply:

- You are running an analysis where the comparative answer will be reviewed by a third party, and you need a reproducible record of every analytic decision made before the comparative answer was seen.
- You want to use the **plasmode candidate-selector** to pick between competing TMLE specifications before unblinding.
- You want to use the **data-quality stress test** to check how the chosen estimator behaves under prespecified threats (measurement error, missingness, unmeasured confounding) before unblinding.
- You are working inside an institutional clean-room arrangement and need the software-side artefacts (lock fingerprint, audit log, decision log, GO / FLAG / STOP record).

If none of these apply, you can stop here and use the plain workflow above. cleanTMLE was designed to be useful at both levels of formality.

---

## Purpose of this vignette

This vignette demonstrates a staged workflow for an observational
comparative-effectiveness analysis on a simulated dataset. Design
and estimator-selection decisions are made before the observed
treatment-outcome association is examined; the goal is reproducible
prespecification and a structured documentation of analytic
decisions that an external reviewer can follow.

In this manuscript and vignette, *outcome-blind* means that
decisions about cohort construction, candidate estimators,
simulation scenarios, decision thresholds, and primary-estimator
selection are made without inspecting the observed
treatment-outcome association. Synthetic outcomes generated for
plasmode simulations may be used; the marginal mean of the
observed outcome on covariates is also used internally to fit the
plasmode generator. The *pre-outcome decision point* is the
juncture at which the prespecified diagnostics are reviewed and the
analysis is classified GO, FLAG (proceed only with documented
review), or STOP before the observed outcome is unmasked for the
comparative analysis.

cleanTMLE is not a substitute for target-trial design, source-data
validation, phenotype validation, independent oversight, or
subject-matter assessment of identification assumptions. The
worked example below shows reproducible prespecification, not a
regulatory submission template.

## Conceptual workflow

The vignette proceeds in the following conceptual steps:

1. **Stage 0 (precondition)**: feasibility assessment, target-trial
   specification, and protocol / SAP drafting. This is largely
   external work conducted before any cleanTMLE code is run.
2. **Stage 1a**: lock the estimand, covariates, learners, candidate
   grid, and decision thresholds.
3. **Stage 1b**: assess cohort adequacy and marginal event support.
4. **Stage 2a**: propensity score, balance, overlap, ESS, and
   design diagnostics.
5. **Stage 2b**: baseline plasmode candidate selection.
6. **Stage 2c**: plasmode data-quality (DQ) stress testing as a
   quantitative pre-outcome supplement to the fit-for-purpose review.
7. **Stage 3 (optional)**: negative-control outcome checks.
8. **Pre-outcome decision**: GO / FLAG / STOP.
9. **Stage 4**: authorized primary outcome analysis.
10. Post-outcome sensitivity analyses, interpretation, and archived
    analysis record.

The Muntner et al. (2024) clean-room governance framework
(role separation, restricted access, protocols, audit) is the
source of the staging idea and is the broader notion of a clean
room. cleanTMLE is the *software layer* of that workflow: it
records the analysis lock, runs the design diagnostics and the
plasmode evaluation, executes the DQ stress test, applies the
pre-outcome decision rule, and writes the audit trail. It does
not by itself implement clean-room personnel governance. The canonical stage labels used
throughout this vignette are:

```
Stage 0    Feasibility, target-trial specification, protocol / SAP drafting
Stage 1a   Lock estimand, covariates, learners, candidate grid, thresholds
Stage 1b   Cohort adequacy and marginal event support
Stage 2a   PS, balance, overlap, ESS, design diagnostics
Stage 2b   Baseline plasmode candidate selection
Stage 2c   DQ stress testing
Stage 3    Optional negative-control outcome checks
Pre-outcome decision: GO / FLAG / STOP
Stage 4    Authorized primary outcome analysis
Post-outcome: sensitivity analyses and interpretation
```

We use *proceed*, *proceed-with-qualification*, and *do-not-run*
in prose; the software shorthand for the same labels is GO, FLAG,
and STOP.

::: {.callout-warning}
This vignette demonstrates the package workflow. It does not
establish exchangeability, measurement validity, phenotype
validity, linkage accuracy, or regulatory acceptability.
:::

::: {.callout-warning}
Low-replicate examples in this vignette are for workflow demonstration only. Use `n_reps >= 200` before interpreting bias, coverage, or RMSE inferentially.
:::

## Pre-outcome study dossier

The reviewer-facing artefact produced by this workflow is a pre-outcome
study dossier. It is assembled across the stages demonstrated below,
and the review team reads it before authorising the primary analysis.
Its components are:

- Protocol and target-trial timing (prespecified estimand record); see *Define the causal question and estimand*.
- Cohort flow and attrition (`attrition_table()`); see *Stage 1b*.
- Event-process classification (structured event-process table and coherence check); see *From target trial specification to event-process classification*.
- Covariate and missingness summary (`make_table1()` and missingness checks); see *Stage 1a* covariate sanitisation.
- PS overlap and balance (`compute_ps_diagnostics()`, `love_plot()`, `love_plot_threeway()`); see *Stage 2a*.
- Treatment- and IPCW-weight diagnostics (`clean_weight_diagnostics()`, `extreme_weights()`); see *Outcome-blind diagnostics before fitting the final estimator*.
- Event-count and precision adequacy (`checkpoint_cohort_adequacy()`); see *Stage 1b*.
- Baseline plasmode candidate selection (`run_plasmode_feasibility()`, `select_tmle_candidate()`); see *Stage 2b*.
- DQ stress-test results (`run_plasmode_dq_stress()`, `summarize_dq_degradation()`); see *Stage 2c*.
- Negative-control eligibility and attrition (`run_residual_confounding_stage()`); see *Stage 3*.
- GO / FLAG / STOP checkpoint dashboard (`gate_all()`; aggregate `checkpoint_dashboard()` planned); see *Pre-outcome decision*.
- Decision log and audit log export (`export_decision_log()`); see *Audit trail and reproducibility*.
- Authorisation record for primary analysis (`authorize_outcome_analysis()`); see *Pre-outcome decision*.

The dossier is the artefact the reviewer reads. Comparative
treatment-outcome estimates are not part of the dossier and are
produced only after authorisation.

### Why TMLE fits this workflow

TMLE is well suited to an outcome-blind staged workflow for a
small number of inter-related reasons:

- The estimand is declared in advance and the estimator is
  defined as the targeted update of an initial outcome model
  toward that estimand, so the analysis target is separable from
  the nuisance fits.
- The nuisance library, the propensity-score truncation, and the
  targeting step can all be prespecified and recorded in the
  analysis lock.
- Targeting decouples the bias of the final estimator from the
  quality of any single nuisance fit, allowing the use of
  adaptive learners without driving the choice from the observed
  outcome association.
- The candidate grid is enumerable: different libraries and
  truncation rules define distinct candidates that can be
  screened on synthetic outcomes before authorisation.
- Inference is based on the efficient influence function, which
  provides a transparent variance estimate compatible with the
  diagnostics computed at earlier stages.

### What each diagnostic answers

| Diagnostic | Question answered | What it does not answer |
|---|---|---|
| PS / overlap / balance / ESS | Empirical design feasibility | Operating characteristics of any specific TMLE candidate |
| Baseline plasmode | Estimator behaviour under a locked synthetic outcome model | Behaviour under the realised outcome process |
| DQ stress testing | Stability under prespecified fit-for-purpose threats | True bias magnitude under unobserved threats |
| Negative-control outcomes | Residual bias through a different channel | Primary-estimator behaviour under stress |
| Post-outcome sensitivity | Robustness to deviations after unblinding | Pre-outcome decisions |

### Stage 0 as precondition

Stage 0 covers the feasibility assessment, target-trial
specification, and protocol / SAP drafting that precede any
analytic code. These activities are typically conducted outside
the software layer, often involving subject-matter experts, data
custodians, and study sponsors. cleanTMLE assumes Stage 0 is
complete: the causal question, eligible population, treatment
strategies, follow-up window, and primary estimand are already
specified in a written protocol before Stage 1a begins. The
remainder of this vignette demonstrates Stages 1a through 4.

---

## Define the causal question and estimand

Before writing any analysis code we state the causal question in
target-trial-emulation style.

| Element | Specification |
|---------|---------------|
| **Population** | Adults with simulated disease (the `sim_func1()` cohort) |
| **Treatment strategies** | Treatment (A = 1) versus control (A = 0), assigned at baseline |
| **Follow-up** | 24 months from cohort entry |
| **Outcome** | Primary event occurring by 24 months (`event_24`) |
| **Causal contrast** | Average treatment effect on the risk-difference scale |

The causal estimand is $\psi = E[Y^{a=1}] - E[Y^{a=0}]$, the
population-level risk difference under two hypothetical interventions.
Because we work with observational data, the quantity we can identify
from the observed distribution is the *statistical estimand*

$$
\psi = E_W\!\bigl[E[Y \mid A=1, W] - E[Y \mid A=0, W]\bigr],
$$

which equals the causal parameter under the standard identification
assumptions (consistency, exchangeability, positivity).
The package estimates this statistical estimand; causal interpretation
relies on the investigator's subject-matter defence of those assumptions.

We choose the risk difference as the primary estimand because it has a
transparent absolute-scale interpretation, admits doubly-robust and
semiparametric-efficient estimation via TMLE, and is directly relevant
to clinical decision making.

---

## From target trial specification to event-process classification

Once the target trial and the statistical estimand are stated,
post-baseline events should not be assigned a generic role of
"censoring." Each event should be mapped to the causal question,
to the corresponding ICH E9(R1) intercurrent-event strategy where
applicable, and to the operational role it plays in estimation.
For example, the primary event is modelled directly, death from
other causes is a competing event, loss to follow-up is a
missing-outcome process handled by IPCW, and treatment switching is
an intercurrent event kept under the assigned treatment strategy.
This classification, the target population and reference group, and
the missing-data plan are prespecified and recorded with the
analysis before the outcome is unmasked.

---

## Stage 1a: Record the analysis specification (lock)

The analysis specification records the treatment, outcome,
covariates, candidate estimators, learner library, simulation
settings, and random seed. Later steps refer back to this object
so that the final analysis can be compared with the prespecified
plan, and so that an external reviewer can reconcile the recorded
inputs with the analyses that were run.

```{r stage1a}
library(cleanTMLE)
set.seed(42)

dat <- sim_func1(n = 400, seed = 42)
```

### Covariate sanitisation

Before locking the specification, we sanitise covariates to ensure they
are safe for SuperLearner: impute missing values, drop zero-variance
columns, and fix formula-unsafe names.

```{r sanitize}
clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  verbose = TRUE)
dat <- clean$data
covs <- clean$covariates
```

```{r lock}
lock <- create_analysis_lock(
  data          = dat,
  treatment     = "treatment",
  outcome       = "event_24",
  covariates    = covs,
  sl_library    = c("SL.glm", "SL.mean"),
  plasmode_reps = 40L,
  seed          = 42L
)
```

::: {.callout-tip}
## Optional: outcome masking

For analyses that require stricter blinding, `mask_outcome()` can
replace the real outcome column with `NA` values inside the lock,
preventing any code from accessing the outcome before Stage 4.
Call `unmask_outcome(lock, original_lock)` when the pre-outcome
gates have been passed and the outcome is needed.
In this vignette we proceed without masking for clarity, but the
functions are available for production workflows.
:::

We prespecify the estimand (a marginal risk difference for the
primary event by 24 months) and a truncation sensitivity plan
alongside the lock, and declare a negative-control outcome. This
metadata does not alter the reproducibility hash (which depends only
on the core analytic specification) but provides the structured
record that an outcome-blind staged review requires.

```{r neg-control-decl}
lock <- define_negative_control(lock, "nc_outcome",
  description = "Binary outcome driven by covariates only; treatment has no causal effect"
)
```

We also prespecify the entire GO / FLAG / STOP decision rule in one
object and attach it to the lock. `attach_decision_thresholds()`
records a separate `thresholds_hash`, so the decision rule — not just
the cohort and estimator definitions — is fingerprinted before any
checkpoint is scored. Each checkpoint below can read its thresholds
from this object via the `dt_*()` helpers.

```{r decision-thresholds}
dt <- decision_thresholds(
  balance_max_smd   = 0.10,
  dq_max_abs_bias   = 0.02,
  nco_rule          = "significance",
  nco_alpha         = 0.05
)
lock <- attach_decision_thresholds(lock, dt)
```

```{r print-lock}
validate_analysis_lock(lock)
print(lock)
```

We initialise the audit log at this point so that every subsequent
stage is recorded.

```{r audit-init}
audit <- create_audit_log(lock)
audit <- record_stage(audit, "Stage 1a",
                      "Analysis lock created and estimand attached")
```

### File-based pipeline support

In production pipelines each stage runs as a separate script.
`save_lock()` / `load_lock()` handle serialisation with automatic
hash re-validation, eliminating the boilerplate of manual RDS
save/load and integrity checks.

```{r save-lock}
# Round-trip demonstration: save, load, verify integrity
tmp_lock_path  <- tempfile("lock_",  fileext = ".rds")
tmp_audit_path <- tempfile("audit_", fileext = ".rds")

save_lock(lock, tmp_lock_path)
save_audit(audit, tmp_audit_path)

# Reload --- load_lock() automatically re-validates the hash
lock_reloaded  <- load_lock(tmp_lock_path)
audit_reloaded <- load_audit(tmp_audit_path)
cat("Lock hash match:", identical(lock$lock_hash, lock_reloaded$lock_hash), "\n")
cat("Audit entries match:", length(audit$entries) == length(audit_reloaded$entries), "\n")

# Clean up temp files
unlink(c(tmp_lock_path, tmp_audit_path))
```

In a real multi-stage pipeline each stage script would save to a
shared results directory (e.g., `save_lock(lock, "results/stage1_lock.rds")`)
and the next stage script would `load_lock("results/stage1_lock.rds")`
at its top.

We also begin populating the decision log, which records the
rationale behind key analytic choices.

```{r dlog-lock}
audit <- record_decision_log_entry(audit,
  stage         = "Stage 1a",
  decision_type = "specification",
  description   = "Analysis lock created with GLM + Mean SL library and 40 plasmode reps",
  rationale     = "Minimal library chosen for vignette reproducibility"
)
```

---

## Stage 1b: Cohort adequacy and marginal event support (Check Point 1)

Before any modelling, we confirm that the cohort has adequate size,
event counts, and positivity to support the planned analysis.
If this checkpoint fails, the analysis stops before any comparative
outcome modelling begins.

```{r cp1}
cp1 <- checkpoint_cohort_adequacy(lock,
  min_n_per_arm  = 50,
  min_events     = 30,
  min_prevalence = 0.01
)
print(cp1)
audit <- record_checkpoint(audit, cp1)
```

The decision is **`r cp1$decision`**.
With `r nrow(dat)` subjects and adequate events in both arms, we proceed.

### Design-stage diagnostics

Before moving to propensity-score modelling we run two additional
design-stage diagnostics.
`estimate_design_precision()` reports the detectable effect size
given the locked sample and event counts, optionally compared against
a target minimum detectable difference (MDD).
`summarize_event_support()` tabulates event counts by treatment arm
and covariate strata, flagging any cells with sparse support.

```{r design-precision}
dp <- estimate_design_precision(lock, target_mdd = 0.05)
print(dp)
```

::: {.callout-note}
The MDD report may show "Feasible: NO" with this small simulated
sample (n = `r nrow(dat)`).
This is expected, because the vignette uses a deliberately small dataset
for fast rendering.
In practice you would either increase sample size or relax the
target MDD.
The MDD check is **informational** and does not trigger STOP; the
pipeline continues.
:::

```{r event-support}
es <- summarize_event_support(lock)
print(es)
```

---

## Stage 2a: PS, balance, overlap, ESS, design diagnostics (Check Point 2)

Stage 2a estimates the propensity score and evaluates overlap, balance,
and effective sample size.
The outcome is never accessed at this stage: only covariates (W) and
treatment (A) are used.

### Propensity score estimation

```{r ps}
ps_fit <- fit_ps_glm(lock, truncate = 0.01)
print(ps_fit)
```

::: {.callout-tip}
## Using external or parallel PS

If you have PS estimated outside cleanTMLE (e.g., via a parallel
SuperLearner cluster), wrap them with `wrap_ps_fit()`:

```r
cl <- parallel::makeCluster(2, type = "PSOCK")
ps_fit <- fit_ps_parallel(lock, cluster = cl, truncate = 0.01)
parallel::stopCluster(cl)

# Or bring-your-own PS:
ps_fit <- wrap_ps_fit(lock, ps_scores = my_ps, method = "external_sl")
```

All downstream functions (`compute_ps_diagnostics`, `checkpoint_balance`,
`run_iptw_workflow`, etc.) work identically with wrapped PS objects.
:::

### Diagnostics

```{r ps-diag}
diag <- compute_ps_diagnostics(ps_fit)
print(diag)
```

```{r ps-overlap, fig.cap = "Propensity score overlap by treatment arm."}
plot(diag)
```

### Structured checkpoint

The `checkpoint_balance()` function converts diagnostics into a
GO / FLAG / STOP decision based on prespecified thresholds for
the maximum weighted SMD and minimum ESS percentage.

```{r cp2}
cp2 <- checkpoint_balance(diag,
  max_smd     = 0.10,
  min_ess_pct = 50,
  lock_hash   = lock$lock_hash
)
print(cp2)
audit <- record_checkpoint(audit, cp2)
```

The decision is **`r cp2$decision`**.
All covariates are well balanced after weighting, and the effective
sample size is adequate.

---

## Stage 2b: Outcome-blind TMLE specification selection

Before accessing the real outcome, we compare *prespecified TMLE
specifications* on plasmode-simulated data.
Each candidate varies the propensity-score truncation threshold and/or
the SuperLearner library used for nuisance estimation.
Plasmode simulation preserves the real covariate distribution and
treatment mechanism but generates synthetic outcomes with a known true
effect, allowing us to evaluate bias, RMSE, and coverage for each
candidate specification.

The purpose is *prespecified specification selection*, not performance
chasing: the selection rule (here, minimum RMSE) is declared in
advance, and the selected TMLE specification is locked into the
analysis lock and carried forward exactly to Stage 4 without
ad hoc modification.

Note that the PS truncation level is not only a numerical
stabilisation parameter: different truncation thresholds imply
different effective target populations. Two candidates with
different truncation rules are therefore paired estimator-and-
estimand objects, and the truncation rule is recorded in the
analysis lock alongside the learner library so the
estimand-vs-estimator distinction remains auditable.

#### Estimator selection versus estimand feasibility

A useful mental check before locking the candidate grid:

- If candidates differ only by nuisance learner, screening step,
  or seed, candidate selection is *estimator* selection. The
  causal estimand is the same across candidates.
- If candidates differ by PS truncation, overlap restriction, or
  eligibility filter, they may target *different* effective
  populations. Selection then has an *estimand* component as well.

In the second case, the prespecified rationale should record the
estimand implication for each candidate before any outcome access.
If the staged workflow shows that only candidates with severe
truncation meet the operating-characteristic thresholds, the
practical conclusion is that the original target population may
not be supported by the data and the estimand itself needs
revision, not just the nuisance fit.

### Define TMLE candidates

```{r tmle-candidates}
# Define candidate TMLE specifications that vary truncation
candidates <- list(
  tmle_candidate("glm_t01", "GLM, trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM, trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05),
  tmle_candidate("glm_mean_t01", "GLM+Mean, trunc=0.01",
                 g_library = c("SL.glm", "SL.mean"), truncation = 0.01)
)
```

### Plasmode feasibility evaluation

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05, 0.10),
  reps            = 40L,
  verbose         = FALSE
)
print(plas)
```

### Select and lock the primary TMLE specification

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- lock_primary_tmle_spec(lock, selected)
print(selected)
```

The selected TMLE specification for Stage 4 is
**`r selected$label`** (truncation = `r selected$truncation`).
This specification is now locked into the analysis lock and will be
used automatically by `fit_tmle_treatment_mechanism()` and
`fit_tmle_outcome_mechanism()`.

## Stage 2c: Plasmode data-quality stress test

The DQ stress test is a **quantitative pre-outcome supplement to
the fit-for-purpose review**, not a formal quantitative bias
analysis (QBA) and not a substitute for source-data validation or
post-outcome QBA. The intended sequence is:

1. Fit-for-purpose review identifies plausible threats.
2. Threats are translated into quantitative severity ranges.
3. Severity ranges are locked before primary outcome access.
4. Synthetic outcomes are generated and perturbed.
5. Candidates are evaluated using the same metrics as baseline
   plasmode selection.
6. The same GO / FLAG / STOP rule is applied.

The DQ stress test documents how candidate estimators behave under
the locked threat range; it does not validate the data or replace
post-outcome QBA.

Four mechanisms are supported:

1. *Covariate missingness*: a fraction of values is set to `NA` and
   imputed with the column median.
2. *Treatment misclassification*: an asymmetric pair of (sensitivity,
   specificity) is applied to A.
3. *Outcome misclassification*: an asymmetric pair of (sensitivity,
   specificity) is applied to Y, optionally per treatment arm.
4. *Unmeasured confounding*: a latent binary U is sampled and shifts
   both the treatment mechanism (via a real PS model fitted on the
   lock data) and the outcome probability by user-specified odds
   ratios.

```{r dq-stress, cache = TRUE}
dq_results <- run_plasmode_dq_stress(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05),
  # 60 (rather than 40) replicates stabilise the MCAR/measurement-error cells,
  # whose worst-case |bias| is otherwise driven up by Monte Carlo noise close to
  # the 0.02 threshold; at 60 reps the margin below the STOP threshold is
  # comfortable so the illustrative gate reliably returns FLAG, not STOP.
  reps            = 60L,
  data_quality_scenarios = list(
    # Covariate missingness under three mechanisms, each median-imputed.
    # MCAR is unbiased by construction; MAR (treatment-dependent) and MNAR
    # (value-dependent) make the imputation biased and are stronger tests.
    covariate_missingness = list(            # MCAR
      fractions = c(0.05, 0.10, 0.20)
    ),
    covariate_missingness_mar = list(        # MAR (treatment-dependent)
      fractions = c(0.10, 0.20), treatment_OR = 3
    ),
    covariate_missingness_mnar = list(       # MNAR (value-dependent)
      fractions = c(0.10, 0.20), strength = 1.5
    ),
    treatment_misclass = list(
      sensitivity = c(0.99, 0.95, 0.90),
      specificity = c(0.99, 0.95, 0.90)
    ),
    outcome_misclass = list(
      sensitivity = c(0.95, 0.90),
      specificity = c(0.99, 0.95)
    ),
    unmeasured_confounding = list(
      U_prevalence   = 0.20,
      U_treatment_OR = c(1.5, 2.0),
      U_outcome_OR   = c(1.5, 2.0)
    )
  ),
  verbose = FALSE
)

deg <- summarize_dq_degradation(dq_results)
head(deg)
```

The *degradation gradient* is the curve of estimator performance
against the severity of a prespecified data-quality threat. It is
analogous to a dose-response curve for estimator fragility. A
shallow gradient suggests a candidate is stable across plausible
severities. A cliff suggests that small increases in degradation
move performance outside the locked criteria.

`summarize_dq_degradation()` returns a long table comparing each
degraded run with the baseline (no degradation): bias change, RMSE
ratio, coverage drop. A robust candidate produces a flat curve; a
fragile candidate shows a cliff. The table can be filtered to the
candidate locked in by `select_tmle_candidate()` and reviewed
alongside the SPIFD2 fit-for-purpose narrative before the gate is
scored.

The DQ stress test is the package's distinguishing contribution
relative to prior outcome-blind plasmode work: standard plasmode
selects the candidate with the best baseline RMSE, while
`run_plasmode_dq_stress()` allows the analyst to select the
candidate that stays inside the gate thresholds across the
plausible range of every realistic data threat.

### Turning the DQ stress test into a gate (Check Point 2c)

`gate_dq()` converts the degradation table into a
`cleantmle_checkpoint` so the DQ result enters the GO / FLAG / STOP
decision on the same footing as the cohort, balance, and
residual-bias checkpoints. It returns STOP if any degraded scenario
for the locked candidate breaches the prespecified bias, coverage,
or RMSE-ratio envelope. Drawing the thresholds from the lock's
`decision_thresholds()` object keeps the whole decision rule in one
fingerprinted place.

```{r gate-dq}
cp_dq <- gate_dq(dq_results,
                 candidate  = selected$candidate_id,
                 thresholds = lock$decision_thresholds)  # NULL is fine too
audit <- record_checkpoint(audit, cp_dq)
cat(sprintf("Check Point 2c (DQ stress): %s\n", cp_dq$decision))
```

```{r audit-stage2b}
audit <- record_stage(audit, "Stage 2b",
  paste("Plasmode evaluation complete; selected TMLE spec:",
        selected$candidate_id))
```

```{r dlog-tmle-select}
audit <- record_decision_log_entry(audit,
  stage         = "Stage 2b",
  decision_type = "estimator_selection",
  description   = paste("Selected TMLE candidate:", selected$candidate_id),
  rationale     = "Minimum RMSE rule applied to plasmode simulation results"
)
```

---

## Stage 3 (optional): Negative-control outcome checks

A negative-control outcome (NCO) screen is an optional pre-outcome
checkpoint that sits between the Stage 2c DQ stress test and the
pre-outcome decision. The NCO screen assesses residual
confounding using a negative-control outcome, a variable that is
known (or strongly believed) to be unaffected by treatment.
If the same analytic pipeline detects an association with the negative
control, this suggests that confounders remain unbalanced after
adjustment.

The negative control outcome `nc_outcome` was generated from
covariates only, with no treatment effect, so any estimated association
is attributable to residual confounding.

The `run_residual_confounding_stage()` wrapper runs the negative control
analysis and generates the checkpoint in a single call, reducing
boilerplate and ensuring the two steps stay paired.

```{r stage3}
stage3 <- run_residual_confounding_stage(lock, ps_fit)
print(stage3)
cp3 <- stage3$checkpoint
audit <- record_checkpoint(audit, cp3)
```

### TMLE-based negative controls (doubly robust)

The default `run_negative_control()` uses IPTW.
`run_negative_control_tmle()` uses the four-step TMLE pipeline,
which is doubly robust and more appropriate for a package built
around TMLE.

```{r nc-tmle}
nc_tmle <- run_negative_control_tmle(lock, "nc_outcome", ps_fit)
cat(sprintf("TMLE NC estimate: %.5f  (p = %.4f, method: %s)\n",
            nc_tmle$estimate, nc_tmle$p_value, nc_tmle$method))
```

The decision is **`r cp3$decision`**.
No significant association with the negative control was detected, so
we proceed to the pre-outcome authorization gate.

---

## Pre-outcome decision

### Interpreting GO / FLAG / STOP

The three labels record the result of a prespecified checkpoint,
not a judgement about the underlying assumptions:

- **GO**: every prespecified threshold for this checkpoint was
  met. The workflow may proceed to the next stage.
- **FLAG**: at least one threshold was not met but the result
  does not block progression. The workflow may proceed under
  documented review.
- **STOP**: at least one threshold was not met and the checkpoint
  blocks progression under the locked rule.

An override does not erase the checkpoint result. It records a
documented decision to proceed despite the checkpoint result.

The pre-outcome decision summarises whether the prespecified
diagnostics support proceeding to outcome analysis, require
additional review, or indicate that the study should stop or be
redesigned. `authorize_outcome_analysis()` inspects the audit
trail and verifies that every required checkpoint has been
recorded with a *proceed* (GO) or, when `allow_flag = TRUE`,
*pause for review* (FLAG) classification. If any checkpoint is
missing or returned *stop* (STOP), the prespecified workflow is
not complete and outcome-analysis functions will not run.

`gate_all()` evaluates all checkpoint objects at once and returns
the composite classification. Pass the DQ checkpoint alongside the
others so a DQ STOP blocks the gate:

```{r gate-all}
composite <- gate_all(cp1, cp2, cp_dq, cp3, allow_flag = TRUE)
print(composite)
```

`authorize_outcome_analysis()` then inspects the audit trail. With
`block_on_any_stop = TRUE` (the default), a STOP recorded by *any*
checkpoint — including the optional DQ-stress gate — blocks
authorisation, even though `"Check Point 2c"` is not in the default
`required_stages`. Required stages are matched on the stage key (the
text before the first colon), so `"Check Point 2"` matches the
balance checkpoint but never the DQ checkpoint.

```{r gate}
gate <- authorize_outcome_analysis(audit)
print(gate)
audit <- record_checkpoint(audit, gate)

# Record the passing gate on the lock. This stamps `.outcome_authorized`, so the
# Stage 4 estimators below run; without it they refuse the outcome. (For a
# masked lock, `unmask_outcome(lock, original_lock, audit = audit)` does the same
# check as it reveals the outcome.)
lock <- assert_outcome_authorized(audit, lock = lock)
```

```{r dlog-gate}
audit <- record_decision_log_entry(audit,
  stage         = "Pre-outcome decision",
  decision_type = "authorization",
  description   = paste("Outcome analysis classification:", gate$decision),
  rationale     = "All preceding checkpoints passed."
)
```

The pre-outcome classification is **`r gate$decision`**. Every required
stage (cohort adequacy, treatment comparability, and the negative
control) returned GO and no checkpoint returned STOP, so
`authorize_outcome_analysis()` authorises the outcome analysis. The
optional data-quality stress checkpoint returned FLAG — its worst RMSE
ratio under the covariate-missingness and outcome-misclassification
threats sits just past the FLAG envelope — which is why the `gate_all()`
composite above reads FLAG. A FLAG on a non-required checkpoint does not
block authorisation, so outcome-analysis functions may now run on the
unmasked data.

### What if a checkpoint fails?

In the example above the required checkpoints returned GO and the
optional DQ-stress checkpoint returned FLAG, which still authorises
under review.
In practice, poor overlap or residual confounding can trigger STOP
decisions that block progression.
The block below constructs a synthetic failure to show how the pipeline
responds:

```{r failure-demo}
# Simulate a STOP scenario: a checkpoint with very high SMD
fake_diag <- diag
fake_diag$smds$smd_weighted <- c(0.25, 0.30, 0.15, 0.08)  # badly imbalanced
fake_cp2 <- checkpoint_balance(fake_diag, max_smd = 0.10,
                                lock_hash = lock$lock_hash)
cat(sprintf("Fake CP2 decision: %s\n", fake_cp2$decision))
cat(sprintf("Rationale: %s\n", fake_cp2$rationale))

# gate_all() blocks when a STOP is present:
fake_composite <- gate_all(cp1, fake_cp2, cp3, allow_flag = FALSE)
cat(sprintf("\nComposite gate: %s\n", fake_composite$decision))
cat(sprintf("Rationale: %s\n", fake_composite$rationale))
```

When the gate returns STOP, no outcome code should execute.
In `run_clean_tmle()` this is checked automatically; in the modular
workflow the analyst inspects the gate decision and branches accordingly.

The most consequential STOP in this workflow usually comes from the
DQ stress test rather than from balance. The block below shows a
DQ-driven STOP: a candidate that passes the baseline plasmode but
whose bias under a prespecified threat exceeds the locked envelope.
Because `authorize_outcome_analysis(block_on_any_stop = TRUE)` honours
any recorded STOP, this blocks the gate even though the DQ checkpoint
is not a *required* stage.

```{r dq-stop-demo}
# A DQ checkpoint that breaches the locked bias envelope returns STOP.
fake_cp_dq <- gate_dq(dq_results,
                      candidate    = selected$candidate_id,
                      max_abs_bias = 0.0001)  # deliberately strict
cat(sprintf("Strict DQ gate: %s -- %s\n",
            fake_cp_dq$decision, fake_cp_dq$rationale))

audit_demo <- record_checkpoint(audit, fake_cp_dq)
gate_demo  <- authorize_outcome_analysis(audit_demo)
cat(sprintf("Authorization with DQ STOP recorded: %s\n",
            gate_demo$decision))
```

---

## Outcome-blind diagnostics before fitting the final estimator

Before the outcome is unmasked, we report a compact set of
diagnostics that support assessment of identification and
estimator stability: treatment overlap, censoring weights (when
present), missingness weights (when applicable), weighted SMDs,
and effective sample size. These diagnostics should be conducted
and the prespecified failure thresholds applied before looking at
outcomes when the workflow allows.

```{r outcome-blind-diagnostics, eval = exists("ps_fit")}
A <- lock$data[[lock$treatment]]
ps <- ps_fit$ps
w  <- ifelse(A == 1, 1 / ps, 1 / (1 - ps))

W_for_smd <- lock$data[, lock$covariates, drop = FALSE]
weight_diag <- clean_weight_diagnostics(weights = w, treatment = A,
                                         covariates = W_for_smd,
                                         max_weight_threshold = 10,
                                         ess_floor = 0.4 * length(w))
print(weight_diag)
```

The maximum-weight threshold (10) and the ESS floor (40 % of N)
are conservative defaults; the prespecified analysis plan should
record values that are appropriate for the study's anticipated
overlap, weight distribution, and sample size. When the study
uses censoring or missingness weights, `clean_weight_diagnostics()`
can be called separately on each weight vector.

---

## Stage 4: Final comparative analysis

Only after all preceding checkpoints pass do we access the real
outcome and run the locked analysis.

### Unadjusted benchmark

```{r crude}
crude <- run_crude_workflow(lock)
cat(sprintf("Crude RD: %.4f (95%% CI: %.4f, %.4f)\n",
            crude$estimate, crude$ci_lower, crude$ci_upper))
```

### PS matching

```{r matching}
match_fit <- run_match_workflow(lock, ps_fit)
print(match_fit)
```

### IPTW

```{r iptw}
iptw_fit <- run_iptw_workflow(lock, ps_fit)
print(iptw_fit)
```

### TMLE in four steps (primary estimator)

TMLE is the prespecified primary estimator, using the specification
selected at Stage 2b and locked into the analysis lock.  The locked
truncation threshold and learner libraries are applied automatically
by the four TMLE functions; no manual specification is needed at
this stage.

The package separates the TMLE procedure into four explicit steps
aligned with the outcome-blind stages:

1. **Treatment mechanism** ($g$-model): estimated at Stage 2 without
   outcome access.  The locked truncation is applied automatically.
2. **Outcome mechanism** ($Q$-model): requires the real outcome and
   therefore belongs to Stage 4.  The locked Q-library is used.
3. **Targeting step**: fluctuates the initial outcome estimates using
   the clever covariate $H(A,W) = A/g(W) - (1-A)/(1-g(W))$.
4. **Estimate extraction**: computes the plug-in ATE, EIC-based SE,
   and confidence interval.

```{r tmle-g}
# Step 1: treatment mechanism (re-uses Stage 2 PS)
g_fit <- fit_tmle_treatment_mechanism(lock, ps_fit)
```

```{r tmle-Q}
# Step 2: outcome mechanism (FIRST access to the real outcome)
Q_fit <- fit_tmle_outcome_mechanism(lock, g_fit)
```

```{r tmle-target}
# Step 3: targeting step
tmle_upd <- run_tmle_targeting_step(g_fit, Q_fit)
```

```{r tmle-extract}
# Step 4: extract final estimate
tmle_fit <- extract_tmle_estimate(tmle_upd)
print(tmle_fit)
```

```{r audit-stage4}
audit <- record_stage(audit, "Stage 4",
  sprintf("Primary TMLE estimate: %.4f", tmle_fit$estimates$ATE$estimate))
```

### Matched-cohort TMLE

`run_matched_tmle()` runs the full TMLE pipeline on a matched subset
without creating a separate lock, addressing a common need in
pharmacoepidemiology.

```{r matched-tmle}
# Extract matched row indices from the matching workflow.
# run_match_workflow() preserves original rownames in matched_data.
matched_idx <- as.integer(rownames(match_fit$matched_data))
cat(sprintf("Matched observations: %d (of %d total)\n",
            length(matched_idx), nrow(dat)))

# Run TMLE on just the matched subset
matched_tmle_fit <- run_matched_tmle(lock, ps_fit,
  subset_idx = matched_idx)
print(matched_tmle_fit)
```

### IPCW-weighted TMLE (missing-outcome sensitivity)

When the outcome is missing for a non-trivial fraction of the
analytic cohort and the analyst suspects the missingness mechanism
is at-random rather than completely-at-random, `run_ipcw_tmle()`
fits a SuperLearner response model
$P(R = 1 \mid A, W)$, builds stabilised inverse-probability-of-
censoring weights, and runs TMLE with the censoring weights so the
ATE targets the full-cohort marginal estimand rather than the
complete-case subset.

```{r ipcw-tmle}
ipcw_fit <- run_ipcw_tmle(lock, ps_fit)
print(ipcw_fit)
```

The returned object is a `tmle_fit` and can be combined with the
other estimators in [`summarize_cleanroom_results()`] or
[`forest_plot()`] for a side-by-side comparison.

### TMLE diagnostics

```{r clever-covariate, fig.cap = "Clever covariate H(A,W) by treatment group."}
clever_covariate_plot(ps_fit = ps_fit, lock = lock)
```

```{r ic-hist, fig.cap = "Efficient influence curve distribution."}
ic_histogram(tmle_fit)
```

### Three-way balance comparison

`compute_matched_smds()` and `love_plot_threeway()` show covariate
balance under three adjustment strategies side by side:

```{r threeway-love, fig.cap = "Unweighted, IPTW-weighted, and matched SMDs."}
m_smds <- compute_matched_smds(dat, "treatment", lock$covariates,
                                subset_idx = matched_idx)
love_plot_threeway(diag, matched_smds = m_smds)
```

### Cross-estimator comparison

```{r compare}
all_fits <- list(
  Crude    = list(estimate = crude$estimate, se = crude$se,
                  ci_lower = crude$ci_lower, ci_upper = crude$ci_upper,
                  p_value  = crude$p_value),
  Matching = match_fit,
  IPTW     = iptw_fit,
  `TMLE (primary)` = tmle_fit
)
summary_tbl <- summarize_cleanroom_results(
  list(Matching = match_fit, IPTW = iptw_fit, `TMLE (primary)` = tmle_fit)
)
summary_tbl
```

TMLE is the locked primary analysis; matching and IPTW serve as
contextual comparators.
Agreement across estimators increases confidence that the result is
not an artefact of a single modelling strategy.

---

## Reporting cumulative risks and contrasts

Treatment-specific risks at the prespecified follow-up time, the
risk difference, and its confidence interval are reported in a
single compact table. The arm risks are the model-based (TMLE
plug-in) risks taken directly from the fit; their difference equals
the reported risk difference.

```{r risk-report-vignette, eval = exists("tmle_fit")}
ate <- tmle_fit$estimates$ATE
# Model-based (TMLE plug-in) treatment-specific risks, taken directly from the
# fit rather than reconstructed from crude arm means; their difference equals
# the reported ATE.
r1 <- tmle_fit$risks$treated
r0 <- tmle_fit$risks$control

risk_report <- data.frame(
  treatment_strategy = c("Treated", "Control"),
  n_observed         = c(sum(lock$data[[lock$treatment]] == 1),
                         sum(lock$data[[lock$treatment]] == 0)),
  risk               = c(r1, r0),
  risk_difference    = c(NA, ate$estimate),
  ci_lower           = c(NA, ate$ci_lower),
  ci_upper           = c(NA, ate$ci_upper))
knitr::kable(risk_report, digits = 4,
             caption = "Treatment-specific risks at 24 months and the risk-difference contrast (modular TMLE).")
```

### Why hazard ratios are secondary in this workflow

cleanTMLE emphasises cumulative-risk contrasts because they map
directly to treatment-specific risks under specified strategies and
are generally more interpretable for regulatory RWE studies.
Hazard ratios may be useful as secondary summaries when
proportional hazards is plausible and the estimand is explicitly
hazard-based, but they should not replace prespecified
cumulative-risk contrasts when the causal question concerns
absolute or relative risk by a clinically meaningful follow-up
time.

---

## Sensitivity analysis

### Truncation sensitivity

The prespecified sensitivity plan varies the PS truncation threshold
to assess how much the IPTW estimate depends on extreme weights.

```{r sensitivity-trunc}
sens <- sensitivity_truncation(lock,
  thresholds = c(0.01, 0.025, 0.05, 0.10))
print(sens, row.names = FALSE)
```

### E-value

The E-value quantifies the minimum strength of unmeasured confounding
(on the risk-ratio scale) needed to explain away the observed
association.

```{r evalue}
ate <- tmle_fit$estimates$ATE
# Confounding-adjusted risk ratio from the model-based (TMLE plug-in) arm
# risks the estimator targeted, not the crude control risk. The CI bound
# maps the risk-difference lower limit onto the risk-ratio scale at the
# adjusted control risk.
r1_hat <- tmle_fit$risks$treated
r0_hat <- tmle_fit$risks$control
rr_hat <- r1_hat / r0_hat
rr_lo  <- (r0_hat + ate$ci_lower) / r0_hat
ev <- compute_evalue(rr_hat, ci_bound = min(rr_lo, 1/rr_lo))
cat(sprintf("E-value: %.2f (CI bound: %.2f)\n", ev["e_value"], ev["e_value_ci"]))
```

---

## Audit trail and reproducibility

The audit log documents every stage decision, its timestamp, and
the lock hash, providing a traceable record of the analytic path.

```{r audit-trail}
print(audit)
```

```{r audit-df}
trail_df <- export_audit_trail(audit)
knitr::kable(trail_df, caption = "Audit trail for the staged analysis.")
```

### Decision log

The decision log collects the rationale behind key analytic choices
made throughout the staged workflow.
`export_decision_log()` returns these entries as a data frame suitable
for archival or tabular display.

```{r decision-log}
dlog_df <- export_decision_log(audit)
knitr::kable(dlog_df, caption = "Decision log for the staged analysis.")
```

A structured decision-log entry recording the Stage 2c outcome and an
on-disk export looks like this:

```{r decision-log-example, eval = FALSE}
audit <- init_decision_log(lock)
audit <- log_decision_entry(audit,
  stage     = "Stage 2c",
  diagnostic = "DQ stress test minimax",
  criterion  = "All candidates within bias/coverage thresholds across severities",
  decision   = "GO",
  rationale  = "Minimax candidate glm_t01 selected; full-cohort plasmode confirms.",
  reviewer   = "lead_analyst",
  y_visible  = FALSE,
  action     = "Lock glm_t01 as primary; record DQ summary in dossier."
)
export_decision_log(audit, "decision_log.csv")
```

---

## Summary

The worked example illustrates the canonical stages:

1. **Stage 0**: feasibility, target-trial specification, and
   protocol / SAP drafting (precondition, external to the package).
2. **Stage 1a**: lock the estimand, covariates, learners,
   candidate grid, and decision thresholds.
3. **Stage 1b**: cohort adequacy and marginal event support
   (Check Point 1, with design-precision diagnostics).
4. **Stage 2a**: propensity score, balance, overlap, ESS, and
   design diagnostics (Check Point 2).
5. **Stage 2b**: baseline plasmode candidate selection using the
   prespecified rule.
6. **Stage 2c**: plasmode DQ stress testing on the locked
   severity ranges from the fit-for-purpose review.
7. **Stage 3 (optional)**: negative-control outcome checks.
8. **Pre-outcome decision**: GO (proceed),
   FLAG (proceed-with-qualification), or STOP (do-not-run), based
   on the recorded checkpoint results.
9. **Stage 4**: authorized primary outcome analysis on the
   unmasked data.
10. **Post-outcome**: sensitivity analyses, interpretation, and
    archived analysis record (analysis specification, audit trail,
    and decision log).

The package includes additional functions used along the way:
`sanitize_covariates()` for covariate preparation,
`run_negative_control_tmle()` for residual-bias screening, the
composite `gate_all()` for evaluating checkpoint objects together,
`run_matched_tmle()` and the four-step TMLE pipeline for the final
analysis, `compute_evalue()` and `sensitivity_truncation()` for
sensitivity analyses, and `run_plasmode_dq_stress()` with
`summarize_dq_degradation()` for studies that want to combine
candidate selection with a quantitative pre-outcome stress test.

### Integration with real-world pipelines

The package now supports common production patterns:

- **External PS**: `wrap_ps_fit()` wraps PS from parallel SuperLearner,
  Bayesian models, or external software into a compatible `ps_fit`.
- **Parallel SuperLearner**: `fit_ps_parallel()` accepts a PSOCK cluster.
- **File-based pipelines**: `save_lock()` / `load_lock()` and
  `save_audit()` / `load_audit()` handle RDS serialisation with
  automatic hash re-validation.
- **YAML config**: `create_analysis_lock_from_yaml()` reads analytic
  specifications from a YAML configuration file.
- **Matched-cohort TMLE**: `run_matched_tmle()` runs TMLE on a subset
  without creating a separate lock.

The package does not enforce institutional governance (role separation,
independent review boards, blinded data managers); those are
organisational responsibilities.
What it does provide is software scaffolding that makes the staged
workflow reproducible and auditable.

```{r session-info}
sessionInfo()
```
