---
title: "cleanTMLE Function Reference"
subtitle: "API Patterns and Quick Examples"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{cleanTMLE Function Reference}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

This article is a compact index of every exported function in
**cleanTMLE**, organised by outcome-blind stage.
For a narrative worked example, see the companion vignette
*A Staged Outcome-Blind TMLE Workflow with cleanTMLE*.

```{r load}
library(cleanTMLE)
dat <- sim_func1(n = 400, seed = 42)
```

---

## Outcome-access taxonomy

The function index below labels each exported function by what it
uses internally from the outcome side of the data:

- **No outcome access**: uses W, A, metadata, lock objects,
  synthetic outcomes, or diagnostics not involving Y.
- **Marginal Y only**: uses event counts or marginal outcome
  support (no estimation of the A--Y association).
- **Synthetic Y only**: uses plasmode-generated outcomes; no
  observed Y.
- **Negative-control Y only**: uses prespecified NC outcomes,
  not the primary outcome.
- **Primary Y**:estimates the observed
  treatment-primary-outcome association; requires authorization.

These labels describe what exported package functions use
internally, not what a user with raw data access could inspect
outside the package.

---

## Function Index

The table below lists every exported function, its outcome-blind
stage, and what it uses from the outcome side of the data.

| Stage | Function | Outcome? | Purpose |
|-------|----------|:--------:|---------|
| Util | `expit()`, `logit()` | n/a | Inverse-logit / logit transforms |
| Data | `sim_func1()` | n/a | Simulate causal-inference data |
| 1a | `create_analysis_lock()` | No | Lock analytic specification |
| 1a | `validate_analysis_lock()` | No | Verify lock integrity |
| 1a | `attach_estimand()` | No | Attach estimand metadata |
| 1a | `declare_sensitivity_plan()` | No | Declare a sensitivity plan |
| 1a | `define_negative_control()` | No | Register a negative control variable |
| 1b | `checkpoint_cohort_adequacy()` | Marginal Y only | Check Point 1: sample size / events |
| 1b | `estimate_design_precision()` | Marginal Y only | Design-stage precision/MDD |
| 1b | `summarize_event_support()` | Marginal Y only | Event counts by arm |
| 2 | `fit_ps_glm()` | No | Logistic regression PS |
| 2 | `fit_ps_superlearner()` | No | SuperLearner PS |
| 2 | `compute_ps_diagnostics()` | No | Overlap, ESS, SMDs |
| 2 | `checkpoint_balance()` | No | Check Point 2: balance decision |
| 2b | `tmle_candidate()` | No | Define a TMLE candidate specification |
| 2b | `expand_tmle_candidate_grid()` | No | Generate candidate grid |
| 2b | `validate_tmle_candidates()` | No | Validate candidate list |
| 2b | `run_plasmode_feasibility()` | Synthetic Y only | Plasmode simulation evaluation |
| 2b | `summarize_plasmode_results()` | No | Print plasmode metrics |
| 2b | `select_tmle_candidate()` | No | Select best TMLE specification |
| 2b | `lock_primary_tmle_spec()` | No | Lock selected spec in the lock |
| 2b | `get_primary_tmle_spec()` | No | Retrieve locked spec |
| 2b | `gate_check()` | No | GO / FLAG / STOP for plasmode |
| 3 | `run_negative_control()` | Negative-control Y only | Negative control analysis |
| 3 | `run_residual_confounding_stage()` | Negative-control Y only | Stage 3 wrapper |
| 3 | `checkpoint_residual_bias()` | No | Check Point 3: residual bias |
| Gate | `authorize_outcome_analysis()` | No | Pre-outcome GO/STOP gate |
| Gate | `assert_outcome_authorized()` | No | Gate assertion (errors if not GO) |
| 4 | `run_crude_workflow()` | Primary Y | Unadjusted risk difference |
| 4 | `run_match_workflow()` | Primary Y | PS matching |
| 4 | `run_iptw_workflow()` | Primary Y | Stabilised IPTW |
| 4 | `fit_tmle_treatment_mechanism()` | No | TMLE g-step (re-uses PS) |
| 4 | `fit_tmle_outcome_mechanism()` | Primary Y | TMLE Q-step |
| 4 | `run_tmle_targeting_step()` | Primary Y | TMLE targeting / fluctuation |
| 4 | `extract_tmle_estimate()` | Primary Y | Final TMLE estimate |
| 4 | `fit_tmle_candidate_set()` | Primary Y | Fit multiple TMLE specs |
| 4 | `fit_final_workflows()` | Primary Y | Run all Stage 4 workflows |
| 4 | `sensitivity_truncation()` | Primary Y | Truncation sensitivity |
| Any |`compute_evalue()` | No | E-value for unmeasured confounding |
| Any |`summarize_cleanroom_results()` | No | Cross-workflow comparison table |
| Any |`record_decision_log_entry()` | No | Add structured decision log entry |
| Any |`export_decision_log()` | No | Export decision log as data frame |
| Any |`summarize_stage_path()` | No | Compact stage narrative |
| Any |`create_audit_log()` | No | Initialise audit trail |
| Any |`record_stage()` | No | Append stage entry |
| Any |`record_checkpoint()` | No | Append checkpoint entry |
| Any |`export_audit_trail()` | No | Export trail as data frame |
| Any |`build_stage_manifest()` | No | Print stage path summary |
| Util | `mask_outcome()` | No | Mask outcome column with NA |
| Util | `unmask_outcome()` | No | Restore masked outcome |
| Wrapper | `run_clean_tmle()` | Primary Y | Full staged clean-room pipeline |
| Wrapper | `init_decision_log()` | No | Create empty decision log |
| Wrapper | `log_decision_entry()` | No | Append row to decision log |
| Wrapper | `save_decision_log()` | No | Write decision log to CSV |
| Plot | `love_plot()` | No | SMD dot-plot (weighted vs unweighted) |
| Plot | `ic_histogram()` | No | Influence curve histogram |
| Integ | `wrap_ps_fit()` | No | Wrap external PS scores into ps_fit |
| Integ | `fit_ps_parallel()` | No | SuperLearner PS with parallel cluster |
| Integ | `save_lock()` / `load_lock()` | No | Serialise / deserialise lock to RDS |
| Integ | `save_audit()` / `load_audit()` | No | Serialise / deserialise audit log |
| 3 | `run_negative_control_tmle()` | Negative-control Y only | TMLE-based negative control (doubly robust) |
| 4 | `run_matched_tmle()` | Primary Y | Modular TMLE on a matched subset |
| Integ | `sanitize_covariates()` | No | Clean covariates for SuperLearner |
| Integ | `create_analysis_lock_from_yaml()` | No | Lock from YAML config file |
| Plot | `clever_covariate_plot()` | No | H(A,W) distribution by group |
| Plot | `love_plot_threeway()` | No | Unweighted / IPTW / matched SMDs |
| Integ | `compute_matched_smds()` | No | SMDs on a matched subset |
| 1 | `attrition_table()` | No outcome access | Cohort flow / attrition documentation |
| 1 | `get_final_cohort()` | No | Extract filtered data from attrition table |
| 1 | `clean_event_process_table()` | No outcome access | Event-process classification table |
| 1 | `clean_check_event_processes()` | No outcome access | Verify event-process specification |
| 2 | `clean_weight_diagnostics()` | No outcome access | Weight ESS, percentiles, max-weight thresholds |
| 3  | `refine_ps_after_nco()` | Negative-control Y only | Iterative PS refinement after NCO signal |
| 2c | `run_plasmode_dq_stress()` | Synthetic Y only | DQ stress test (missingness, misclass, U) |
| 2c | `summarize_dq_degradation()` | No | DQ degradation comparison table |
| Gate | `gate_all()` | No | Composite GO/FLAG/STOP across checkpoints |

**Marginal Y only**: accesses event counts or marginal outcome support, not the A--Y association.
**Synthetic Y only**: generates simulated outcomes; does not use the real outcome association.
**Negative-control Y only**: accesses prespecified negative-control outcomes, not the primary outcome.
**Primary Y**: estimates the observed treatment--primary-outcome association and requires authorization.

---

## Utilities

```{r utils}
expit(0)            # 0.5
logit(0.5)          # 0
expit(logit(0.73))  # 0.73
```

---

## Stage 1a: Lock, Estimand, Plans

```{r lock}
lock <- create_analysis_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  sl_library = c("SL.glm", "SL.mean"),
  plasmode_reps = 40L,
  seed       = 42L
)

lock <- attach_estimand(lock,
  description          = "Treatment effect on 24-month event risk",
  contrast             = "risk_difference",
  statistical_estimand = "E_W[E(Y|A=1,W) - E(Y|A=0,W)]"
)

lock <- declare_sensitivity_plan(lock, "truncation",
  description = "Vary PS truncation",
  settings    = list(thresholds = c(0.01, 0.05, 0.10)))

lock <- define_negative_control(lock, "nc_outcome")

validate_analysis_lock(lock)
print(lock)
```

### Outcome Masking

```{r masking}
masked <- mask_outcome(lock)
all(is.na(masked$data[["event_24"]]))
unmasked <- unmask_outcome(masked, lock)
```

---

## Stage 1b: Cohort Adequacy (Check Point 1)

```{r cp1}
cp1 <- checkpoint_cohort_adequacy(lock)
print(cp1)
```

---

## Design-Stage Diagnostics

```{r design-precision}
dp <- estimate_design_precision(lock, target_mdd = 0.05)
print(dp)
summarize_event_support(lock)
```

---

## Stage 2a: Design diagnostics (Check Point 2)

```{r ps}
ps_fit <- fit_ps_glm(lock, truncate = 0.01)
diag   <- compute_ps_diagnostics(ps_fit)
print(diag)
```

```{r overlap, fig.cap = "PS overlap."}
plot(diag)
```

```{r cp2}
cp2 <- checkpoint_balance(diag, lock_hash = lock$lock_hash)
print(cp2)
```

---

## Stage 2b: Baseline plasmode candidate selection

```{r candidates}
# Define TMLE candidate specifications
candidates <- list(
  tmle_candidate("glm_t01", "GLM trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05)
)
# Or use the default grid:
# candidates <- expand_tmle_candidate_grid()
```

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(lock, tmle_candidates = candidates,
                                 effect_sizes = c(0.05, 0.10),
                                 reps = 40L)
print(plas)
```

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- lock_primary_tmle_spec(lock, selected)
print(selected)
```

---

## Stage 2c: Data-quality stress testing

`run_plasmode_dq_stress()` perturbs the synthetic outcome under
locked data-quality severity ranges. The severity values are not
chosen here; they come from external sources (validation studies,
data profiling, or expert elicitation) and are locked before the
stress test runs.

### Outcome-blind DGP mode

The plasmode + DQ stress test runs in one of three
data-generating-process modes, following the FIORD outcome-blind
simulation spectrum [@nance2026fiord]. The DGP mode is recorded
in the analysis lock alongside the candidates and severity ranges,
and the audit log surfaces which mode was used:

- **`dgp_mode = "investigator_specified"`** — fully external DGP,
  for example a structural-equation specification using
  `simcausal`. Strictly outcome-blind because no part of the
  generator is fit to the study data. Highest risk of
  mis-specification; lowest risk of unblinding.
- **`dgp_mode = "hybrid"`** (current cleanTMLE default) — uses
  the empirical covariate distribution and the fitted
  treatment / censoring mechanism, with an investigator-specified
  or covariate-only marginal outcome model $E[Y \mid W]$. The
  joint $(A, Y) \mid W$ relationship is *not* used. This is the
  middle of the FIORD spectrum and the recommended default for
  most RWE studies.
- **`dgp_mode = "full_empirical"`** — uses a parametric bootstrap
  from the fitted joint $(W, A, Y)$. **Not strictly outcome-blind
  unless run on held-out or external data**, because the outcome
  mechanism is fit to the study data the analyst can still see.
  Documented as a power-user option and labelled non-outcome-blind
  in the audit log.

The current release implements the hybrid mode; the
investigator-specified and full-empirical modes are documented as
planned extensions (see *Planned helpers* below).

The example below shows two scenarios with locked severity values
and the downstream summarisation and gating calls.

```{r dq-stress-example, eval = FALSE}
# Severity values below are illustrative. In practice they are
# locked before this chunk runs and come from:
#   - validation studies (e.g. measured sens/spec for the outcome
#     phenotype against a gold-standard chart review),
#   - data profiling (e.g. observed missingness fractions), or
#   - expert elicitation (e.g. plausible OR ranges for an
#     unmeasured confounder).

dq_results <- run_plasmode_dq_stress(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05),
  reps            = 30L,
  data_quality_scenarios = list(
    # Scenario 1: outcome misclassification, two (sens, spec) pairs
    outcome_misclass = list(
      sensitivity = c(0.95, 0.90),
      specificity = c(0.99, 0.95)
    ),
    # Scenario 2: unmeasured confounding, two beta (OR) values
    unmeasured_confounding = list(
      U_prevalence   = 0.20,
      U_treatment_OR = c(1.5, 2.0),
      U_outcome_OR   = c(1.5, 2.0)
    )
  ),
  verbose = FALSE
)

# Summarise degradation relative to the baseline plasmode
deg <- summarize_dq_degradation(dq_results)
head(deg)

# Score all checkpoints together (GO / FLAG / STOP)
composite <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
print(composite)

# Or check the audit-trail-based pre-outcome gate
gate <- authorize_outcome_analysis(audit)
print(gate)
```

::: {.callout-warning}
## Common misuse

- Do not use DQ stress testing to claim causal validity.
- Do not interpret low-replicate examples inferentially.
- Do not treat truncation as purely numerical when it changes
  the effective target population.
- Do not use the package audit log as a substitute for
  institutional audit trails.
- Do not interpret a GO decision as proof that assumptions hold.
:::

---

## Stage 3: Negative-control checks (Check Point 3)

```{r nc}
nc  <- run_negative_control(lock, "nc_outcome", ps_fit)
cp3 <- checkpoint_residual_bias(nc, lock_hash = lock$lock_hash)
print(nc)
print(cp3)
```

---

## Audit Trail (initialise early for gate)

```{r audit}
audit <- create_audit_log(lock)
audit <- record_stage(audit, "Stage 1a", "Lock created")
audit <- record_checkpoint(audit, cp1)
audit <- record_checkpoint(audit, cp2)
audit <- record_checkpoint(audit, cp3)
```

## Gate and authorisation

```{r gate}
gate <- authorize_outcome_analysis(audit)
print(gate)
```

---

## Stage 4: Authorised primary analysis

```{r crude}
run_crude_workflow(lock)
```

```{r matching}
match_fit <- run_match_workflow(lock, ps_fit)
print(match_fit)
```

```{r iptw}
iptw_fit <- run_iptw_workflow(lock, ps_fit)
print(iptw_fit)
```

```{r tmle}
g_fit    <- fit_tmle_treatment_mechanism(lock, ps_fit)
Q_fit    <- fit_tmle_outcome_mechanism(lock, g_fit)
tmle_upd <- run_tmle_targeting_step(g_fit, Q_fit)
tmle_fit <- extract_tmle_estimate(tmle_upd)
print(tmle_fit)
```

```{r compare}
summarize_cleanroom_results(
  list(Matching = match_fit, IPTW = iptw_fit, TMLE = tmle_fit)
)
```

---

## Sensitivity and Diagnostics

```{r trunc-sens}
sensitivity_truncation(lock, thresholds = c(0.01, 0.05, 0.10))
```

```{r evalue}
compute_evalue(1.3, ci_bound = 1.05)
```

---

## Model Specification DSL

The `specify_models()` + `identify_*()` API provides a pipe-friendly
interface for time-to-event estimation functions.

```{r spec}
spec <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph") |>
  identify_subject(id)
print(spec)
```

---

## Time-to-Event Estimation

```{r ipwrisk, cache = TRUE}
fit_ipw <- estimate_ipwrisk(spec, risk_time = c(12, 24),
                             weight_type = "iptw", nboot = 20, seed = 42)
print(fit_ipw)
```

```{r ipwrisk-plot, fig.cap = "IPTW risk curves."}
plot(fit_ipw)
```

```{r gcomp}
spec_gc <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment)

fit_gc <- estimate_gcomprisk(spec_gc, risk_time = c(12, 24))
print(fit_gc)
```

```{r aipw}
spec_aipw <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph")

fit_aipw <- estimate_aipwrisk(spec_aipw, risk_time = c(12, 24))
print(fit_aipw)
```

```{r ipwhr}
hr_fit <- estimate_ipwhr(spec)
print(hr_fit)
```

```{r compare-tte}
compare_fits(IPW = fit_ipw, GComp = fit_gc, AIPW = fit_aipw, risk_time = 24)
```

---

## Tables and Weight Diagnostics

```{r table1}
make_table1(dat, vars = c("age", "sex", "biomarker", "comorbidity"),
            by = "treatment", smd = TRUE)
```

```{r table2}
make_table2(fit_ipw, risk_time = 24)
```

```{r wt-summary}
make_wt_summary_table(fit_ipw)
```

```{r extreme}
extreme_weights(fit_ipw, k = 5)
```

---

## Audit Trail (continued)

```{r audit-final}
audit <- record_stage(audit, "Stage 4", "Final estimation complete")
export_audit_trail(audit)
```

### Decision Log

```{r decision-log}
audit <- record_decision_log_entry(audit, "Stage 2", "model_specification",
                                   "Selected GLM PS model")
export_decision_log(audit)
summarize_stage_path(audit)
```

```{r manifest}
build_stage_manifest(audit)
```

---

## Real-World Pipeline Integration

When applying cleanTMLE to actual data, several practical concerns
arise: externally fitted PS, dirty covariate matrices, file-based
multi-stage scripts, and matched-cohort analyses.
This section demonstrates the integration utilities that address
those needs.

### Dirty covariates → clean matrix

Real registry data often has missing values, zero-variance columns,
or column names with special characters.
`sanitize_covariates()` handles all of these in one call:

```{r sanitize}
clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  verbose = TRUE)
cat("Covariates after cleaning:", paste(clean$covariates, collapse = ", "), "\n")
```

### Bring your own propensity scores

If PS were estimated outside cleanTMLE (parallel SuperLearner,
Bayesian model, external software), `wrap_ps_fit()` wraps them into
a `ps_fit` so all downstream functions work identically:

```{r wrap-ps}
my_ps <- predict(glm(treatment ~ age + sex + biomarker + comorbidity,
                      data = dat, family = binomial()), type = "response")
ext_ps_fit <- wrap_ps_fit(lock, ps_scores = my_ps, method = "custom_glm")
ext_diag   <- compute_ps_diagnostics(ext_ps_fit)
print(ext_diag$ess)
```

### Serialise lock and audit between stages

In production pipelines, each stage runs as a separate R script.
`save_lock()` / `load_lock()` serialise with automatic hash
re-validation on reload:

```{r save-load-lock}
# Round-trip: save, reload, verify
tmp <- tempfile("lock_", fileext = ".rds")
save_lock(lock, tmp)
lock_rt <- load_lock(tmp)
cat("Hash intact:", identical(lock$lock_hash, lock_rt$lock_hash), "\n")
unlink(tmp)
```

### TMLE-based negative controls

The default `run_negative_control()` uses IPTW.
`run_negative_control_tmle()` is doubly robust:more consistent
with a TMLE-centred pipeline:

```{r nc-tmle}
nc_tmle <- run_negative_control_tmle(lock, "nc_outcome", ps_fit)
print(nc_tmle)
cat("Method:", nc_tmle$method, "\n")
```

### Matched-cohort TMLE

`run_matched_tmle()` runs the full TMLE update on a matched subset
without creating a separate lock:

```{r matched-tmle}
matched_idx <- c(which(dat$treatment == 1)[1:50],
                 which(dat$treatment == 0)[1:50])
matched_fit <- run_matched_tmle(lock, ps_fit, subset_idx = matched_idx,
                                override_clean_room = TRUE)
print(matched_fit)
```

### Diagnostic plots

```{r clever-plot, fig.cap = "Clever covariate H(A,W) by treatment group."}
clever_covariate_plot(ps_fit = ps_fit, lock = lock)
```

`compute_matched_smds()` + `love_plot_threeway()` compare balance
across all three adjustment strategies:

```{r threeway-love, fig.cap = "Unweighted vs IPTW vs matched SMDs."}
m_smds <- compute_matched_smds(dat, "treatment",
  c("age", "sex", "biomarker", "comorbidity"),
  subset_idx = matched_idx)
love_plot_threeway(diag, matched_smds = m_smds)
```

### Composite gating

`gate_all()` evaluates all checkpoints at once ---
convenient when the pipeline has many stages:

```{r gate-all}
composite <- gate_all(cp1, cp2, cp3, allow_flag = TRUE)
print(composite)
```

### Plasmode candidate-grid expansion (workshop-driven, 0.2)

The 0.2 release adds eight axes to the plasmode candidate grid plus four new helpers, motivated by Phillips et al. 2023 *Practical considerations for specifying a super learner*, Gruber et al. 2023 *Developing a Targeted Learning-based SAP* and its supplement, Gruber et al. 2023 *Evaluating and improving RWE with TL*, and Gruber & van der Laan 2009.

**New `tmle_candidate()` arguments** (all back-compatible):

- `variance_method ∈ c("IF","cv_IF","robust")`:IF is the current default; `cv_IF` uses fold-specific Q,g; `robust` requests the upstream `tmle::tmle(variance.method = "tmle")` plug-in variance (Tran et al. 2018).
- `cv_scheme ∈ c("none","cv_tmle","sample_split")`:promotes CV-TMLE to a first-class axis; `cv_V` lets you pin V (otherwise auto-set from the Phillips rule).
- `estimator ∈ c("tmle","aipw","onestep")`:recorded on the spec; AIPW is post-hoc-computable via `compute_aipw(g_fit, Q_fit)`.
- `discrete_sl`:request the discrete super learner (`method.NNLS`, eSL in the dSL pool).
- `screener ∈ c("All","corP","corRank","glmnet","randomForest")`:couples a screener with every Q, g learner inside CV (Phillips p. 1283).
- `tmle_control`:list of upstream `tmle::tmle()` controls (`fluctuation`, `alpha`, `target.gwt`, `automate`, `min.retain`, `cv_Qinit`).
- `match_spec`:list `list(method, ratio, caliper)` for matched-cohort candidates.
- `truncation` now accepts either numeric or one of the named rules `"sqrt_n_ln_n"`, `"fixed_001"`, `"fixed_025"`, `"fixed_05"`. The named rule is resolved at fit time when `n` is known.

**New `expand_tmle_candidate_grid()` arguments**: `estimators`, `cv_schemes`, `variance_methods`, `discrete_sl`, `screeners`, `max_candidates`. The function Cartesian-products the requested axes and caps the grid at `max_candidates`.

**New `select_tmle_candidate()` rules**:

- `rule = "ci_coverage"`:picks the candidate whose oracle coverage is closest to nominal.
- `rule = "min_se"`:picks the candidate with the smallest mean SE.
- `rule = "composite"`:weighted composite of standardised RMSE, coverage penalty, and SE-vs-SD penalty.
- `rule = "fiord_two_stage"`:the FIORD two-stage selector: Stage 1 screens candidates whose oracle coverage is within `fiord_coverage_tol` (default 0.02) of the target; Stage 2 picks the smallest mean SE among survivors. *Variance-method selection on the locked point estimator (FIORD Stage 2 proper)* is still pending; this rule currently selects the point-estimator candidate.

**New helpers**:

- `compute_n_eff(y, family)`:effective sample size `min(n, 5·n_rare)` for binary outcomes (Phillips p. 1281).
- `recommend_cv_V(n_eff)`:the Phillips V rule (V = 20 for n_eff < 500, V = 10 for n_eff < 5,000, ...).
- `build_sl_library(role, n_eff, p, preset)`:four role-aware library presets (`small_n`, `default`, `rich`, `very_rich`), keyed by effective sample size.
- `resolve_truncation_rule(rule, n)`:numeric pass-through or one of the named rules.

**New post-processors**:

- `run_positivity_diagnostics(ps_fit)`:single tidy object with PS C-statistic, per-arm distribution summary, % truncated, bounded-g quantiles, and a near-positivity-violation flag.
- `compute_G_value(estimate, se)`:the additive bias that would push the CI across the null (Gruber 2023 RWE).
- `run_delta_sensitivity(estimate, se, delta_grid)`:causal-gap sensitivity sweep on top of an existing fit (SAP §8.3.4).
- `compute_aipw(g_fit, Q_fit)`:AIPW estimate from the same nuisances the modular TMLE uses, useful as a candidate-grid comparator.

**Example**:

```{r expanded-candidate-example, eval = FALSE}
# Expanded grid: 2 truncations x 1 library x 2 CV schemes x
# 2 variance methods = 8 candidates.
g <- expand_tmle_candidate_grid(
  truncations      = c(0.01, 0.05),
  libraries        = list(default = c("SL.glm", "SL.glmnet", "SL.gam")),
  cv_schemes       = c("none", "cv_tmle"),
  variance_methods = c("IF", "cv_IF"),
  max_candidates   = 16
)
plas <- run_plasmode_feasibility(lock, tmle_candidates = g, reps = 200)
selected <- select_tmle_candidate(
  plas,
  rule = "fiord_two_stage",
  fiord_target_coverage = 0.95,
  fiord_coverage_tol = 0.02
)
```

### Planned helpers

These helpers are recommended outputs and are planned for later cleanTMLE releases; current users should build them from the documented primitives.

- `clean_target_trial_table()`:render a target-trial timing table from an analysis lock and `attach_estimand()` metadata. Build manually from the lock fields and the estimand object for now.
- `checkpoint_dashboard()`:aggregate per-checkpoint records into a single GO / FLAG / STOP dashboard. Build by binding the rows returned by individual `checkpoint_*()` calls and `gate_all()`.
- `run_plasmode_dq_stress(..., scenarios = with_u_gradient(...))`:extend the unmeasured-confounding scenario from a single fixed latent-U strength to a prespecified grid of strengths, so the DQ stress test produces an E-value-style sensitivity curve as a by-product. Current users can approximate this by calling `run_plasmode_dq_stress()` repeatedly with different `unmeasured_U_strength` parameter values and concatenating the metrics.
- `run_preprotocol_plasmode()`:a pre-protocol variant of the plasmode + DQ stress test that takes user-specified covariate distributions and a hypothesised treatment-mechanism strength instead of a real lock. Use case: deciding which dataset to use before a data-use agreement is signed.
- `run_ctmle_candidate()`, `run_bootstrap_variance()`:C-TMLE (collaborative TMLE; Gruber 2009 §3.2) and nonparametric bootstrap variance paths. Both require wrapping a separate upstream package (`ctmle`, `boot`/`HAL`) and are on the 0.3 roadmap.

```{r session-info}
sessionInfo()
```
