---
title: "cleanTMLE Function Reference"
subtitle: "API Patterns and Quick Examples"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{cleanTMLE Function Reference}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

This article is a compact index of every exported function in
**cleanTMLE**, organised by outcome-blind stage.
For a narrative worked example, see the companion vignette
*A Staged Outcome-Blind TMLE Workflow with cleanTMLE*.

```{r load}
library(cleanTMLE)
dat <- sim_func1(n = 400, seed = 42)

# Estimand/sensitivity declarations and stage-path/stage-manifest
# summaries moved to the companion cleanroomGov package in the
# estimation/governance split. Load it when available; chunks that
# call those helpers are guarded on this flag so the reference still
# knits if the companion package is absent.
has_gov <- requireNamespace("cleanroomGov", quietly = TRUE)
if (has_gov) library(cleanroomGov)
```

---

## Outcome-access taxonomy

The function index below labels each exported function by what it
uses internally from the outcome side of the data:

- **No outcome access**: uses W, A, metadata, lock objects,
  synthetic outcomes, or diagnostics not involving Y.
- **Marginal Y only**: uses event counts or marginal outcome
  support (no estimation of the A--Y association).
- **Synthetic Y only**: uses plasmode-generated outcomes; no
  observed Y.
- **Negative-control Y only**: uses prespecified NC outcomes,
  not the primary outcome.
- **Primary Y**: estimates the observed
  treatment-primary-outcome association. Since 0.2.0 the software
  guard on these functions enforces outcome masking always, and a
  recorded authorisation only on locks that opted into enforcement
  (`create_analysis_lock(enforce = TRUE)`).

These labels describe what exported package functions use
internally, not what a user with raw data access could inspect
outside the package.

## The 0.2.0 workflow: decide the estimand before outcomes

The package's core question is whether a comparison is estimable, and with
which estimand, before the outcome is unlocked. The workflow is five verbs:
lock, assess, map, simulate, estimate. The example below builds a cohort
with a severe practical positivity violation, the situation the Rescue.Co
case study meets on its protocol contrast, and shows the design stage
routing the analysis to a feasible estimand.

```{r workflow-020, message=FALSE}
set.seed(7)
n  <- 800
x1 <- rnorm(n); x2 <- rbinom(n, 1, 0.4)
g  <- plogis(-2.2 + 3.2 * x1 + 0.5 * x2)   # a covariate nearly separates arms
A  <- rbinom(n, 1, g)
y  <- rbinom(n, 1, plogis(-2 + 0.5 * x1 + 0.4 * A))
sev_dat <- data.frame(x1 = x1, x2 = x2, treatment = A, outcome = y)

sev_lock <- create_analysis_lock(sev_dat, "treatment", "outcome",
                                 c("x1", "x2"),
                                 sl_library = "SL.glm", seed = 7L)
sev_lock <- declare_estimand_ladder(sev_lock, primary = "ATE",
                                    fallbacks = c("trimmed_ATE", "ATT",
                                                  "ATO"))

sev_ps  <- fit_ps(sev_lock, method = "glm")
sev_sup <- assess_support(sev_ps)
sev_sup
```

The verdict travels with every downstream estimate. `plot(sup)` draws the
mirrored propensity histogram with the reference band shaded. The
feasibility table then says which estimands the same fit can support: the
ATE's weights explode where the ATT's control weights stay bounded by
`max g / (1 - g)` and the ATO's weights are bounded by one by construction.

```{r feasibility-020}
sev_fea <- estimand_feasibility(sev_ps, profile_vars = c("x1", "x2"))
sev_fea
head(sev_fea$unsupported)
```

The outcome-blind support simulation asks the same question dynamically:
across a prespecified family of outcome surfaces (confounding along the
propensity direction, effect modification along it, optional curvature),
which estimators stay within a bias tolerance at near-nominal coverage.
Replicates are generated under the generate-treatment plasmode design;
keeping the observed treatment (the pre-0.2.0 baseline) induces a
positivity violation by construction (Shaw et al. 2025) and survives only
as a deprecated option.

```{r simulate-020, eval=FALSE}
sev_sim <- simulate_support(sev_lock, sev_ps, reps = 200)
plot(sev_sim)      # the support map the review reads
```

Estimation runs the declared ladder: the primary when feasible, otherwise
every feasible fallback, each row labelled with its estimand, verdict,
caveat and implausibility flags, and every switch logged.

```{r ladder-020, message=FALSE, warning=FALSE}
res <- run_estimand_ladder(sev_lock, sev_ps, sl_library = "SL.glm",
                           verbose = FALSE)
res
```

`design_report(sev_lock, sev_sup, sev_fea)` assembles what a review team reads before
outcomes are unlocked, with a one-line recommendation; its framing follows
the blinded validity-diagnostics gate of Conover et al. (2025), where a
failed comparison is labelled inestimable rather than estimated anyway.

---

## Function Index

The table below lists every exported function, its outcome-blind
stage, and what it uses from the outcome side of the data. A few
rows marked *(cleanroomGov)* are governance helpers that moved to
the companion **cleanroomGov** package in the estimation/governance
split; they are documented here for continuity but are exported by
`cleanroomGov`, not `cleanTMLE`. Functions superseded in 0.2.0 (the
checkpoint, gate-token, audit and decision-log layer) keep working,
emit a once-per-session pointer to their replacement, and are grouped
under Superseded on the pkgdown reference index; their rows below are
kept for continuity.

| Stage | Function | Outcome? | Purpose |
|-------|----------|:--------:|---------|
| Design | `fit_ps()` | No | One front door for propensity fitting: `method = "superlearner"`, `"glm"`, or `"external"` with `scores` |
| Design | `assess_support()` | No | Graded overlap verdict (PASS/FLAG/SEVERE/FAIL), balance table, near-deterministic strata, tree search, mirrored plot; `thresholds =` a named list of overrides |
| Design | `estimand_feasibility()` | No | Per-estimand weights, ESS, removals, target population, verdict; `profile_vars =` adds the removed-versus-kept profile as `$unsupported` |
| Design | `declare_estimand_ladder()` | No | Pre-register primary, fallbacks, trigger, sensitivity floors; `candidate =` binds the selected TMLE specification |
| Design | `simulate_support()` | Synthetic | Outcome-blind support map (generate-treatment plasmode, per-estimand truths); `surface =` a named list; `design = "parametric_bootstrap"` gives the Petersen check, labelled optimistic |
| Design | `design_report()` | No | The pre-outcome bundle a review team reads, with a recommendation and the TARGET emulation table (`$emulation`, `protocol =`) |
| Design | `create_contrast_locks()` | No | One lock per contrast from a multi-level treatment, shared covariates and controls |
| Design | `run_negative_control_ladder()` | NC | Registered controls on nested cohorts; names what turns null after restriction |
| Design | `check_process_indicators()` | No | Care-process collider check conditional on severity |
| 4 | `estimate_effect()` | Primary | The one estimation front door: `estimand = "ATE"/"trimmed_ATE"/"ATT"/"ATO"/"matched_ATT"`, `estimator = "tmle"/"iptw"/"match"/"crude"`, `missing = "auto"/"ipcw"/"complete_case"`; the implausibility guard rides on every result |
| 4 | `run_estimand_ladder()` | Primary | Estimate the declared primary when feasible plus every feasible fallback, logged |
| Util | `SL.glm.pca()`, `SL.glm.pca5()`, `SL.glm.pca10()` | n/a | PCA-GLM SuperLearner learners (rotation fitted per fold) |
| Util | `SL.glmnet.ridge()`, `SL.glmnet.enet()` | n/a | Ridge and elastic-net glmnet learners |
| Util | `expit()`, `logit()` | n/a | Inverse-logit / logit transforms |
| Data | `sim_func1()` | n/a | Simulate causal-inference data |
| 1a | `create_analysis_lock()` | No | Lock analytic specification |
| 1a | `validate_analysis_lock()` | No | Verify lock integrity |
| 1a | `attach_estimand()` | No | Attach estimand metadata *(cleanroomGov)* |
| 1a | `declare_sensitivity_plan()` | No | Declare a sensitivity plan *(cleanroomGov)* |
| 1a | `define_negative_control()` | No | Register a negative control variable |
| 1b | `checkpoint_cohort_adequacy()` | Marginal Y only | Check Point 1: sample size / events |
| 1b | `estimate_design_precision()` | Marginal Y only | Design-stage precision/MDD |
| 1b | `summarize_event_support()` | Marginal Y only | Event counts by arm |
| 2 | `fit_ps_glm()` | No | Logistic regression PS |
| 2 | `fit_ps_superlearner()` | No | SuperLearner PS |
| 2 | `compute_ps_diagnostics()` | No | Overlap, ESS, SMDs |
| 2 | `checkpoint_balance()` | No | Check Point 2: balance decision |
| 2b | `tmle_candidate()` | No | Define a TMLE candidate specification |
| 2b | `expand_tmle_candidate_grid()` | No | Generate candidate grid |
| 2b | `validate_tmle_candidates()` | No | Validate candidate list |
| 2b | `run_plasmode_feasibility()` | Synthetic Y only | Plasmode simulation evaluation |
| 2b | `summarize_plasmode_results()` | No | Print plasmode metrics |
| 2b | `select_tmle_candidate()` | No | Select best TMLE specification |
| 2b | `lock_primary_tmle_spec()` | No | Lock selected spec in the lock |
| 2b | `get_primary_tmle_spec()` | No | Retrieve locked spec |
| 2b | `gate_check()` | No | GO / FLAG / STOP for plasmode |
| 3 | `run_negative_control()` | Negative-control Y only | Negative control analysis |
| 3 | `run_residual_confounding_stage()` | Negative-control Y only | Stage 3 wrapper |
| 3 | `checkpoint_residual_bias()` | No | Check Point 3: residual bias |
| Gate | `authorize_outcome_analysis()` | No | Pre-outcome GO/STOP gate |
| Gate | `assert_outcome_authorized()` | No | Gate assertion (errors if not GO) |
| 4 | `run_crude_workflow()` | Primary Y | Unadjusted risk difference |
| 4 | `run_match_workflow()` | Primary Y | PS matching |
| 4 | `run_iptw_workflow()` | Primary Y | Stabilised IPTW |
| 4 | `fit_tmle_treatment_mechanism()` | No | TMLE g-step (re-uses PS) |
| 4 | `fit_tmle_outcome_mechanism()` | Primary Y | TMLE Q-step |
| 4 | `run_tmle_targeting_step()` | Primary Y | TMLE targeting / fluctuation |
| 4 | `extract_tmle_estimate()` | Primary Y | Final TMLE estimate |
| 4 | `fit_tmle_candidate_set()` | Primary Y | Fit multiple TMLE specs |
| 4 | `fit_final_workflows()` | Primary Y | Run all Stage 4 workflows |
| 4 | `sensitivity_truncation()` | Primary Y | Truncation sensitivity |
| Any |`compute_evalue()` | No | E-value for unmeasured confounding |
| Any |`summarize_cleanroom_results()` | No | Cross-workflow comparison table |
| Any |`record_decision_log_entry()` | No | Add structured decision log entry |
| Any |`export_decision_log()` | No | Export decision log as data frame |
| Any |`summarize_stage_path()` | No | Compact stage narrative *(cleanroomGov)* |
| Any |`create_audit_log()` | No | Initialise audit trail |
| Any |`record_stage()` | No | Append stage entry |
| Any |`record_checkpoint()` | No | Append checkpoint entry |
| Any |`export_audit_trail()` | No | Export trail as data frame |
| Any |`build_stage_manifest()` | No | Print stage path summary *(cleanroomGov)* |
| Util | `mask_outcome()` | No | Mask outcome column with NA |
| Util | `unmask_outcome()` | No | Restore masked outcome |
| Wrapper | `run_clean_tmle()` | Primary Y | Full staged clean-room pipeline |
| Wrapper | `init_decision_log()` | No | Create empty decision log |
| Wrapper | `log_decision_entry()` | No | Append row to decision log |
| Wrapper | `save_decision_log()` | No | Write decision log to CSV |
| Plot | `love_plot()` | No | SMD dot-plot (weighted vs unweighted) |
| Plot | `ic_histogram()` | No | Influence curve histogram |
| Integ | `wrap_ps_fit()` | No | Wrap external PS scores into ps_fit |
| Integ | `fit_ps_parallel()` | No | SuperLearner PS with parallel cluster |
| Integ | `save_lock()` / `load_lock()` | No | Serialise / deserialise lock to RDS |
| Integ | `save_audit()` / `load_audit()` | No | Serialise / deserialise audit log |
| 3 | `run_negative_control_tmle()` | Negative-control Y only | TMLE-based negative control (doubly robust) |
| 4 | `run_matched_tmle()` | Primary Y | Modular TMLE on a matched subset |
| Integ | `sanitize_covariates()` | No | Clean covariates for SuperLearner |
| Integ | `create_analysis_lock_from_yaml()` | No | Lock from YAML config file |
| Plot | `clever_covariate_plot()` | No | H(A,W) distribution by group |
| Plot | `love_plot_threeway()` | No | Unweighted / IPTW / matched SMDs |
| Integ | `compute_matched_smds()` | No | SMDs on a matched subset |
| 1 | `attrition_table()` | No outcome access | Cohort flow / attrition documentation |
| 1 | `get_final_cohort()` | No | Extract filtered data from attrition table |
| 1 | `clean_event_process_table()` | No outcome access | Event-process classification table *(cleanroomGov)* |
| 1 | `clean_check_event_processes()` | No outcome access | Verify event-process specification *(cleanroomGov)* |
| 2 | `clean_weight_diagnostics()` | No outcome access | Weight ESS, percentiles, max-weight thresholds |
| 3  | `refine_ps_after_nco()` | Negative-control Y only | Iterative PS refinement after NCO signal |
| 2c | `run_plasmode_dq_stress()` | Synthetic Y only | DQ stress test (missingness, misclass, U) |
| 2c | `summarize_dq_degradation()` | No | DQ degradation comparison table |
| Gate | `gate_all()` | No | Composite GO/FLAG/STOP across checkpoints |

**Marginal Y only**: accesses event counts or marginal outcome support, not the A--Y association.
**Synthetic Y only**: generates simulated outcomes; does not use the real outcome association.
**Negative-control Y only**: accesses prespecified negative-control outcomes, not the primary outcome.
**Primary Y**: estimates the observed treatment--primary-outcome association and requires authorization.

---

## Utilities

`expit()` and `logit()` went internal in the 0.2.0 consolidation
(`stats::plogis()` and `stats::qlogis()` are the base equivalents):

```{r utils}
cleanTMLE:::expit(0)                       # 0.5
cleanTMLE:::logit(0.5)                     # 0
cleanTMLE:::expit(cleanTMLE:::logit(0.73)) # 0.73
```

---

## Stage 1a: Lock, Estimand, Plans

```{r lock}
lock <- create_analysis_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  sl_library = c("SL.glm", "SL.mean"),
  negative_controls = "nc_outcome",
  plasmode_reps = 40L,
  seed       = 42L
)
lock <- declare_estimand_ladder(lock,
  primary   = "ATE",
  fallbacks = c("trimmed_ATE", "ATT", "ATO"),
  trigger   = "SEVERE")
print(lock)
```

`attach_estimand()` and `declare_sensitivity_plan()` moved to the
companion **cleanroomGov** package (see the Function Index). When
that package is installed they attach estimand metadata and a
sensitivity plan to the lock:

```{r estimand, eval = has_gov}
lock <- attach_estimand(lock,
  description          = "Treatment effect on 24-month event risk",
  contrast             = "risk_difference",
  statistical_estimand = "E_W[E(Y|A=1,W) - E(Y|A=0,W)]"
)

lock <- declare_sensitivity_plan(lock, "truncation",
  description = "Vary PS truncation",
  settings    = list(thresholds = c(0.01, 0.05, 0.10)))
```

### Outcome Masking

```{r masking}
masked <- mask_outcome(lock)
all(is.na(masked$data[["event_24"]]))
unmasked <- unmask_outcome(masked, lock)
```

### Lock persistence

```{r persistence}
p <- tempfile(fileext = ".rds")
save_lock(lock, p)
lock <- load_lock(p)   # re-validates the fingerprint
unlink(p)
```

---

## Stage 1b: Design precision and event support

```{r design-precision}
dp <- estimate_design_precision(lock, target_mdd = 0.05)
print(dp)
dp$event_support
```

---

## Stage 2a: Propensity, support verdict, feasibility, balance

```{r ps}
ps_fit <- fit_ps(lock, method = "glm", truncate = 0.01)
sup    <- assess_support(ps_fit)
print(sup)
```

```{r overlap, fig.cap = "Mirrored propensity histograms with the band shaded."}
plot(sup)
```

```{r love-fns, fig.cap = "Standardised mean differences."}
love_plot(sup)
```

```{r feasibility-fns}
fea <- estimand_feasibility(ps_fit)
print(fea)
```

---

## Stage 2b: Support map and plasmode candidate selection

```{r support-sim-fns, cache = TRUE}
sim <- simulate_support(lock, ps_fit, reps = 10, verbose = FALSE)
print(sim)
```

```{r candidates}
# Define TMLE candidate specifications
candidates <- list(
  tmle_candidate("glm_t01", "GLM trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05)
)
```

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(lock, tmle_candidates = candidates,
                                 effect_sizes = c(0.05, 0.10),
                                 reps = 40L)
print(plas)
```

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- declare_estimand_ladder(lock,
  primary   = "ATE",
  fallbacks = c("trimmed_ATE", "ATT", "ATO"),
  trigger   = "SEVERE",
  candidate = selected)
print(selected)
```

---

## Stage 2c: Data-quality stress testing

`run_plasmode_dq_stress()` perturbs the synthetic outcome under
locked data-quality severity ranges. The severity values are not
chosen here; they come from external sources (validation studies,
data profiling, or expert elicitation) and are locked before the
stress test runs.

### Outcome-blind DGP mode

The plasmode + DQ stress test runs in one of three
data-generating-process modes, following the FIORD outcome-blind
simulation spectrum [@nance2026fiord]. The DGP mode is recorded
in the analysis lock alongside the candidates and severity ranges,
and the audit log surfaces which mode was used:

- **`dgp_mode = "investigator_specified"`**: fully external DGP,
  for example a structural-equation specification using
  `simcausal`. Strictly outcome-blind because no part of the
  generator is fit to the study data. Highest risk of
  mis-specification; lowest risk of unblinding.
- **`dgp_mode = "hybrid"`** (current cleanTMLE default): uses
  the empirical covariate distribution and the fitted
  treatment / censoring mechanism, with an investigator-specified
  or covariate-only marginal outcome model $E[Y \mid W]$. The
  joint $(A, Y) \mid W$ relationship is *not* used. This is the
  middle of the FIORD spectrum and the recommended default for
  most RWE studies.
- **`dgp_mode = "full_empirical"`**: uses a parametric bootstrap
  from the fitted joint $(W, A, Y)$. **Not strictly outcome-blind
  unless run on held-out or external data**, because the outcome
  mechanism is fit to the study data the analyst can still see.
  Documented as a power-user option and labelled non-outcome-blind
  in the audit log.

The current release implements the hybrid mode; the
investigator-specified and full-empirical modes are documented as
planned extensions (see *Planned helpers* below).

The example below shows two scenarios with locked severity values
and the downstream summarisation and gating calls.

```{r dq-stress-example, eval = FALSE}
# Severity values below are illustrative. In practice they are
# locked before this chunk runs and come from:
#   - validation studies (e.g. measured sens/spec for the outcome
#     phenotype against a gold-standard chart review),
#   - data profiling (e.g. observed missingness fractions), or
#   - expert elicitation (e.g. plausible OR ranges for an
#     unmeasured confounder).

dq_results <- run_plasmode_dq_stress(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05),
  reps            = 30L,
  data_quality_scenarios = list(
    # Practical positivity violation
    near_positivity = list(slopes = c(2.0, 3.0)),
    # Covariate missingness: MCAR, MAR (treatment-dependent), MNAR
    # (value-dependent). All median-imputed; MAR/MNAR bias the imputation.
    covariate_missingness      = list(fractions = c(0.10, 0.20)),
    covariate_missingness_mar  = list(fractions = c(0.10, 0.20), treatment_OR = 3),
    covariate_missingness_mnar = list(fractions = c(0.10, 0.20), strength = 1.5),
    # Outcome misclassification, two (sens, spec) pairs
    outcome_misclass = list(
      sensitivity = c(0.95, 0.90),
      specificity = c(0.99, 0.95)
    ),
    # Unmeasured confounding, two (OR) values
    unmeasured_confounding = list(
      U_prevalence   = 0.20,
      U_treatment_OR = c(1.5, 2.0),
      U_outcome_OR   = c(1.5, 2.0)
    )
  ),
  verbose = FALSE
)

# Summarise degradation relative to the baseline plasmode
deg <- summary(dq_results)
head(deg)
```

The stress test is robust to non-converging cells. If a degraded scenario
is severe enough that fewer than two replicates converge for a candidate,
that cell reports `NA` for `emp_sd` and `se_cal` (and drops out of the
SE-calibration summary) rather than aborting the run; `bias`, `rmse`, and
`coverage` are still reported from the replicates that did converge, and
`n_converged` records how many contributed to each row.

::: {.callout-warning}
## Common misuse

- Do not use DQ stress testing to claim causal validity.
- Do not interpret low-replicate examples inferentially.
- Do not treat truncation as purely numerical when it changes
  the effective target population.
- Do not use the design log as a substitute for institutional
  audit trails.
- Do not interpret a proceed recommendation as proof that
  assumptions hold.
:::

---

## Stage 3: Negative controls and the collider check

```{r nc}
nc <- run_negative_control_ladder(lock,
  restrictions = list("comorbidity-free" = lock$data$comorbidity == 0),
  method = "tmle", ps_method = "glm", verbose = FALSE)
print(nc)
```

```{r process}
proc <- check_process_indicators(lock, indicators = "biomarker")
print(proc)
```

(The `biomarker` column stands in for a care-process indicator here;
in a real study the indicators are post-treatment care-process
variables kept out of the adjustment set.)

## Pre-outcome decision: the design report

```{r report}
design_report(lock, sup, fea, simulation = sim, nc_ladder = nc)
```

---

## Stage 4: Estimation

```{r ladder-est}
fit <- run_estimand_ladder(lock, ps_fit)
print(fit)
```

```{r effect}
tmle_fit  <- estimate_effect(lock, ps_fit, estimand = "ATE",
                             estimator = "tmle", return_steps = TRUE)
crude_fit <- estimate_effect(lock, estimand = "ATE", estimator = "crude")
iptw_fit  <- estimate_effect(lock, ps_fit, estimand = "ATE",
                             estimator = "iptw")
match_fit <- estimate_effect(lock, ps_fit, estimand = "matched_ATT")
print(tmle_fit)
```

```{r compare, fig.cap = "Cross-estimator forest plot; TMLE is the locked primary."}
forest_plot(list(Crude = crude_fit, IPTW = iptw_fit,
                 `Matched ATT` = match_fit, TMLE = tmle_fit))
```

---

## Sensitivity and Diagnostics

```{r trunc-sens}
resolve_truncation_rule("sqrt_n_ln_n", n = nrow(dat))  # adaptive default
```

```{r delta-evalue}
ate <- tmle_fit$estimates$ATE
head(run_delta_sensitivity(ate$estimate, ate$se))
compute_evalue(1.3, ci_bound = 1.05)
```

---

## Model Specification DSL

::: {.callout-warning}
## Experimental: outside the tested v0.1 scope

The `specify_models()` DSL and the time-to-event estimators that
follow (`estimate_ipwrisk()`, `estimate_gcomprisk()`,
`estimate_aipwrisk()`, `estimate_ipwhr()`) are experimental
extensions outside the tested v0.1 scope, which covers binary
point-treatment studies with a marginal risk-difference estimand.
They emit a runtime warning to that effect and their interfaces and
results should be treated as provisional.
:::

The `specify_models()` + `identify_*()` API provides a pipe-friendly
interface for time-to-event estimation functions.

```{r spec}
spec <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph") |>
  identify_subject(id)
print(spec)
```

---

## Time-to-Event Estimation

```{r ipwrisk, cache = TRUE}
fit_ipw <- estimate_ipwrisk(spec, risk_time = c(12, 24),
                             weight_type = "iptw", nboot = 20, seed = 42)
print(fit_ipw)
```

```{r ipwrisk-plot, fig.cap = "IPTW risk curves."}
plot(fit_ipw)
```

```{r gcomp}
spec_gc <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment)

fit_gc <- estimate_gcomprisk(spec_gc, risk_time = c(12, 24))
print(fit_gc)
```

```{r aipw}
spec_aipw <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph")

fit_aipw <- estimate_aipwrisk(spec_aipw, risk_time = c(12, 24))
print(fit_aipw)
```

```{r ipwhr}
hr_fit <- estimate_ipwhr(spec)
print(hr_fit)
```

```{r compare-tte}
# Every estimate_*risk() fit carries a $risk table with the same core
# columns, so the estimators compare with plain data-frame work.
tte <- list(IPW = fit_ipw, GComp = fit_gc, AIPW = fit_aipw)
do.call(rbind, lapply(names(tte), function(nm) {
  r <- tte[[nm]]$risk
  cbind(estimator = nm,
        r[r$time == 24, c("time", "group", "risk", "surv")])
}))
```

---

## Tables and Weight Diagnostics

```{r table1}
make_table1(dat, vars = c("age", "sex", "biomarker", "comorbidity"),
            by = "treatment", smd = TRUE)
```

```{r table2}
make_table2(fit_ipw, risk_time = 24)
```

```{r wt-diag}
A <- lock$data[[lock$treatment]]
w <- ifelse(A == 1, 1 / ps_fit$ps, 1 / (1 - ps_fit$ps))
clean_weight_diagnostics(weights = w, treatment = A,
                         covariates = lock$data[, lock$covariates])
```

---

## The design log

The design log lives on the lock: one timestamped row per recorded
decision (ladder declarations, candidate attachment, estimand
switches, overrides). `save_lock()` archives it with the
specification, and ladder results carry the entries their run added.

```{r design-log}
knitr::kable(lock$design_log)
```

`summarize_stage_path()` and `build_stage_manifest()` moved to the
companion **cleanroomGov** package (see the Function Index). When it
is installed they render a compact stage narrative and a stage-path
manifest from that package's audit-trail objects:

```{r stage-path, eval = FALSE}
summarize_stage_path(audit)   # cleanroomGov
build_stage_manifest(audit)   # cleanroomGov
```

---

## Real-World Pipeline Integration

When applying cleanTMLE to actual data, several practical concerns
arise: externally fitted PS, dirty covariate matrices, file-based
multi-stage scripts, and matched-cohort analyses.
This section demonstrates the integration utilities that address
those needs.

### Dirty covariates → clean matrix

Real registry data often has missing values, zero-variance columns,
or column names with special characters.
`sanitize_covariates()` handles all of these in one call:

```{r sanitize}
clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  verbose = TRUE)
cat("Covariates after cleaning:", paste(clean$covariates, collapse = ", "), "\n")
```

### Bring your own propensity scores

If PS were estimated outside cleanTMLE (parallel SuperLearner,
Bayesian model, external software), `fit_ps(method = "external")`
wraps them into a `ps_fit` so all downstream functions work
identically:

```{r wrap-ps}
my_ps <- predict(glm(treatment ~ age + sex + biomarker + comorbidity,
                      data = dat, family = binomial()), type = "response")
ext_ps_fit <- fit_ps(lock, method = "external", scores = my_ps)
ext_sup    <- assess_support(ext_ps_fit)
ext_sup$verdict
```

### Serialise lock and audit between stages

In production pipelines, each stage runs as a separate R script.
`save_lock()` / `load_lock()` serialise with automatic hash
re-validation on reload:

```{r save-load-lock}
# Round-trip: save, reload, verify
tmp <- tempfile("lock_", fileext = ".rds")
save_lock(lock, tmp)
lock_rt <- load_lock(tmp)
cat("Hash intact:", identical(lock$lock_hash, lock_rt$lock_hash), "\n")
unlink(tmp)
```

### Negative controls, doubly robust

The negative-control ladder's `method = "tmle"` (the default) fits
each control doubly robustly, with the propensity refitted per rung;
`method = "unadjusted"` gives the two-proportion balance-check
reading. Both were shown in Stage 3 above.

### Matched-cohort TMLE

`estimate_effect(estimand = "matched_ATT")` matches on the fitted
propensity (greedy 1:1 within a caliper) and runs the full TMLE
update on the matched subset, without creating a separate lock. The
matched analysis was shown in Stage 4 above.

### Diagnostic plots

```{r clever-plot, fig.cap = "Clever covariate H(A,W) by treatment group."}
plot(tmle_fit, type = "clever")
```

```{r ic-plot, fig.cap = "Efficient influence curve distribution."}
plot(tmle_fit, type = "ic")
```

`love_plot(sup, matched = <data.frame>)` adds a matched-cohort SMD
series (columns `variable`, `smd`) to the unweighted-versus-weighted
display shown at Stage 2a.

### Plasmode candidate-grid expansion (workshop-driven, 0.2)

The 0.2 release adds eight axes to the plasmode candidate grid plus four new helpers, motivated by Phillips et al. 2023 *Practical considerations for specifying a super learner*, Gruber et al. 2023 *Developing a Targeted Learning-based SAP* and its supplement, Gruber et al. 2023 *Evaluating and improving RWE with TL*, and Gruber & van der Laan 2009.

**New `tmle_candidate()` arguments** (all back-compatible):

- `variance_method ∈ c("IF","cv_IF","robust")`:IF is the current default; `cv_IF` uses fold-specific Q,g; `robust` requests the upstream `tmle::tmle(variance.method = "tmle")` plug-in variance (Tran et al. 2018).
- `cv_scheme ∈ c("none","cv_tmle","sample_split")`:promotes CV-TMLE to a first-class axis; `cv_V` lets you pin V (otherwise auto-set from the Phillips rule).
- `estimator ∈ c("tmle","aipw","onestep")`:recorded on the spec; AIPW is post-hoc-computable via `compute_aipw(g_fit, Q_fit)`.
- `discrete_sl`:request the discrete super learner (`method.NNLS`, eSL in the dSL pool).
- `screener ∈ c("All","corP","corRank","glmnet","randomForest")`:couples a screener with every Q, g learner inside CV (Phillips p. 1283).
- `tmle_control`:list of upstream `tmle::tmle()` controls (`fluctuation`, `alpha`, `target.gwt`, `automate`, `min.retain`, `cv_Qinit`).
- `match_spec`:list `list(method, ratio, caliper)` for matched-cohort candidates.
- `truncation` now accepts either numeric or one of the named rules `"sqrt_n_ln_n"`, `"fixed_001"`, `"fixed_025"`, `"fixed_05"`. The named rule is resolved at fit time when `n` is known.

**Grid expansion**: `expand_tmle_candidate_grid()` (internal since the
0.2.0 consolidation; reach it with `cleanTMLE:::` if needed)
Cartesian-products the requested axes (`estimators`, `cv_schemes`,
`variance_methods`, `discrete_sl`, `screeners`) and caps the grid at
`max_candidates`; most analyses declare a handful of candidates
directly with `tmle_candidate()`.

**New `select_tmle_candidate()` rules**:

- `rule = "ci_coverage"`:picks the candidate whose oracle coverage is closest to nominal.
- `rule = "min_se"`:picks the candidate with the smallest mean SE.
- `rule = "composite"`:weighted composite of standardised RMSE, coverage penalty, and SE-vs-SD penalty.
- `rule = "fiord_two_stage"`:the FIORD two-stage selector: Stage 1 screens candidates whose oracle coverage is within `fiord_coverage_tol` (default 0.02) of the target; Stage 2 picks the smallest mean SE among survivors. *Variance-method selection on the locked point estimator (FIORD Stage 2 proper)* is still pending; this rule currently selects the point-estimator candidate.

**Helpers** (exported):

- `build_sl_library(role, n_eff, p, preset)`:role-aware library presets (`small_n`, `default`, `rich`, `very_rich`, `rwe_wide`), keyed by effective sample size; computes `n_eff` from `y` when not supplied (Phillips p. 1281) and returns the recommended V.
- `resolve_truncation_rule(rule, n)`:numeric pass-through or one of the named rules, including the adaptive `"sqrt_n_ln_n"`.

**Post-processors**:

- `assess_support(ps_fit)`:the graded overlap assessment (replaces the earlier `run_positivity_diagnostics()` tidy object, which now wraps it).
- `run_delta_sensitivity(estimate, se, delta_grid)`:additive-bias sensitivity sweep on top of an existing fit, with the g-value (the bias that pushes the CI across the null; Gruber 2023 RWE) attached.
- `compute_aipw(g_fit, Q_fit)` (internal):AIPW estimate from the same nuisances the modular TMLE uses.

**Example**:

```{r expanded-candidate-example, eval = FALSE}
# Expanded grid: 2 truncations x 1 library x 2 CV schemes x
# 2 variance methods = 8 candidates.
g <- cleanTMLE:::expand_tmle_candidate_grid(
  truncations      = c(0.01, 0.05),
  libraries        = list(default = c("SL.glm", "SL.glmnet", "SL.gam")),
  cv_schemes       = c("none", "cv_tmle"),
  variance_methods = c("IF", "cv_IF"),
  max_candidates   = 16
)
plas <- run_plasmode_feasibility(lock, tmle_candidates = g, reps = 200)
selected <- select_tmle_candidate(
  plas,
  rule = "fiord_two_stage",
  fiord_target_coverage = 0.95,
  fiord_coverage_tol = 0.02
)
```

### Planned helpers

These helpers are recommended outputs and are planned for later cleanTMLE releases; current users should build them from the documented primitives.

- `clean_target_trial_table()`:render a target-trial timing table from an analysis lock and `attach_estimand()` metadata. `design_report(protocol = ...)` renders the TARGET-style emulation table today; build richer versions from the lock fields for now.
- `run_plasmode_dq_stress(..., scenarios = with_u_gradient(...))`:extend the unmeasured-confounding scenario from a single fixed latent-U strength to a prespecified grid of strengths, so the DQ stress test produces an E-value-style sensitivity curve as a by-product. Current users can approximate this by calling `run_plasmode_dq_stress()` repeatedly with different `unmeasured_U_strength` parameter values and concatenating the metrics.
- `run_preprotocol_plasmode()`:a pre-protocol variant of the plasmode + DQ stress test that takes user-specified covariate distributions and a hypothesised treatment-mechanism strength instead of a real lock. Use case: deciding which dataset to use before a data-use agreement is signed.
- `run_ctmle_candidate()`, `run_bootstrap_variance()`:C-TMLE (collaborative TMLE; Gruber 2009 §3.2) and nonparametric bootstrap variance paths. Both require wrapping a separate upstream package (`ctmle`, `boot`/`HAL`) and are on the 0.3 roadmap.

```{r session-info}
sessionInfo()
```
