---
title: "A Staged Outcome-Blind TMLE Workflow with cleanTMLE"
subtitle: "Worked Example"
author: "cleanTMLE Authors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
vignette: >
  %\VignetteIndexEntry{A Staged Outcome-Blind TMLE Workflow with cleanTMLE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  warning    = FALSE,
  message    = FALSE
)
library(cleanTMLE)
```

## Read this first

This vignette has two halves. The **first half** teaches the core estimator and shows what a plain end-to-end analysis looks like using `cleanTMLE` as a regular RWE estimation package. The **second half** adds the staged-analysis layer (the analysis lock, the support and feasibility stage, the estimand ladder, the design report and its log) for studies that need a reproducible pre-outcome dossier.

If you are new to targeted-learning methods, read the first half straight through. The vocabulary you need is in the **Concept primer** below. The governance vocabulary (clean room, dossier, design log) does not appear until the staged-workflow half; by then you will have already seen Table 1, the propensity-score overlap, and a risk-curve plot, so the governance ideas have something concrete to attach to.

## A clinical anchor

We use a small simulated cohort (`sim_func1()`) that we will treat as if it came from a pharmaco-epidemiology study. Read it as:

- **Population.** Adult patients enrolled at a tertiary hospital network.
- **Treatment** (`treatment`). A new oral medication started at enrolment versus standard care.
- **Outcome** (`event_24`). A binary event at 24 months (for example, a major cardiovascular event).
- **Baseline covariates** (`age`, `sex`, `biomarker`, `comorbidity`). Standard pre-treatment demographics and risk factors recorded at enrolment.

The numbers are simulated, but the question we are asking is the one a real pharmaco-epi study would ask: **does the new medication change the 24-month event risk, after we adjust for baseline differences between the treated and untreated groups?**

```{r clinical-anchor}
set.seed(2026)
dat <- sim_func1(n = 600, seed = 2026)
head(dat[, c("treatment", "event_24", "age", "sex", "biomarker",
             "comorbidity")])
cat("\nGroup sizes:\n")
table(treatment = dat$treatment, useNA = "ifany")
cat("\nRaw 24-month event rates by treatment arm:\n")
round(tapply(dat$event_24, dat$treatment, mean, na.rm = TRUE), 3)
```

## Concept primer

Eight terms that appear repeatedly in this vignette and in the targeted-learning literature, in plain language:

- **Estimand.** The exact quantity you want to estimate. For us: the population-average 24-month event-risk difference if everyone took the new medication versus if no one did.
- **Propensity score (PS).** The probability of receiving treatment given a person's baseline covariates. Predicted from a model fit on the data.
- **Overlap (positivity).** Whether treated and untreated groups have similar baseline profiles. Poor overlap means adjustment cannot recover the marginal effect for the full cohort.
- **SMD (standardised mean difference).** A balance metric: the difference in covariate means between treated and untreated, standardised by the pooled SD. Smaller is better; < 0.10 is the convention.
- **ESS (effective sample size, Kish).** The "real" sample size after weighting; a few heavy weights can shrink ESS far below `n`.
- **TMLE (targeted maximum-likelihood estimation).** A four-step recipe: (1) predict the outcome from covariates and treatment; (2) predict the propensity score; (3) *target* the outcome predictions using the propensity score so the resulting treatment-effect estimate has the right calibration; (4) read off the contrast and its standard error.
- **Clever covariate.** The specific function of the propensity score that TMLE uses in step 3. You do not have to compute it by hand; `run_clean_tmle()` does it.
- **EIC (efficient influence curve) variance.** The standard-error formula TMLE uses. Interpretable as a regular Wald-type SE under standard regularity conditions.

Two more terms come up in the second half of the vignette: **plasmode simulation** (a simulation that keeps the real covariate and treatment structure but generates a synthetic outcome with a known truth) and **IPCW** (inverse-probability-of-censoring weighting, used to handle missing outcomes).

## The plain workflow: four estimators on one figure

Before introducing any governance machinery, we run the full estimator menu on the clinical anchor and compare them on a single figure. This is the workflow you can lift straight into a methods-development analysis.

```{r plain-workflow}
# 1. Lock the specification: data, variable roles, learner library, seed.
plain_lock <- create_analysis_lock(
  data       = dat,
  treatment  = "treatment",
  outcome    = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  sl_library = c("SL.glm", "SL.glmnet", "SL.mean"),
  seed       = 2026
)

# 2. Propensity score, then the support assessment: overlap verdict,
#    weight diagnostics, ESS, c-statistic, and the balance table.
ps  <- fit_ps(plain_lock)
sup <- assess_support(ps)
sup
```

```{r plain-overlap, fig.cap = "Mirrored propensity histogram by treatment arm, reference band shaded. Both arms should inhabit the same range; a one-sided pile-up signals positivity strain."}
plot(sup)
```

The support plot is the first thing a reviewer should see. If the two arms
do not cover the same propensity range, adjustment cannot recover a
marginal contrast for the full eligible cohort, and the estimand question
of the second half of this vignette is live. `love_plot(sup)` draws the
balance view of the same object.

```{r plain-estimators, results = "hide"}
# 3. Bind the analysis choice to the lock: the estimand plan plus the
#    TMLE specification the plasmode selector would normally pick.
primary_spec <- tmle_candidate(
  candidate_id = "plain_tmle",
  label        = "TMLE for the plain workflow",
  q_library    = c("SL.glm", "SL.glmnet"),
  g_library    = c("SL.glm", "SL.glmnet"),
  truncation   = 0.01
)
plain_lock <- declare_estimand_ladder(plain_lock, primary = "ATE",
                                      candidate = primary_spec)

# 4. Four estimators through the one front door, same lock, same PS.
crude <- estimate_effect(plain_lock, estimand = "ATE", estimator = "crude")
match <- estimate_effect(plain_lock, ps, estimand = "ATE",
                         estimator = "match")
iptw  <- estimate_effect(plain_lock, ps, estimand = "ATE",
                         estimator = "iptw")

# 5. TMLE, with the four modular steps returned for inspection: the
#    treatment mechanism, the outcome mechanism, the targeting step, and
#    the extracted estimate (the clever covariate lives in $steps).
tmle <- estimate_effect(plain_lock, ps, estimand = "ATE",
                        estimator = "tmle", missing = "complete_case",
                        return_steps = TRUE)
```

```{r plain-summary}
# 6. Combine the four estimates in one table.
#    Each result has $estimate; CI may be on the top level (run_*_workflow)
#    or under $estimates$ATE (modular TMLE extract).
pick_ci <- function(x) {
  if (!is.null(x$estimates) && !is.null(x$estimates$ATE)) {
    list(est = x$estimates$ATE$estimate,
         lo  = x$estimates$ATE$ci_lower,
         hi  = x$estimates$ATE$ci_upper)
  } else {
    se_or_na <- if (is.null(x$se) || length(x$se) == 0) NA_real_ else x$se
    lo_top   <- if (is.null(x$ci_lower)) x$estimate - 1.96 * se_or_na else x$ci_lower
    hi_top   <- if (is.null(x$ci_upper)) x$estimate + 1.96 * se_or_na else x$ci_upper
    list(est = x$estimate, lo = lo_top, hi = hi_top)
  }
}
plain_results <- data.frame(
  estimator = c("Crude (unadjusted)", "PS matching",
                "IPTW (stabilised)", "TMLE (full cohort)"),
  estimate  = c(pick_ci(crude)$est, pick_ci(match)$est,
                pick_ci(iptw)$est,  pick_ci(tmle)$est),
  ci_lower  = c(pick_ci(crude)$lo,  pick_ci(match)$lo,
                pick_ci(iptw)$lo,   pick_ci(tmle)$lo),
  ci_upper  = c(pick_ci(crude)$hi,  pick_ci(match)$hi,
                pick_ci(iptw)$hi,   pick_ci(tmle)$hi),
  stringsAsFactors = FALSE
)
knitr::kable(plain_results, digits = 4,
             caption = "Four estimators on the plain-mode lock.")
```

(`estimate_effect(missing = "ipcw")` adds an IPCW sensitivity path for studies with missing outcomes; `estimate_gcomprisk()` and `estimate_aipwrisk()` are also available, but they consume a `specify_models()` object rather than the `cleanroom_lock` shown here. See the function-reference vignette for both.)

```{r plain-forest, fig.cap = "Forest plot of the four estimators. The dashed line marks no effect; the adjusted estimators agree closely and shift the crude estimate.", fig.height = 3.5}
library(ggplot2)
ggplot(plain_results,
       aes(x = estimate, y = factor(estimator, levels = rev(plain_results$estimator)))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.2) +
  labs(x = "Risk difference (95% CI)", y = NULL,
       title = "Four estimators on one figure",
       subtitle = "Crude vs PS matching vs IPTW vs TMLE") +
  theme_minimal(base_size = 11)
```

A few things to read from the forest plot:

- The three adjusted estimators (PS matching, IPTW, TMLE) agree closely. When the design diagnostics look reasonable, this is exactly what you should see; substantial disagreement is a signal to slow down.
- The crude (unadjusted) estimate sits noticeably further from the others. The shift between crude and the adjusted estimators is the contribution of the baseline confounders.
- The confidence intervals are not the same width. IPTW's IF-based SE assumes the propensity score is known; in practice it has been estimated, so the IPTW SE is conservative. TMLE's SE accounts for the estimation of the propensity score.
- The matched estimator's CI is wider than the full-cohort TMLE's CI because matching restricts the analysis to the overlap subset and discards unmatched controls.

That is the entire core workflow. Everything in the rest of this vignette is *governance and audit machinery* on top of these four estimators, for studies that need a reproducible pre-outcome dossier, and is not a different estimator.

## Deciding the estimand before outcomes

A traditional clean room blinds the analyst: the outcome team does not see
the treatment-outcome association until the design is fixed. What it cannot
do is tell the design team whether the design supports the estimand they
plan to estimate. That failure is quiet. On a cohort with a severe practical
positivity violation, the average treatment effect can come back with a
narrow confidence interval and be wrong by an order of magnitude, because
truncation limits variance but not extrapolation, and nothing in the
estimator's own output distinguishes such an estimate from a real finding.
In the Rescue.Co case study that motivates this package, the protocol
contrast produced a time-to-arrival ATE of +14,637 minutes against a crude
difference of seven hours the other way, with a confidence interval
excluding zero.

The design stage below is what the package adds over blinding alone, and
none of it touches the outcome. The example is a cohort where one covariate
nearly separates the arms.

```{r estimand-decision, message=FALSE}
set.seed(11)
n  <- 800
x1 <- rnorm(n); x2 <- rbinom(n, 1, 0.4)
A  <- rbinom(n, 1, plogis(-2.2 + 3.2 * x1 + 0.5 * x2))
y  <- rbinom(n, 1, plogis(-2 + 0.5 * x1 + 0.4 * A))
sev <- data.frame(x1 = x1, x2 = x2, treatment = A, outcome = y)

lock_s <- create_analysis_lock(sev, "treatment", "outcome", c("x1", "x2"),
                             sl_library = "SL.glm", seed = 11L)
lock_s <- declare_estimand_ladder(lock_s, primary = "ATE",
                                  fallbacks = c("trimmed_ATE", "ATT", "ATO"),
                                  trigger = "SEVERE")
ps_s  <- fit_ps(lock_s, method = "glm")
sup_s <- assess_support(ps_s)
sup_s
```

```{r estimand-decision-plot, fig.height=3.8}
plot(sup_s)
```

The graded verdict (PASS, FLAG, SEVERE, FAIL, following the review-team
decisions of Muntner et al. 2024) is computed from the share of the sample
outside a prespecified propensity band and the largest inverse-probability
weight, escalated when any covariate stratum is near-deterministic in
treatment. The feasibility table then reads the same fit per estimand: the
effect among the treated needs control weights bounded by
`max g / (1 - g)`, and the overlap-weighted ATO's weights are bounded by
one, so both remain estimable where the ATE is not.

```{r estimand-decision-fea}
estimand_feasibility(ps_s)
```

`estimand_feasibility(ps_s, profile_vars = ...)` also profiles the patients
outside the band (its `$unsupported` table), which is what trimming would
remove; trimming changes the estimand and the profile is what makes the
change concrete. `simulate_support()` asks the
dynamic version of the question, generating synthetic outcomes over a
prespecified family of surfaces and reporting, per estimand, where its
estimator stays within a bias tolerance at near-nominal coverage. Estimation
then runs the declared ladder, labels every row with its estimand and
verdict, refuses an infeasible primary unless an override reason is
recorded, and logs every switch:

```{r estimand-decision-ladder, message=FALSE, warning=FALSE}
run_estimand_ladder(lock_s, ps_s, sl_library = "SL.glm", verbose = FALSE)
```

That is the package's answer to the question a clean room alone cannot
answer: not only was the analyst blinded, but the comparison the study
promised was checked for estimability before anyone saw an outcome, and the
estimand that was finally estimated is the one the data can support, chosen
by a rule that was declared in advance.

## When to read the rest

Continue reading if any of the following apply:

- You are running an analysis where the comparative answer will be reviewed by a third party, and you need a reproducible record of every analytic decision made before the comparative answer was seen.
- You want to use the **plasmode candidate-selector** to pick between competing TMLE specifications before unblinding.
- You want to use the **data-quality stress test** to check how the chosen estimator behaves under prespecified threats (measurement error, missingness, unmeasured confounding) before unblinding.
- You are working inside an institutional clean-room arrangement and want the software-enforced two-pass contract (`create_analysis_lock(enforce = TRUE)`), whose locks require a recorded design review before estimation runs.

If none of these apply, you can stop here and use the workflows above. cleanTMLE was designed to be useful at both levels of formality.

---

## Purpose of this vignette

This vignette demonstrates a staged workflow for an observational
comparative-effectiveness analysis on a simulated dataset. Design
and estimator-selection decisions are made before the observed
treatment-outcome association is examined; the goal is reproducible
prespecification and a structured documentation of analytic
decisions that an external reviewer can follow.

In this manuscript and vignette, *outcome-blind* means that
decisions about cohort construction, candidate estimators,
simulation scenarios, decision thresholds, and primary-estimator
selection are made without inspecting the observed
treatment-outcome association. Synthetic outcomes generated for
plasmode simulations may be used; the marginal mean of the
observed outcome on covariates is also used internally to fit the
plasmode generator. The *pre-outcome decision point* is the
juncture at which the prespecified diagnostics are reviewed and the
analysis is classified GO, FLAG (proceed only with documented
review), or STOP before the observed outcome is unmasked for the
comparative analysis.

cleanTMLE is not a substitute for target-trial design, source-data
validation, phenotype validation, independent oversight, or
subject-matter assessment of identification assumptions. The
worked example below shows reproducible prespecification, not a
regulatory submission template.

## Conceptual workflow

The vignette proceeds in the following conceptual steps:

1. **Stage 0 (precondition)**: feasibility assessment, target-trial
   specification, and protocol / SAP drafting. This is largely
   external work conducted before any cleanTMLE code is run.
2. **Stage 1a**: lock the estimand, covariates, learners, candidate
   grid, and decision thresholds.
3. **Stage 1b**: assess cohort adequacy and marginal event support.
4. **Stage 2a**: propensity score, balance, overlap, ESS, and
   design diagnostics.
5. **Stage 2b**: baseline plasmode candidate selection.
6. **Stage 2c**: plasmode data-quality (DQ) stress testing as a
   quantitative pre-outcome supplement to the fit-for-purpose review.
7. **Stage 3 (optional)**: negative-control outcome checks.
8. **Pre-outcome decision**: GO / FLAG / STOP.
9. **Stage 4**: authorized primary outcome analysis.
10. Post-outcome sensitivity analyses, interpretation, and archived
    analysis record.

The Muntner et al. (2024) clean-room governance framework
(role separation, restricted access, protocols, audit) is the
source of the staging idea and is the broader notion of a clean
room. cleanTMLE is the *software layer* of that workflow: it
records the analysis lock, runs the design diagnostics and the
plasmode evaluation, executes the DQ stress test, applies the
pre-outcome decision rule, and writes the audit trail. It does
not by itself implement clean-room personnel governance. The canonical stage labels used
throughout this vignette are:

```
Stage 0    Feasibility, target-trial specification, protocol / SAP drafting
Stage 1a   Lock estimand, covariates, learners, candidate grid, thresholds
Stage 1b   Cohort adequacy and marginal event support
Stage 2a   PS, balance, overlap, ESS, design diagnostics
Stage 2b   Baseline plasmode candidate selection
Stage 2c   DQ stress testing
Stage 3    Optional negative-control outcome checks
Pre-outcome decision: GO / FLAG / STOP
Stage 4    Authorized primary outcome analysis
Post-outcome: sensitivity analyses and interpretation
```

We use *proceed*, *proceed-with-qualification*, and *do-not-run*
in prose; the software shorthand for the same labels is GO, FLAG,
and STOP.

::: {.callout-warning}
This vignette demonstrates the package workflow. It does not
establish exchangeability, measurement validity, phenotype
validity, linkage accuracy, or regulatory acceptability.
:::

::: {.callout-warning}
Low-replicate examples in this vignette are for workflow demonstration only. Use `n_reps >= 200` before interpreting bias, coverage, or RMSE inferentially.
:::

## Pre-outcome study dossier

The reviewer-facing artefact produced by this workflow is a pre-outcome
study dossier. It is assembled across the stages demonstrated below,
and the review team reads it before authorising the primary analysis.
Its components are:

- Protocol and target-trial timing (the lock record, plus
  `design_report(protocol = ...)` for a TARGET-style emulation table);
  see *Define the causal question and estimand*.
- Cohort flow and attrition (`attrition_table()`); see *Stage 1b*.
- Event-process classification (structured event-process table and coherence check); see *From target trial specification to event-process classification*.
- Covariate and missingness summary (`make_table1()` and `sanitize_covariates()`); see *Stage 1a* covariate sanitisation.
- PS overlap, support verdict, and balance (`fit_ps()`, `assess_support()`, `love_plot()`); see *Stage 2a*.
- Estimand feasibility and the declared ladder (`estimand_feasibility()`, `declare_estimand_ladder()`); see *Deciding the estimand before outcomes* above.
- Treatment- and IPCW-weight diagnostics (`clean_weight_diagnostics()`); see *Outcome-blind diagnostics before fitting the final estimator*.
- Event-count and precision adequacy (`estimate_design_precision()`); see *Stage 1b*.
- Outcome-blind support map (`simulate_support()`); see *Deciding the estimand before outcomes*.
- Baseline plasmode candidate selection (`run_plasmode_feasibility()`, `select_tmle_candidate()`); see *Stage 2b*.
- DQ stress-test results (`run_plasmode_dq_stress()` and its `summary()`); see *Stage 2c*.
- Negative controls across cohort restrictions (`run_negative_control_ladder()`); see *Stage 3*.
- Care-process collider check (`check_process_indicators()`); see *Stage 3*.
- The assembled dossier and recommendation (`design_report()`); see *Pre-outcome decision*.

The dossier is the artefact the reviewer reads. Comparative
treatment-outcome estimates are not part of the dossier and are
produced only after the review.

### Why TMLE fits this workflow

TMLE is well suited to an outcome-blind staged workflow for a
small number of inter-related reasons:

- The estimand is declared in advance and the estimator is
  defined as the targeted update of an initial outcome model
  toward that estimand, so the analysis target is separable from
  the nuisance fits.
- The nuisance library, the propensity-score truncation, and the
  targeting step can all be prespecified and recorded in the
  analysis lock.
- Targeting decouples the bias of the final estimator from the
  quality of any single nuisance fit, allowing the use of
  adaptive learners without driving the choice from the observed
  outcome association.
- The candidate grid is enumerable: different libraries and
  truncation rules define distinct candidates that can be
  screened on synthetic outcomes before authorisation.
- Inference is based on the efficient influence function, which
  provides a transparent variance estimate compatible with the
  diagnostics computed at earlier stages.

### What each diagnostic answers

| Diagnostic | Question answered | What it does not answer |
|---|---|---|
| PS / overlap / balance / ESS | Empirical design feasibility | Operating characteristics of any specific TMLE candidate |
| Baseline plasmode | Estimator behaviour under a locked synthetic outcome model | Behaviour under the realised outcome process |
| DQ stress testing | Stability under prespecified fit-for-purpose threats | True bias magnitude under unobserved threats |
| Negative-control outcomes | Residual bias through a different channel | Primary-estimator behaviour under stress |
| Post-outcome sensitivity | Robustness to deviations after unblinding | Pre-outcome decisions |

### Stage 0 as precondition

Stage 0 covers the feasibility assessment, target-trial
specification, and protocol / SAP drafting that precede any
analytic code. These activities are typically conducted outside
the software layer, often involving subject-matter experts, data
custodians, and study sponsors. cleanTMLE assumes Stage 0 is
complete: the causal question, eligible population, treatment
strategies, follow-up window, and primary estimand are already
specified in a written protocol before Stage 1a begins. The
remainder of this vignette demonstrates Stages 1a through 4.

---

## Define the causal question and estimand

Before writing any analysis code we state the causal question in
target-trial-emulation style.

| Element | Specification |
|---------|---------------|
| **Population** | Adults with simulated disease (the `sim_func1()` cohort) |
| **Treatment strategies** | Treatment (A = 1) versus control (A = 0), assigned at baseline |
| **Follow-up** | 24 months from cohort entry |
| **Outcome** | Primary event occurring by 24 months (`event_24`) |
| **Causal contrast** | Average treatment effect on the risk-difference scale |

The causal estimand is $\psi = E[Y^{a=1}] - E[Y^{a=0}]$, the
population-level risk difference under two hypothetical interventions.
Because we work with observational data, the quantity we can identify
from the observed distribution is the *statistical estimand*

$$
\psi = E_W\!\bigl[E[Y \mid A=1, W] - E[Y \mid A=0, W]\bigr],
$$

which equals the causal parameter under the standard identification
assumptions (consistency, exchangeability, positivity).
The package estimates this statistical estimand; causal interpretation
relies on the investigator's subject-matter defence of those assumptions.

We choose the risk difference as the primary estimand because it has a
transparent absolute-scale interpretation, admits doubly-robust and
semiparametric-efficient estimation via TMLE, and is directly relevant
to clinical decision making.

---

## From target trial specification to event-process classification

Once the target trial and the statistical estimand are stated,
post-baseline events should not be assigned a generic role of
"censoring." Each event should be mapped to the causal question,
to the corresponding ICH E9(R1) intercurrent-event strategy where
applicable, and to the operational role it plays in estimation.
For example, the primary event is modelled directly, death from
other causes is a competing event, loss to follow-up is a
missing-outcome process handled by IPCW, and treatment switching is
an intercurrent event kept under the assigned treatment strategy.
This classification, the target population and reference group, and
the missing-data plan are prespecified and recorded with the
analysis before the outcome is unmasked.

---

## Stage 1a: Record the analysis specification (lock)

The analysis specification records the treatment, outcome,
covariates, candidate estimators, learner library, simulation
settings, and random seed. Later steps refer back to this object
so that the final analysis can be compared with the prespecified
plan, and so that an external reviewer can reconcile the recorded
inputs with the analyses that were run.

```{r stage1a}
library(cleanTMLE)
set.seed(42)

dat <- sim_func1(n = 400, seed = 42)
```

### Covariate sanitisation

Before locking the specification, we sanitise covariates to ensure they
are safe for SuperLearner: impute missing values, drop zero-variance
columns, and fix formula-unsafe names.

```{r sanitize}
clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  verbose = TRUE)
dat <- clean$data
covs <- clean$covariates
```

```{r lock}
lock <- create_analysis_lock(
  data          = dat,
  treatment     = "treatment",
  outcome       = "event_24",
  covariates    = covs,
  sl_library    = c("SL.glm", "SL.mean"),
  negative_controls = "nc_outcome",
  plasmode_reps = 40L,
  seed          = 42L
)
print(lock)
```

::: {.callout-tip}
## Optional: outcome masking

For analyses that require stricter blinding, `create_analysis_lock(mask
= TRUE)` (or `mask_outcome()` on an existing lock) replaces the real
outcome column with `NA` values inside the lock, so no code can read
the outcome before estimation; `unmask_outcome(lock, original_lock)`
restores it when the design review is complete. Stronger still is
keeping the outcome columns in a separate store and joining them only
at estimation, which is what the package's case study does. In this
vignette we proceed without masking for clarity.
:::

The estimand plan is declared next, and it is the key
pre-registration: the primary estimand, the ordered fallbacks the
analysis will accept, and the support verdict that triggers a move
down the ladder. Switching estimands after seeing results is a garden
of forking paths; switching by this rule, declared before outcome
access, is a design decision the log records.

```{r ladder-decl}
lock <- declare_estimand_ladder(lock,
  primary   = "ATE",
  fallbacks = c("trimmed_ATE", "ATT", "ATO"),
  trigger   = "SEVERE"
)
```

Decision thresholds for the stress test (the maximum acceptable bias,
the coverage floor) are arguments of the functions that use them, and
every declaration lands in the lock's design log, so the rule is on
the record before any diagnostic is scored.

### File-based pipeline support

In production pipelines each stage runs as a separate script.
`save_lock()` / `load_lock()` handle serialisation with automatic
hash re-validation, eliminating the boilerplate of manual RDS
save/load and integrity checks.

```{r save-lock}
tmp_lock_path <- tempfile("lock_", fileext = ".rds")
save_lock(lock, tmp_lock_path)
lock_reloaded <- load_lock(tmp_lock_path)  # re-validates the hash
cat("Lock hash match:",
    identical(lock$lock_hash, lock_reloaded$lock_hash), "\n")
unlink(tmp_lock_path)
```

In a real multi-stage pipeline each stage script would save to a
shared results directory (e.g., `save_lock(lock, "results/stage1_lock.rds")`)
and the next stage script would `load_lock("results/stage1_lock.rds")`
at its top. The design log rides inside the lock, so saving the lock
archives the decision record with it.

---

## Stage 1b: Cohort adequacy and marginal event support

Before any modelling, we confirm that the cohort has adequate size,
event counts, and precision to support the planned analysis. If it
does not, the analysis stops here, before any comparative outcome
modelling begins. `estimate_design_precision()` reads only marginal
event counts (never the treatment-outcome association): it reports the
minimum detectable difference (MDD) given the locked sample and event
counts, compares it against a target when one is supplied, and its
`$event_support` element tabulates events by arm so sparse cells are
visible.

```{r design-precision}
dp <- estimate_design_precision(lock, target_mdd = 0.05)
print(dp)
```

::: {.callout-note}
The MDD report may show "Feasible: NO" with this small simulated
sample (n = `r nrow(dat)`).
This is expected, because the vignette uses a deliberately small dataset
for fast rendering.
In practice you would either increase sample size or relax the
target MDD.
The MDD check is **informational** and does not stop the pipeline;
it informs the review.
:::

---

## Stage 2a: Propensity fit, support verdict, feasibility, balance

Stage 2a fits the propensity score and grades the design: overlap,
implied weights, effective sample size, balance, and which estimands
the fit can support. The outcome is never accessed at this stage; only
covariates (W) and treatment (A) are used.

### Propensity score estimation

`fit_ps()` is the one front door: `method = "superlearner"` (the
default), `"glm"` for a logistic fit, or `"external"` with scores
estimated outside the package.

```{r ps}
ps_fit <- fit_ps(lock, method = "glm", truncate = 0.01)
print(ps_fit)
```

::: {.callout-tip}
## Bring-your-own propensity scores

Scores fitted outside cleanTMLE (a parallel SuperLearner cluster,
a Bayesian model, another language) enter through the same front
door and work identically downstream:

```r
ps_fit <- fit_ps(lock, method = "external", scores = my_ps)
```
:::

### The support assessment

`assess_support()` grades the overlap: the share of the sample with
untruncated propensity outside the band, the largest implied weight,
ESS per arm, the c-statistic (discrimination is bad news here), a
stratum scan and a shallow-tree search for pockets where treatment is
nearly deterministic, and a balance table. The verdict (PASS, FLAG,
SEVERE, FAIL) carries a caveat sentence that travels with every
downstream estimate.

```{r support}
sup <- assess_support(ps_fit)
print(sup)
```

```{r ps-overlap, fig.cap = "Mirrored propensity histograms: treated above the axis, controls below, band shaded."}
plot(sup)
```

```{r love, fig.cap = "Standardised mean differences before and after weighting."}
love_plot(sup)
```

### Per-estimand feasibility

The same fitted propensity is then read per estimand:
the weights each estimand implies, their maximum, ESS by arm, who is
effectively removed, and a verdict on the same scale. This is the
table that tells you whether the declared primary is answerable and
which fallbacks are.

```{r feasibility}
fea <- estimand_feasibility(ps_fit,
  profile_vars = c("age", "biomarker"))
print(fea)
```

On this well-overlapped simulated design every estimand passes. The
*Deciding the estimand before outcomes* section at the top of this
vignette shows the same table on a severe-overlap design, where it
does the work it was built for.

---

## Stage 2b: Outcome-blind TMLE specification selection

Before accessing the real outcome, we compare *prespecified TMLE
specifications* on plasmode-simulated data.
Each candidate varies the propensity-score truncation threshold and/or
the SuperLearner library used for nuisance estimation.
Plasmode simulation preserves the real covariate distribution and
treatment mechanism but generates synthetic outcomes with a known true
effect, allowing us to evaluate bias, RMSE, and coverage for each
candidate specification.

The purpose is *prespecified specification selection*, not performance
chasing: the selection rule (here, minimum RMSE) is declared in
advance, and the selected TMLE specification is locked into the
analysis lock and carried forward exactly to Stage 4 without
ad hoc modification.

Note that the PS truncation level is not only a numerical
stabilisation parameter: different truncation thresholds imply
different effective target populations. Two candidates with
different truncation rules are therefore paired estimator-and-
estimand objects, and the truncation rule is recorded in the
analysis lock alongside the learner library so the
estimand-vs-estimator distinction remains auditable.

#### Estimator selection versus estimand feasibility

A useful mental check before locking the candidate grid:

- If candidates differ only by nuisance learner, screening step,
  or seed, candidate selection is *estimator* selection. The
  causal estimand is the same across candidates.
- If candidates differ by PS truncation, overlap restriction, or
  eligibility filter, they may target *different* effective
  populations. Selection then has an *estimand* component as well.

In the second case, the prespecified rationale should record the
estimand implication for each candidate before any outcome access.
If the staged workflow shows that only candidates with severe
truncation meet the operating-characteristic thresholds, the
practical conclusion is that the original target population may
not be supported by the data and the estimand itself needs
revision, not just the nuisance fit.

### Define TMLE candidates

```{r tmle-candidates}
# Define candidate TMLE specifications that vary truncation
candidates <- list(
  tmle_candidate("glm_t01", "GLM, trunc=0.01",
                 g_library = c("SL.glm"), truncation = 0.01),
  tmle_candidate("glm_t05", "GLM, trunc=0.05",
                 g_library = c("SL.glm"), truncation = 0.05),
  tmle_candidate("glm_mean_t01", "GLM+Mean, trunc=0.01",
                 g_library = c("SL.glm", "SL.mean"), truncation = 0.01)
)
```

### Plasmode feasibility evaluation

```{r plasmode, cache = TRUE}
plas <- run_plasmode_feasibility(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05, 0.10),
  reps            = 40L,
  verbose         = FALSE
)
print(plas)
```

### Select and lock the primary TMLE specification

```{r select}
selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock <- declare_estimand_ladder(lock,
  primary   = "ATE",
  fallbacks = c("trimmed_ATE", "ATT", "ATO"),
  trigger   = "SEVERE",
  candidate = selected
)
print(selected)
```

The selected TMLE specification for estimation is
**`r selected$label`** (truncation = `r selected$truncation`).
Re-declaring the ladder with `candidate = selected` attaches the
specification to the ladder, so downstream estimation honours its
truncation and libraries automatically, and the re-declaration is a
new design-log entry rather than a silent overwrite.

## Stage 2c: Plasmode data-quality stress test

The DQ stress test is a **quantitative pre-outcome supplement to
the fit-for-purpose review**, not a formal quantitative bias
analysis (QBA) and not a substitute for source-data validation or
post-outcome QBA. The intended sequence is:

1. Fit-for-purpose review identifies plausible threats.
2. Threats are translated into quantitative severity ranges.
3. Severity ranges are locked before primary outcome access.
4. Synthetic outcomes are generated and perturbed.
5. Candidates are evaluated using the same metrics as baseline
   plasmode selection.
6. The same GO / FLAG / STOP rule is applied.

The DQ stress test documents how candidate estimators behave under
the locked threat range; it does not validate the data or replace
post-outcome QBA.

Four mechanisms are supported:

1. *Covariate missingness*: a fraction of values is set to `NA` and
   imputed with the column median.
2. *Treatment misclassification*: an asymmetric pair of (sensitivity,
   specificity) is applied to A.
3. *Outcome misclassification*: an asymmetric pair of (sensitivity,
   specificity) is applied to Y, optionally per treatment arm.
4. *Unmeasured confounding*: a latent binary U is sampled and shifts
   both the treatment mechanism (via a real PS model fitted on the
   lock data) and the outcome probability by user-specified odds
   ratios.

```{r dq-stress, cache = TRUE}
dq_results <- run_plasmode_dq_stress(
  lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.05),
  # 60 (rather than 40) replicates stabilise the MCAR/measurement-error cells,
  # whose worst-case |bias| is otherwise driven up by Monte Carlo noise close to
  # the 0.02 threshold; at 60 reps the margin below the STOP threshold is
  # comfortable so the illustrative gate reliably returns FLAG, not STOP.
  reps            = 60L,
  data_quality_scenarios = list(
    # Covariate missingness under three mechanisms, each median-imputed.
    # MCAR is unbiased by construction; MAR (treatment-dependent) and MNAR
    # (value-dependent) make the imputation biased and are stronger tests.
    covariate_missingness = list(            # MCAR
      fractions = c(0.05, 0.10, 0.20)
    ),
    covariate_missingness_mar = list(        # MAR (treatment-dependent)
      fractions = c(0.10, 0.20), treatment_OR = 3
    ),
    covariate_missingness_mnar = list(       # MNAR (value-dependent)
      fractions = c(0.10, 0.20), strength = 1.5
    ),
    treatment_misclass = list(
      sensitivity = c(0.99, 0.95, 0.90),
      specificity = c(0.99, 0.95, 0.90)
    ),
    outcome_misclass = list(
      sensitivity = c(0.95, 0.90),
      specificity = c(0.99, 0.95)
    ),
    unmeasured_confounding = list(
      U_prevalence   = 0.20,
      U_treatment_OR = c(1.5, 2.0),
      U_outcome_OR   = c(1.5, 2.0)
    )
  ),
  verbose = FALSE
)

deg <- summary(dq_results)
head(deg)
```

The *degradation gradient* is the curve of estimator performance
against the severity of a prespecified data-quality threat. It is
analogous to a dose-response curve for estimator fragility. A
shallow gradient suggests a candidate is stable across plausible
severities. A cliff suggests that small increases in degradation
move performance outside the locked criteria.

`summary()` on the stress-test result returns a long table comparing
each degraded run with the baseline (no degradation): bias change,
RMSE ratio, coverage drop. A robust candidate produces a flat curve;
a fragile candidate shows a cliff. The table can be filtered to the
candidate locked in by `select_tmle_candidate()` and reviewed
alongside the SPIFD2 fit-for-purpose narrative in the design report.
When a candidate breaches the declared bias or coverage envelope
under a plausible severity, the reading is the Muntner "conduct
additional analyses" or "terminate" recommendation for that
specification, and the review is recorded in the design log. When
every candidate breaches, the honest conclusion is usually about the
data source or the estimand, not the nuisance fit.

The DQ stress test is one of the package's two distinguishing
contributions relative to prior outcome-blind plasmode work (the
other is the support map): standard plasmode selects the candidate
with the best baseline RMSE, while `run_plasmode_dq_stress()` lets
the analyst select the candidate that stays inside the declared
envelope across the plausible range of every prespecified threat,
via `select_tmle_candidate(rule = "min_max_rmse")`.

---

## Stage 3 (optional): Negative controls across cohort restrictions

A negative-control outcome (NCO) screen is an optional pre-outcome
check that sits between the Stage 2c DQ stress test and the
pre-outcome decision. It assesses residual confounding using a
negative-control outcome, a variable that is known (or strongly
believed) to be unaffected by treatment. If the same analytic
pipeline detects an association with the negative control, this
suggests that confounders remain unbalanced after adjustment.

The negative control `nc_outcome` (registered on the lock at Stage
1a) was generated from covariates only, with no treatment effect, so
any estimated association is attributable to residual confounding.

`run_negative_control_ladder()` fits every registered control on
every nested cohort restriction, so the review sees not only whether
a control fails but which restriction makes it pass. A control that
fails on the full cohort and turns null after a pre-registered
exclusion localises the confounding to the excluded stratum, which a
single-cohort check cannot show. Controls too rare to fit are
labelled inestimable rather than dropped silently.

```{r stage3}
nc <- run_negative_control_ladder(lock,
  restrictions = list(
    "comorbidity-free" = lock$data$comorbidity == 0
  ),
  method = "tmle", ps_method = "glm", verbose = FALSE)
print(nc)
```

On this simulated design the control is null on both rungs, as it
should be: the covariates that drive `nc_outcome` are in the
adjustment set. A companion check for care-process variables,
`check_process_indicators()`, regresses each named indicator on
treatment given the remaining covariates and flags indicators the
treatment predicts; such variables belong outside the adjustment set,
where conditioning on them would open collider paths.

---

## Pre-outcome decision

The pre-outcome decision summarises whether the prespecified
diagnostics support proceeding to outcome analysis, require
additional review, or indicate that the study should stop or be
redesigned; these are the three review-team recommendations of
Muntner et al. (2024), and the design report states which one the
evidence supports. `design_report()` assembles the dossier the review
team reads: the lock record, the support verdict, the feasibility
table, the declared ladder, the support map when one was run, and the
negative-control ladder, closed by a one-line recommendation.

The support map is the outcome-blind simulation described in
*Deciding the estimand before outcomes* at the top of this vignette:
`simulate_support()` resamples the covariates, redraws treatment from
the fitted propensity (the generate-treatment plasmode design), and
generates synthetic outcomes over a prespecified surface family, so
each estimand's estimator is scored where the truth is known. A small
run makes the report complete here; use far more replicates in
practice.

```{r support-sim, cache = TRUE}
sim <- simulate_support(lock, ps_fit, reps = 10, verbose = FALSE)
```

```{r design-report}
design_report(lock, sup, fea, simulation = sim, nc_ladder = nc)
```

On this simulated design the recommendation is to proceed: the
support verdict is PASS, the declared primary is feasible, and the
negative control is null on every rung. The report reads differently
on the severe-overlap design shown at the top of this vignette,
where it recommends the ladder switch and names the fallback
estimands.

### What if the design fails the review?

Two refusals are built into estimation, and both are logged rather
than silent. First, an estimand whose feasibility verdict meets the
declared trigger is refused by `run_estimand_ladder()` unless an
`override_reason` is supplied, and a supplied override becomes a
design-log entry. Second, every estimate passes an implausibility
guard against bounds the raw data impose (sign against a much larger
crude difference, magnitude several times the crude contrast, implied
arm risks outside the observed range), and a flagged row prints its
reason. The refusal path is shown on the severe-overlap design at the
top of this vignette; the guard column appears in every ladder result
below.

---

## Outcome-blind diagnostics before fitting the final estimator

Before the outcome is unmasked, we report a compact set of
diagnostics that support assessment of identification and
estimator stability: treatment overlap, censoring weights (when
present), missingness weights (when applicable), weighted SMDs,
and effective sample size. These diagnostics should be conducted
and the prespecified failure thresholds applied before looking at
outcomes when the workflow allows.

```{r outcome-blind-diagnostics, eval = exists("ps_fit")}
A <- lock$data[[lock$treatment]]
ps <- ps_fit$ps
w  <- ifelse(A == 1, 1 / ps, 1 / (1 - ps))

W_for_smd <- lock$data[, lock$covariates, drop = FALSE]
weight_diag <- clean_weight_diagnostics(weights = w, treatment = A,
                                         covariates = W_for_smd,
                                         max_weight_threshold = 10,
                                         ess_floor = 0.4 * length(w))
print(weight_diag)
```

The maximum-weight threshold (10) and the ESS floor (40 % of N)
are conservative defaults; the prespecified analysis plan should
record values that are appropriate for the study's anticipated
overlap, weight distribution, and sample size. When the study
uses censoring or missingness weights, `clean_weight_diagnostics()`
can be called separately on each weight vector.

---

## Stage 4: Final comparative analysis

Only after the design review is complete do we access the real
outcome and run the locked analysis. The declared route is the
ladder itself: `run_estimand_ladder()` estimates the primary when it
is feasible plus every feasible fallback, labels each row with its
estimand and support verdict, and logs any switch or override.

```{r ladder-run}
fit <- run_estimand_ladder(lock, ps_fit)
print(fit)
```

On this well-supported design the ladder reduces to the declared
primary plus labelled alternatives; the interesting behaviour, the
refused ATE and the logged switch, appears on the severe-overlap
design at the top of this vignette.

### One estimand through the front door

`estimate_effect()` estimates a single declared estimand, with the
estimator family and the missing-outcome handling as arguments.
`return_steps = TRUE` exposes the modular TMLE quartet for
inspection: the treatment mechanism (fitted at Stage 2a without
outcome access, with the locked truncation applied), the outcome
mechanism (the first access to the real outcome), the targeting step
that fluctuates the initial fit along the clever covariate
$H(A,W) = A/g(W) - (1-A)/(1-g(W))$, and the extracted estimate with
its influence-function SE.

```{r estimate-effect}
tmle_fit <- estimate_effect(lock, ps_fit,
  estimand = "ATE", estimator = "tmle", return_steps = TRUE)
print(tmle_fit)
names(tmle_fit$steps)
```

### Comparators and sensitivity estimators through the same door

The estimator families that a conventional analysis reports as
context are arguments, not separate pipelines. Matching runs the ATT
on the matched cohort; IPCW targets the full-cohort estimand under a
missingness-at-random model rather than the complete-case subset.

```{r comparators}
crude_fit <- estimate_effect(lock, estimand = "ATE",
                             estimator = "crude")
iptw_fit  <- estimate_effect(lock, ps_fit, estimand = "ATE",
                             estimator = "iptw")
match_fit <- estimate_effect(lock, ps_fit, estimand = "matched_ATT")
ipcw_fit  <- estimate_effect(lock, ps_fit, estimand = "ATE",
                             estimator = "tmle", missing = "ipcw")
```

```{r compare, fig.cap = "Cross-estimator comparison. TMLE is the locked primary; the others are labelled comparators, not interchangeable estimates of one parameter."}
forest_plot(list(
  `Crude`           = crude_fit,
  `IPTW`            = iptw_fit,
  `Matched ATT`     = match_fit,
  `TMLE (primary)`  = tmle_fit,
  `TMLE + IPCW`     = ipcw_fit
))
```

TMLE is the locked primary analysis; matching and IPTW serve as
contextual comparators. Agreement across estimators increases
confidence that the result is not an artefact of a single modelling
strategy, and disagreement between the matched and full-cohort rows
is itself information about where the effect lives.

### TMLE diagnostics

`plot()` on a TMLE fit exposes the estimator's own diagnostics: the
influence-curve distribution (whose heavy tails warn about unstable
inference) and the clever covariate (whose extremes are the
positivity strain made visible).

```{r ic-hist, fig.cap = "Efficient influence curve distribution."}
plot(tmle_fit, type = "ic")
```

```{r clever, fig.cap = "Clever covariate by treatment group."}
plot(tmle_fit, type = "clever")
```

---

## Reporting cumulative risks and contrasts

Treatment-specific risks at the prespecified follow-up time, the
risk difference, and its confidence interval are reported in a
single compact table. The arm risks are the model-based (TMLE
plug-in) risks taken directly from the fit; their difference equals
the reported risk difference.

```{r risk-report-vignette, eval = exists("tmle_fit")}
ate <- tmle_fit$estimates$ATE
# Model-based (TMLE plug-in) treatment-specific risks, taken directly from the
# fit rather than reconstructed from crude arm means; their difference equals
# the reported ATE.
r1 <- tmle_fit$risks$treated
r0 <- tmle_fit$risks$control

risk_report <- data.frame(
  treatment_strategy = c("Treated", "Control"),
  n_observed         = c(sum(lock$data[[lock$treatment]] == 1),
                         sum(lock$data[[lock$treatment]] == 0)),
  risk               = c(r1, r0),
  risk_difference    = c(NA, ate$estimate),
  ci_lower           = c(NA, ate$ci_lower),
  ci_upper           = c(NA, ate$ci_upper))
knitr::kable(risk_report, digits = 4,
             caption = "Treatment-specific risks at 24 months and the risk-difference contrast (modular TMLE).")
```

### Why hazard ratios are secondary in this workflow

cleanTMLE emphasises cumulative-risk contrasts because they map
directly to treatment-specific risks under specified strategies and
are generally more interpretable for regulatory RWE studies.
Hazard ratios may be useful as secondary summaries when
proportional hazards is plausible and the estimand is explicitly
hazard-based, but they should not replace prespecified
cumulative-risk contrasts when the causal question concerns
absolute or relative risk by a clinically meaningful follow-up
time.

---

## Sensitivity analysis

### Truncation sensitivity

The prespecified sensitivity plan varies the PS truncation threshold
to assess how much the weighted estimates depend on extreme weights.
Truncation is an estimand-relevant choice, not only a numerical one,
so each level is reported as its own labelled row rather than folded
into a single number. `resolve_truncation_rule()` documents the
default rule (the adaptive $5/(\sqrt{n}\,\log n)$ bound) and the
fixed alternatives; a compact grid comes from calling
`estimate_effect()` at each level.

```{r sensitivity-trunc}
trunc_grid <- c(0.01, 0.025, 0.05)
sens <- do.call(rbind, lapply(trunc_grid, function(tt) {
  ps_t <- fit_ps(lock, method = "glm", truncate = tt)
  f <- estimate_effect(lock, ps_t, estimand = "ATE",
                       estimator = "iptw")
  data.frame(truncation = tt, estimate = f$estimate,
             ci_lower = f$ci_lower, ci_upper = f$ci_upper)
}))
print(sens, row.names = FALSE)
```

### Additive-bias delta sensitivity

`run_delta_sensitivity()` shifts a point estimate by a prespecified
grid of additive biases and reports where the confidence interval
crosses the null, including the g-value, the smallest additive bias
that would move the interval across zero.

```{r delta-sens}
ate <- tmle_fit$estimates$ATE
delta <- run_delta_sensitivity(ate$estimate, ate$se)
head(delta)
```

### E-value

The E-value quantifies the minimum strength of unmeasured confounding
(on the risk-ratio scale) needed to explain away the observed
association.

```{r evalue}
ate <- tmle_fit$estimates$ATE
# Confounding-adjusted risk ratio from the model-based (TMLE plug-in) arm
# risks the estimator targeted, not the crude control risk. The CI bound
# maps the risk-difference lower limit onto the risk-ratio scale at the
# adjusted control risk.
r1_hat <- tmle_fit$risks$treated
r0_hat <- tmle_fit$risks$control
rr_hat <- r1_hat / r0_hat
rr_lo  <- (r0_hat + ate$ci_lower) / r0_hat
ev <- compute_evalue(rr_hat, ci_bound = min(rr_lo, 1/rr_lo))
cat(sprintf("E-value: %.2f (CI bound: %.2f)\n", ev["e_value"], ev["e_value_ci"]))
```

---

## The design log and reproducibility

The design log lives on the lock and accumulates one timestamped row
per recorded decision: the ladder declarations, any re-declaration
with a selected candidate, estimand switches with their triggering
verdict, and overrides with their reasons. Because it rides inside
the lock, `save_lock()` archives the decision record with the
specification it governed, and the estimation results carry their own
copy of the entries the run generated.

```{r design-log}
knitr::kable(lock$design_log,
             caption = "Design log for the staged analysis.")
```

An external reviewer reconciles three artefacts: the lock (whose hash
fingerprints the data shape, variable roles, and seed), the design
log (what was declared and decided, and when), and the labelled
estimation output (what ran, under which verdicts, with which flags).
The reviewer never needs the analyst's recollection of which choice
was made when.

---

## Summary

The worked example illustrates the canonical stages:

1. **Stage 0**: feasibility, target-trial specification, and
   protocol / SAP drafting (precondition, external to the package).
2. **Stage 1a**: lock the specification (`create_analysis_lock()`)
   and declare the estimand ladder
   (`declare_estimand_ladder()`).
3. **Stage 1b**: cohort adequacy, design precision, and marginal
   event support (`estimate_design_precision()`).
4. **Stage 2a**: propensity fit (`fit_ps()`), the graded support
   assessment (`assess_support()`), per-estimand feasibility
   (`estimand_feasibility()`), and balance (`love_plot()`).
5. **Stage 2b**: the outcome-blind support map
   (`simulate_support()`) and plasmode candidate selection
   (`run_plasmode_feasibility()`, `select_tmle_candidate()`).
6. **Stage 2c**: plasmode DQ stress testing
   (`run_plasmode_dq_stress()`) on the locked severity ranges from
   the fit-for-purpose review.
7. **Stage 3 (optional)**: negative controls across cohort
   restrictions (`run_negative_control_ladder()`) and the
   care-process collider check (`check_process_indicators()`).
8. **Pre-outcome decision**: the assembled dossier and
   recommendation (`design_report()`), in the vocabulary of Muntner
   et al. (2024): proceed, proceed with qualification, or stop.
9. **Stage 4**: estimation of the declared primary and feasible
   fallbacks (`run_estimand_ladder()`, `estimate_effect()`).
10. **Post-outcome**: sensitivity analyses (`compute_evalue()`,
    `run_delta_sensitivity()`), interpretation, and the archived
    record (the lock with its design log, and the labelled
    estimates).

### Integration with real-world pipelines

The package supports common production patterns:

- **External PS**: `fit_ps(method = "external", scores = ...)` wraps
  scores from parallel SuperLearner, Bayesian models, or external
  software into a compatible `ps_fit`.
- **File-based pipelines**: `save_lock()` / `load_lock()` handle RDS
  serialisation with automatic hash re-validation, and the design log
  travels inside the lock.
- **Multi-arm studies**: `create_contrast_locks()` builds one lock
  per prespecified pairwise contrast on a shared cohort, each with
  its own support assessment, ladder, and design report.
- **Institutional enforcement**: locks created with `enforce = TRUE`
  withhold estimation until a recorded design review, for settings
  that want the two-pass discipline enforced in software; outcome
  masking is available to every lock.

The package does not enforce institutional governance (role separation,
independent review boards, blinded data managers); those are
organisational responsibilities.
What it does provide is software scaffolding that makes the staged
workflow reproducible and auditable.

```{r session-info}
sessionInfo()
```
