---
title: "Recreating causalRisk-style workflows + TMLE extensions"
author: "causalRiskeR Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recreating causalRisk-style workflows + TMLE extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Installation

```{r install, eval = !requireNamespace("causalRiskeR", quietly = TRUE)}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("amertens/cleanTMLE", subdir = "causalRiskeR")
```

## Overview

This vignette demonstrates how to use **causalRiskeR** for causal inference
in time-to-event settings. We cover:

1. Data simulation
2. Model specification using the `specify_models()` + `identify_*()` DSL
3. IPW cumulative risk estimation (overall, treatment-stratified, IPTW+IPCW)
4. Competing risks
5. Weight diagnostics, propensity score overlap, trimming/truncation
6. G-computation and AIPW (doubly robust) estimators
7. Hazard ratio estimation with `estimate_ipwhr()`
8. TMLE extensions: point-treatment, survival, and longitudinal (LMTP)
9. Tables and plots

## 1. Simulate Data

```{r simulate-data}
library(causalRiskeR)

# Generate simulated dataset
dat <- sim_func1(n = 1000, seed = 42)
head(dat)
str(dat)

# Summary of key variables
table(dat$treatment)
table(dat$event)
table(dat$event_type)
summary(dat$time)
```

The simulated data has:
- **Confounders**: age, sex, biomarker, comorbidity
- **Treatment**: binary (0/1), confounded by age, sex, biomarker
- **Time-to-event**: with right censoring and competing risks
- True treatment effect: protective (HR ~ 0.7)

## 2. Model Specification

causalRiskeR uses a pipe-friendly specification DSL. First, create a
specification object, then progressively identify model components.

```{r spec-basic}
# Basic specification: outcome + treatment + censoring
spec <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph") |>
  identify_subject(id)

print(spec)
```

## 3. IPW Cumulative Risk Estimation

### 3.1 Overall Risk (no treatment contrast)

```{r ipw-overall}
spec_overall <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event")

fit_overall <- estimate_ipwrisk(spec_overall, risk_time = c(6, 12, 18, 24, 30, 36))
print(fit_overall)
plot(fit_overall)
```

### 3.2 Treatment-Stratified Risk (unadjusted)

```{r ipw-stratified}
spec_strat <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event") |>
  identify_treatment(treatment)  # no PS model = unweighted

fit_strat <- estimate_ipwrisk(spec_strat, risk_time = c(6, 12, 18, 24, 30, 36))
print(fit_strat)
plot(fit_strat)
```

### 3.3 IPTW + IPCW Adjusted Risk

```{r ipw-adjusted}
fit_ipw <- estimate_ipwrisk(
  spec,
  risk_time = c(6, 12, 18, 24, 30, 36),
  weight_type = "iptw",
  nboot = 100,
  seed = 42
)
print(fit_ipw)
plot(fit_ipw)
```

### 3.4 Risk Difference and Risk Ratio

```{r ipw-effects}
# Risk Difference
plot(fit_ipw, effect = "RD")

# Risk Ratio
plot(fit_ipw, effect = "RR")
```

### 3.5 SMR Weights (ATT targeting)

```{r ipw-smr}
fit_smr <- estimate_ipwrisk(
  spec,
  risk_time = c(12, 24),
  weight_type = "smr"
)
print(fit_smr)
```

## 4. Competing Risks

```{r competing-risks}
spec_cr <- specify_models(data = dat) |>
  identify_outcome(event, type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph") |>
  identify_competing_risk(event_type, event_value = 1)

fit_cr <- estimate_ipwrisk(spec_cr, risk_time = c(6, 12, 18, 24, 30, 36))
print(fit_cr)
plot(fit_cr)
```

## 5. Weight Diagnostics

### 5.1 Weight Summary Table

```{r wt-summary}
make_wt_summary_table(fit_ipw)
```

### 5.2 Extreme Weights

```{r extreme-wt}
extreme_weights(fit_ipw, k = 5)
```

### 5.3 Propensity Score Overlap

```{r ps-hist}
hist(fit_ipw, type = "ps")
```

### 5.4 Weight Distribution

```{r wt-hist}
hist(fit_ipw, type = "weights")
```

### 5.5 Trimming and Truncation

```{r trim-trunc}
# Trimmed at 1st/99th percentile of PS
fit_trim <- estimate_ipwrisk(spec, risk_time = c(12, 24), trim = 0.01)
print(fit_trim)

# Truncated weights at 99th percentile
fit_trunc <- estimate_ipwrisk(spec, risk_time = c(12, 24), trunc = 0.99)
print(fit_trunc)
```

## 6. Table 1: Baseline Characteristics

```{r table1}
# Unweighted Table 1
make_table1(dat, vars = c("age", "sex", "biomarker", "comorbidity"),
            by = "treatment", smd = TRUE)
```

```{r table1-weighted}
# Weighted Table 1 (using IPW fit)
make_table1(fit_ipw, vars = c("age", "sex", "biomarker", "comorbidity"),
            weighted = TRUE, smd = TRUE)
```

## 7. Table 2: Results Summary

```{r table2}
make_table2(fit_ipw, risk_time = 24)
```

## 8. G-Computation

```{r gcomp}
spec_gc <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment)

fit_gc <- estimate_gcomprisk(spec_gc, risk_time = c(6, 12, 18, 24, 30, 36))
print(fit_gc)
plot(fit_gc)
```

## 9. AIPW (Doubly Robust)

```{r aipw}
spec_aipw <- specify_models(data = dat) |>
  identify_outcome(event, formula = ~ treatment + age + sex + biomarker,
                   type = "time_to_event") |>
  identify_treatment(treatment, formula = ~ age + sex + biomarker) |>
  identify_censoring(censored, formula = ~ age + sex, model = "coxph")

fit_aipw <- estimate_aipwrisk(
  spec_aipw,
  risk_time = c(6, 12, 18, 24, 30, 36)
)
print(fit_aipw)
plot(fit_aipw)
```

## 10. Comparing Estimators

```{r compare}
comp <- compare_fits(
  IPW = fit_ipw,
  GComp = fit_gc,
  AIPW = fit_aipw,
  risk_time = 24
)
print(comp)
```

## 11. Hazard Ratio Estimation

```{r ipwhr}
hr_fit <- estimate_ipwhr(spec)
print(hr_fit)

# Forest plot
forest_plot(hr_fit)

# HR data for custom processing
hr_data(hr_fit)

# Table 2 for HR
make_table2(hr_fit)
```

## 12. TMLE Extensions

### 12.1 Point-Treatment TMLE (Binary Outcome)

We create a binary outcome: event by 24 months.

```{r tmle-point, eval = requireNamespace("tmle", quietly = TRUE) && requireNamespace("SuperLearner", quietly = TRUE)}
dat$event_24 <- as.integer(dat$event == 1 & dat$time <= 24)

fit_tmle <- estimate_tmle_risk_point(
  data = dat,
  treatment = "treatment",
  outcome = "event_24",
  covariates = c("age", "sex", "biomarker", "comorbidity"),
  sl_library = c("SL.glm", "SL.mean"),
  truncate = 0.01
)
print(fit_tmle)
make_table2(fit_tmle)
```

### 12.2 Survival TMLE

```{r surv-tmle, eval = requireNamespace("tmle", quietly = TRUE) && requireNamespace("SuperLearner", quietly = TRUE)}
fit_surv_tmle <- estimate_surv_tmle(
  data = dat,
  treatment = "treatment",
  time = "time",
  event = "event",
  covariates = c("age", "sex", "biomarker"),
  target_times = c(12, 24),
  sl_library = c("SL.glm", "SL.mean")
)
print(fit_surv_tmle)
plot(fit_surv_tmle)
```

### 12.3 Longitudinal TMLE (LMTP)

```{r lmtp, eval = requireNamespace("lmtp", quietly = TRUE)}
# Simple example with 2 time points
set.seed(99)
n_lmtp <- 200
lmtp_dat <- data.frame(
  L1 = rnorm(n_lmtp),
  A1 = rbinom(n_lmtp, 1, 0.5),
  L2 = rnorm(n_lmtp),
  A2 = rbinom(n_lmtp, 1, 0.5),
  Y = rbinom(n_lmtp, 1, 0.3)
)

fit_lmtp <- estimate_lmtp(
  data = lmtp_dat,
  treatment_vars = c("A1", "A2"),
  outcome = "Y",
  baseline = c("L1"),
  time_varying = list(c("L2")),
  shift_fun = function(data, trt) rep(1, nrow(data)),
  outcome_type = "binomial"
)
print(fit_lmtp)
```

## 13. Summary

This vignette demonstrated the core causalRiskeR workflow:

| Step | Function | Description |
|------|----------|-------------|
| 1 | `sim_func1()` | Simulate data |
| 2 | `specify_models()` | Create specification |
| 3 | `identify_*()` | Declare model components |
| 4 | `estimate_ipwrisk()` | IPW cumulative risk |
| 5 | `estimate_gcomprisk()` | G-computation risk |
| 6 | `estimate_aipwrisk()` | AIPW doubly robust risk |
| 7 | `estimate_ipwhr()` | Weighted Cox HR |
| 8 | `estimate_tmle_risk_point()` | Point TMLE |
| 9 | `estimate_surv_tmle()` | Survival TMLE |
| 10 | `estimate_lmtp()` | Longitudinal TMLE |
| 11 | `make_table1()` | Baseline table |
| 12 | `make_table2()` | Results table |
| 13 | `make_wt_summary_table()` | Weight diagnostics |
| 14 | `extreme_weights()` | Extreme weights |
| 15 | `hist()` | PS/weight histograms |
| 16 | `forest_plot()` | Forest plot for HRs |
| 17 | `compare_fits()` | Compare estimators |

```{r session-info}
sessionInfo()
```
