---
title: "Design dossier"
params:
  dossier: ""
format:
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
---

```{r}
d <- readRDS(params$dossier)
kt <- function(x) knitr::kable(x, digits = 3, row.names = FALSE)
```

Lock `r substr(d$lock_hash, 1, 12)`, dossier `r substr(d$dossier_hash, 1, 12)`,
cleanTMLE `r d$version`, built `r d$created`.

## Decision

Primary estimand: **`r if (is.na(d$decision$primary)) "none feasible" else d$decision$primary`**.
Negative-control verdict: **`r d$decision$nc_verdict`**.

```{r, results = "asis"}
pr <- d$decision$primary
if (!is.na(pr) && isTRUE(d$decision$unresolved[[pr]]))
  cat("The primary estimand's status rests on at least one cell whose 95% interval",
      "still contained a tolerance at max_reps; that cell was decided on its point",
      "estimate.\n")
```

```{r}
kt(data.frame(estimand = names(d$decision$status), status = unlist(d$decision$status),
              unresolved = unlist(d$decision$unresolved[names(d$decision$status)])))
```

```{r}
if (length(d$decision$candidates))
  kt(do.call(rbind, lapply(names(d$decision$candidates), function(e) {
    x <- d$decision$candidates[[e]]
    data.frame(estimand = e, candidate = x$id, learners = paste(x$learners, collapse = ", "),
               truncation = x$truncation)
  })))
```

## Design

Overlap grade: **`r d$design$overlap_grade`**; share outside the band:
`r round(d$design$share_outside_band, 3)`.

```{r}
kt(d$design$overlap); kt(d$design$ess); kt(d$design$balance)
if (!is.null(d$design$edge) && nrow(d$design$edge)) kt(d$design$edge)
```

## Simulation

`r d$simulation$scope`

`r d$simulation$decision_note`

```{r}
kt(d$simulation$verdict)
if (!is.null(d$simulation$extension) && nrow(d$simulation$extension))
  kt(d$simulation$extension)
kt(d$simulation$metrics[c("estimand", "candidate", "surface", "reps", "bias", "coverage",
                          "coverage_lower", "coverage_upper", "rmse", "fail_rate", "pass",
                          "unresolved")])
```

## Negative controls

```{r}
if (is.null(d$nc)) cat("None declared.")
```

```{r}
if (!is.null(d$nc)) kt(d$nc$by_rung)
```

```{r}
if (!is.null(d$nc) && !is.null(d$nc$by_domain)) kt(d$nc$by_domain)
```

```{r}
if (!is.null(d$nc)) kt(d$nc$table)
```

## Notes

```{r, results = "asis"}
cat(paste0("- ", d$notes), sep = "\n")
```
